GMD data: first steps
========================================================

Date: 14.07.2014  
Author: Heike Sprenger  

### set working directory
```{r set working directory}
# setwd("D:/work/repos/trost_metabolomics")
# setwd("~/work/repos/trost_metabolomics")
```

[solution for issue with working directory and knitr](https://github.com/yihui/knitr/issues/277)

### Load workspace, packages and scripts
```{r load workspace, message=FALSE}
# load packages
library(knitr)
library(ggplot2)
library(reshape)
library(reshape2)
library(pander)
library(plyr)
library(RODBC)
# library(RMySQL)
# library(yaml)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# options(width=600)

# load workspace
```


### Set up database connection
```{r set up database connection}
dbhandle <- odbcDriverConnect('driver={SQL Server}; 
                              server=r710sqljahu; 
                              database=GMD; 
                              trusted_connection=true')

# login <- yaml.load_file("../libpurzel/login.yaml")

# phenotyper <- dbConnect(MySQL(), user=login$user, password=login$passwd, dbname=login$db, host=login$host)  

# gmd_connection <- dbConnect(MySQL(), 
#                             user = "MPIMP-GOLM\HSprenger", 
#                             password = Potato, 
#                             dbname = r710sqljahu, 
#                             host = gmd.mpimp-golm.mpg.de)  
```


### read table with GMD ids and "real" experiment names
```{r read table with GMD ids and "real" experiment names}
gmd_trost_experiments <- read.table("data/gmd_trost_experiments.txt", sep="\t", header=TRUE)
# order table by ID
gmd_trost_experiments <- gmd_trost_experiments[order(gmd_trost_experiments$id),]

pander(head(gmd_trost_experiments))
```


### get trost_TagList from GMD
trost_TagList shows all experiments that belong to the TROST project.
```{r get trost_TagList from GMD}
trost_TagList <- sqlQuery(dbhandle, 'select * from tf.TagList where comment=\'trost\' or name like \'%trost%\'')
levels(trost_TagList$id)
# order table
trost_TagList <- trost_TagList[order(trost_TagList$id),]
```


#### join trost_TagList and gmd_trost_experiments tables
```{r join trost_TagList and gmd_trost_experiments tables}
trost_TagList <- join(trost_TagList, gmd_trost_experiments, by="id")
colnames(trost_TagList)
# remove duplicate column
trost_TagList <- trost_TagList[,-6]

# save table
write.table(trost_TagList, "data/trost_TagList.txt", sep="\t")
```


### Query for table of ALL TROST chromatogram IDs and respective names (e.g. 11290if_9) from GMD which are used as IDs in Phenotyper
```{r Query for table of ALL TROST chromatogram IDs and respective names}
trost_gmd_ids <- sqlQuery(dbhandle, 'SELECT GC_Chromatogram.name, 
                                            tf.IntensityValue.FK_chromatogram 
                                     FROM tf.IntensityValue 
                                     INNER JOIN GC_Chromatogram 
                                          ON tf.IntensityValue.FK_chromatogram = GC_Chromatogram.id 
                                     INNER JOIN tf.TagList 
                                          ON tf.IntensityValue.FK_TagList = tf.TagList.id
                                     WHERE (tf.TagList.comment = \'trost\') or 
                                           (tf.TagList.name like \'%trost%\')
                                     GROUP BY FK_chromatogram, GC_Chromatogram.name')

colnames(trost_gmd_ids) <- c("GMD_id", "chromatogram")
dim(trost_gmd_ids)
# 5723 2

pander(head(trost_gmd_ids))

# order table by chromatogram id
trost_gmd_ids <- trost_gmd_ids[order(trost_gmd_ids$chromatogram),]

# save table
write.table(trost_gmd_ids, "data/trost_gmd_ids.txt", sep="\t")
```


### Import TROST phenotyper query results 
####(gained by metaquery and masterquery --> phenotyper_metadata.Rmd)
### and join phenotyper results with trost_gmd_ids by common GMD_id
```{r Import TROST phenotyper query results}
phenotyper_result_joined <- read.table("data/phenotyper_result_joined.txt", sep="\t", header=TRUE)
phenotyper_result_joined_valdis <- read.table("data/phenotyper_result_joined_valdis.txt", sep="\t", header=TRUE)
dim(phenotyper_result_joined_valdis)
# 4555 24

phenotyper_result_joined_with_gmd_ids <- join(phenotyper_result_joined_valdis, trost_gmd_ids, by="GMD_id")

# order table by chromatogram id
phenotyper_result_joined_with_gmd_ids <- phenotyper_result_joined_with_gmd_ids[order(phenotyper_result_joined_with_gmd_ids$chromatogram),]

# save table
write.table(phenotyper_result_joined_with_gmd_ids, "data/phenotyper_result_joined_with_gmd_ids.txt", sep="\t")
```

### Import analytes overlap table 
#### (overlap regarding 6 selected TROST experiments, not the QC experiments) includes 110 analytes (without contaminations/internal standards)
```{r Import analytes overlap table}
analytes_6sel_exp <- read.table("data/gmd_metabolites_overlap_6selected_exp.txt", sep="\t", header=TRUE)
#analytes_7sel_exp <- read.table("data/gmd_metabolites_overlap_7selected_exp.txt", sep="\t", header=TRUE, allowEscapes=T)

# number of ALL overlapping analytes
length(analytes_6sel_exp$analyte)
# 110
#length(analytes_7sel_exp$analyte)
# 107

# order table
analytes_6sel_exp_sort <- analytes_6sel_exp[order(analytes_6sel_exp$analyte),]
#analytes_7sel_exp_sort <- analytes_7sel_exp[order(analytes_7sel_exp$analyte),]

# save sorted table
write.table(analytes_6sel_exp_sort, "data/analytes_6sel_exp_sort.txt", sep="\t")
#write.table(analytes_7sel_exp_sort, "data/analytes_7sel_exp_sort.txt", sep="\t", eol = "\r")
```


### sessionInfo
```{r sessionInfo}
sessionInfo()
```
