---
title: "GMD analytes for all Trost experiments"
author: "Heike Sprenger"
date: "Friday, November 06, 2015"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---

## Set working directory
```{r set working directory}
# setwd("D:/work/repos/trost_metabolomics")
# setwd("~/work/repos/trost_metabolomics")
```

[solution for issue with working directory and knitr](https://github.com/yihui/knitr/issues/277)

## Load workspace, packages and scripts
```{r load workspace, message=FALSE, warning=FALSE}
# load packages
library(knitr)
library(ggplot2)
library(reshape)
library(reshape2)
library(pander)
library(plyr)
library(RODBC)
library(DT)
# library(RMySQL)
# library(yaml)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# options(width=600)

# load workspace
# load("gmd_data.RData")
```


## Source R functions
```{r source R functions}
source("../libpurzel/func_get_experiment_id.R")
source("../libpurzel/func_get_gmd_analytes.R")
source("../libpurzel/func_create_analytes_table.R")
```


## Set up database connection
```{r set up database connection}
dbhandle <- odbcDriverConnect('driver={SQL Server}; 
                              server=r710sqljahu; 
                              database=GMD; 
                              trusted_connection=true')
```


## Function to get experiment_id
```{r define experiment id}
trost_TagList <- read.table("data/trost_TagList.txt", sep="\t", header=TRUE)

experiment_id <- func_get_experiment_id("JKI_Shelter_2014")
```


## Load ALL GMD raw values
```{r load ALL GMD raw values}
all_values_cast <- read.table("data/gmd_raw_values_matrix.txt", sep = "\t", header = T, check.names = F)

dim(all_values_cast)
# 5723 samples
# 548 analytes

pander(all_values_cast[1:3,1:3])
```


## Get analytes for specific experiment
```{r get analytes for specific experiment}

gmd_analytes_jki_shelter_2014 <- func_get_gmd_analytes( func_get_experiment_id("JKI_Shelter_2014") )

gmd_analytes_mpi_test_1 <- func_get_gmd_analytes( func_get_experiment_id("MPI_Test1_2") )
gmd_analytes_mpi_test_2 <- func_get_gmd_analytes( func_get_experiment_id("MPI_Test2") )
gmd_analytes_jki_test_1 <- func_get_gmd_analytes( func_get_experiment_id("JKI_Test1") )
gmd_analytes_jki_test_2 <- func_get_gmd_analytes( func_get_experiment_id("JKI_Test2") )
gmd_analytes_mpi_feld_2011 <- func_get_gmd_analytes( func_get_experiment_id("MPI_Feld_2011") )
gmd_analytes_mpi_feld_2012 <- func_get_gmd_analytes( func_get_experiment_id("MPI_Feld_2012") )
gmd_analytes_jki_feld_2012 <- func_get_gmd_analytes( func_get_experiment_id("JKI_Feld_2012") )

# overlap
gmd_analytes_overlap_6 <- Reduce(intersect, list(gmd_analytes_mpi_test_1$analyteID,
                              gmd_analytes_mpi_test_2$analyteID,
                              gmd_analytes_jki_test_1$analyteID,
                              gmd_analytes_mpi_feld_2011$analyteID,
                              gmd_analytes_mpi_feld_2012$analyteID,
                              gmd_analytes_jki_feld_2012$analyteID))

gmd_analytes_overlap_7 <- Reduce(intersect, list(gmd_analytes_mpi_test_1$analyteID,
                              gmd_analytes_mpi_test_2$analyteID,
                              gmd_analytes_jki_test_1$analyteID,
                              gmd_analytes_jki_test_2$analyteID,
                              gmd_analytes_mpi_feld_2011$analyteID,
                              gmd_analytes_mpi_feld_2012$analyteID,
                              gmd_analytes_jki_feld_2012$analyteID))

length(gmd_analytes_overlap_6)
length(gmd_analytes_overlap_7)

# for all trost/valdis experiments
gmd_analytes <- func_get_all_gmd_analytes()
# 551 analytes
pander(head(gmd_analytes))
```


## Load inventory list (from AG Kopka)
Problem with hash "#" symbol http://stackoverflow.com/questions/7086945/how-can-read-numeral-signs-as-part-of-a-column-header
```{r load inventory list}
inventory_list <- read.table("data/inventory_AFE.txt", header = T, sep = "\t", check.names = F, comment.char = "")
inventory_list_part <- inventory_list[, 1:12]
colnames(inventory_list_part)

# use DT library to show a formatted table
datatable(inventory_list_part) %>%
  formatStyle(
    'Class',
    backgroundColor = styleEqual(
      c("Reagent (Contamination)", "Internal Standard (Retention Time)"), c('red','orange')
    )
  )

length(intersect(gmd_analytes$analyteID, inventory_list[,1]))
length(intersect(gmd_analytes_overlap_6, inventory_list[,1]))
length(intersect(gmd_analytes_overlap_7, inventory_list[,1]))

inventory_list_subset_6 <- subset(inventory_list, inventory_list[,1] %in% gmd_analytes_overlap_6)
inventory_list_subset_7 <- subset(inventory_list, inventory_list[,1] %in% gmd_analytes_overlap_7)

write.table(inventory_list_subset_6, "data/inventory_list_subset_6exp.txt", sep = "\t", row.names = F)
write.table(inventory_list_subset_7, "data/inventory_list_subset_7exp.txt", sep = "\t", row.names = F)
```


## Remove several analytes

**remove internal standards and contaminations**

**remove analytes that are present as duplicates, prefer the other derivate of the same metabolite**

* A117006-101, Proline
* A152002-101, Aspartic acid
* A163001-101, Glutamic acid
* A177004-101, Glucose
* A180002-101, Fructose
* A187002-101, Fructose
* A189002-101, Glucose
* A239001-101, Glucose-6-phosphate

```{r remove several analytes}
analytes_to_remove <- which(inventory_list_part$Class %in% c("Internal Standard (Quantification)",
                                                             "Internal Standard (Retention Time)",
                                                             "Reagent (Contamination)") |
                            inventory_list_part$MPIMP_ID_Analyte %in% c("A117006-101", "A152002-101", 
                                                                        "A163001-101", "A177004-101",
                                                                        "A180002-101", "A187002-101",
                                                                        "A189002-101", "A239001-101"))

# create additional column for selected/removed analytes
inventory_list_part$select <- "select"
inventory_list_part$select[analytes_to_remove] <- "remove"
```


## Join gmd_analytes table and inventory list
```{r join gmd_analytes table and inventory list}
gmd_analytes_inventory <- merge(gmd_analytes, inventory_list_part, 
                                by.x = "analyteID", by.y = "MPIMP_ID_Analyte")
```


## Define sets of specific experiments
```{r define sets of specific experiments}
# 6 experiments used for publication
experiments_6 <- c("MPI_Test1_2", "MPI_Test2", "JKI_Test1", 
                   "MPI_Feld_2011", "MPI_Feld_2012", "JKI_Feld_2012")

# 7 experiments used for PhD thesis
experiments_7 <- c("MPI_Test1_2", "MPI_Test2", "JKI_Test1", 
                   "JKI_Test2", "MPI_Feld_2011", "MPI_Feld_2012", "JKI_Feld_2012")

# including all TROST experiments
experiments_15 <- c("MPI_Test1_2", "MPI_Test2", "JKI_Test1", "JKI_Test2", 
                    "MPI_Feld_2011", "MPI_Feld_2012", "JKI_Feld_2012", "JKI_Feld_2012", "Dethlingen_2011",
                    "JKI_Shelter_2011", "JKI_Shelter_2012", "MPI_Pruef1", "MPI_Pruef2", "MPI_Pruef3",
                    "breeder_trials")

# including all TROST and VALDIS experiments
experiments_17 <- c("MPI_Test1_2", "MPI_Test2", "JKI_Test1", "JKI_Test2", 
                    "MPI_Feld_2011", "MPI_Feld_2012", "JKI_Feld_2012", "JKI_Feld_2012", "Dethlingen_2011",
                    "JKI_Shelter_2011", "JKI_Shelter_2012", "MPI_Pruef1", "MPI_Pruef2", "MPI_Pruef3",
                    "breeder_trials", "JKI_Shelter_2014", "MPI_FGH_2014")
```


## Create analytes table for all experiments

**use function to create an analytes table with additional columns per single experiment**
**each single column states "yes" or "no" if analyte was annotated in a specific experiment**
```{r create analytes table for all experiments}
analytes_table_6 <- func_create_analytes_table(gmd_analytes_inventory, experiments_6)
analytes_table_7 <- func_create_analytes_table(gmd_analytes_inventory, experiments_7)
analytes_table_15 <- func_create_analytes_table(gmd_analytes_inventory, experiments_15)
analytes_table_17 <- func_create_analytes_table(gmd_analytes_inventory, experiments_17)

analytes_table_6_overlap <- subset(analytes_table_6, analytes_table_6$count_yes == 6)
analytes_table_7_overlap <- subset(analytes_table_7, analytes_table_7$count_yes == 7)

# use only the selected analytes to get rid of internal standards and contaminations
analytes_table_6_overlap_select <- subset(analytes_table_6_overlap, 
                                          analytes_table_6_overlap$select == "select")

analytes_table_7_overlap_select <- subset(analytes_table_7_overlap, 
                                          analytes_table_7_overlap$select == "select")

dim(analytes_table_6_overlap_select)
dim(analytes_table_7_overlap_select)
```


## Save filtered analytes table for specific experimental sets
```{r save filtered analytes table for specific experimental sets}
# order by FK_Analyte ID
analytes_table_6_overlap_select_ordered <- analytes_table_6_overlap_select[order(analytes_table_6_overlap_select$FK_Analyte), ]
analytes_table_7_overlap_select_ordered <- analytes_table_7_overlap_select[order(analytes_table_7_overlap_select$FK_Analyte), ]

write.table(analytes_table_6_overlap_select, "data/analytes_table_6exp_overlap_select.txt", 
            sep="\t", row.names = F)
write.table(analytes_table_7_overlap_select, "data/analytes_table_7exp_overlap_select.txt", 
            sep="\t", row.names = F)

write.table(analytes_table_6_overlap_select_ordered, "data/analytes_table_6exp_overlap_select_ordered.txt", 
            sep="\t", row.names = F)
write.table(analytes_table_7_overlap_select_ordered, "data/analytes_table_7exp_overlap_select_ordered.txt", 
            sep="\t", row.names = F)
```


### Which analytes have long names
**Could be used to create new lines in analyte name**
```{r which analytes have long names}
long_names_idx <- which(unlist(lapply(as.character(analytes_table_7_overlap_select_ordered$Name), nchar)) > 20)
analytes_table_7_overlap_select_ordered$Name[long_names_idx]
```


## SessionInfo
```{r sessionInfo}
save.image("gmd_data.RData")
sessionInfo()
```