---
title: "GMD subset for specific samples and analytes"
author: "Heike Sprenger"
date: "Friday, November 06, 2015"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---

# Set working directory
```{r set working directory}
# setwd("D:/work/repos/trost_metabolomics")
# setwd("~/work/repos/trost_metabolomics")
```

[solution for issue with working directory and knitr](https://github.com/yihui/knitr/issues/277)

# Load workspace, packages and scripts
```{r load workspace, message=FALSE, warning=FALSE}
# load packages
library(knitr)
library(ggplot2)
library(reshape)
library(reshape2)
library(pander)
library(plyr)
library(DT)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# options(width=600)

# load workspace
# load("gmd_data.RData")
```


# Source R functions
```{r source R functions}
source("../libpurzel/func_compile_subset.R")
```


# Load data
## Load gmd raw data for all TROST/VALDIS experiments
Created with `gmd_raw_values.Rmd`
```{r load gmd raw data for all TROST/VALDIS experiments}
all_values_cast <- read.table("data/gmd_raw_values_matrix.txt", sep = "\t", header = T, check.names = F)
dim(all_values_cast)
# rows: samples (5723)
# cols: analytes (548)

# transpose raw data
all_values_t <- as.data.frame(t(all_values_cast))

dim(all_values_t)
# rows: analytes (548)
# columns: samples (5723)
```


## Load normalized data for pot/field experiments
**Data were normalized separately per cultivation type!**
```{r load normalized data for pot/field experiments}
# 3 pot experiments
values_3exp_pot_late_before <- 
  read.table("../../TROST/GC-MS/Publication/output/merge_pot_experiments/values_norm_log10_late_before_woOutlier.txt", 
             header = T, sep = "\t", check.names = F)
dim(values_3exp_pot_late_before)
# rows: 72 samples
# columns: 107 analytes

# 3 field experiments
values_3exp_field_late <- 
  read.table("../../TROST/GC-MS/Publication/output/merge_field_experiments_late/values_norm_log10.txt", 
             header = T, sep = "\t", check.names = F)
dim(values_3exp_field_late)
# rows: 144 samples
# columns: 107 analytes
```


### Combine normalized data for pot/field experiments
```{r combine normalized data for pot/field experiments}
values_6exp_normalized <- rbind(values_3exp_pot_late_before,
                                values_3exp_field_late)

dim(values_6exp_normalized)
# rows: 216 samples
# columns: 107 analytes

# transpose normalized data
values_6exp_normalized_t <- as.data.frame(t(values_6exp_normalized))
dim(values_6exp_normalized_t)
# rows: 107 analytes
# columns: 216 samples

# order normalized data by colnames (sampleID) and rownames (analyteID)
values_6exp_normalized_t_ordered <- values_6exp_normalized_t[order(rownames(values_6exp_normalized_t)),
                                                             order(colnames(values_6exp_normalized_t))]
```


# Load sample_lists
## Load sample_list of 6 experiments (for publication)
Created with:

* `merge_all_experiments_late.Rmd` for all 6 experiments
* `merge_pot_experiments_late_before.Rmd` for 3 pot experiments
* `merge_field_experiments_late.Rmd` for 3 field experiments

(folder: TROST/GC-MS/Publication/scripts)

```{r load sample_list of 6 experiments}
# all 6 experiments
sample_list_6exp <- read.table("../../TROST/GC-MS/Publication/output/merge_all_experiments_late/factors.txt",
                               header = T, sep = "\t")
dim(sample_list_6exp)
# 216 samples (rows)

# add column with experiment ID for publication
sample_list_6exp$experiment <- sample_list_6exp$trial
levels(sample_list_6exp$experiment) <- c("F4", "G3", "F1", "F2", "G1", "G2")
levels(sample_list_6exp$trial)

# change order of experiment levels
sample_list_6exp$experiment <- factor(sample_list_6exp$experiment, levels = c("G1", "G2", "G3", "F1", "F2", "F4"))
pander(table(sample_list_6exp$trial, sample_list_6exp$experiment))

# order sample_list by chromtogram ID
sample_list_6exp_ordered <- sample_list_6exp[order(sample_list_6exp$chromatogram),]

# order sample_list by experiment, treatment, cultivar
sample_list_6exp_ordered2 <- sample_list_6exp[order(sample_list_6exp$experiment, 
                                                    sample_list_6exp$treatment, 
                                                    sample_list_6exp$cultivar),]
```


## Load sample list of greenhouse and field data separately
```{r load sample list of greenhouse and field data separately}
# 3 field experiments
sample_list_3exp_field_late <- read.table("../../TROST/GC-MS/Publication/output/merge_field_experiments_late/factors.txt",
                               header = T, sep = "\t")
dim(sample_list_3exp_field_late)

# 3 pot experiments
sample_list_3exp_pot_late_before <- read.table("../../TROST/GC-MS/Publication/output/merge_pot_experiments/factors_late_before.txt",
                               header = T, sep = "\t")
dim(sample_list_3exp_pot_late_before)
```


# Load subset of analytes list (107, for 7 experiments)
Created with `gmd_analytes.Rmd`
```{r load subset of analytes list}
analytes_table_7_overlap_select <- read.table("data/analytes_table_7exp_overlap_select.txt",
                                              header = T, sep = "\t")
dim(analytes_table_7_overlap_select)
# 107 analytes (rows)
head(analytes_table_7_overlap_select$FK_Analyte)
```


## Select only part of analytes list that is needed for supplemental table
```{r select only part of analytes list that is needed for supplemental table}
analytes_table_7_overlap_supp <- subset(analytes_table_7_overlap_select,
                                        select = c("FK_Analyte","Name", "Class", "MPIMP_ID_Metabolite", 
                                                   "Sum_Formula", "KEGG_ID_Metabolite", "CAS_ID_Metabolite",
                                                   "Derivative", "analyteID", "CAS_ID_Analyte"))

pander(head(analytes_table_7_overlap_supp))
```


# Apply func_compile_subset
## Apply func_compile_subset for raw data
```{r apply func_compile_subset for raw data}
raw_data_subset <- func_compile_subset(raw_data = all_values_t, 
                                       sample_ids = sample_list_6exp$chromatogram,
                                       analyte_ids = analytes_table_7_overlap_supp$FK_Analyte,
                                       sample_list = sample_list_6exp_ordered,
                                       analyte_list = analytes_table_7_overlap_supp)
dim(raw_data_subset)
# rows: 107 analytes
# columns: 225 = 216 samples + 9 columns with additional information about analytes

sample_list_6exp_ordered2_part <- subset(sample_list_6exp_ordered2, 
                                         select = c("cultivar", "treatment", "experiment", "cultivation"))
```


## Apply func_compile_subset for normalized data
```{r apply func_compile_subset for normalized data}
normalized_data_subset <- func_compile_subset(raw_data = values_6exp_normalized_t_ordered, 
                                       sample_ids = sample_list_6exp$chromatogram,
                                       analyte_ids = analytes_table_7_overlap_supp$FK_Analyte,
                                       sample_list = sample_list_6exp_ordered,
                                       analyte_list = analytes_table_7_overlap_supp)
dim(normalized_data_subset)
```


# Save final supplemental table for manuscript
```{r save final supplemental table for manuscript}
write.table(raw_data_subset, "output/supplemental_table_6exp_raw_data.txt", sep = "\t", row.names = F)
write.table(normalized_data_subset, "output/supplemental_table_6exp_normalized_data.txt", sep = "\t", row.names = F)
write.table(sample_list_6exp_ordered2_part, "output/supplemental_table_6exp_description.txt", sep = "\t", row.names = F)
```


# SessionInfo
```{r sessionInfo}
save.image("gmd_data.RData")
sessionInfo()
```