---
title: GMD data - first steps
author: "Heike Sprenger"
date: "July 14, 2014"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---

## Set working directory
```{r set working directory}
# setwd("D:/work/repos/trost_metabolomics")
# setwd("~/work/repos/trost_metabolomics")
```

[solution for issue with working directory and knitr](https://github.com/yihui/knitr/issues/277)

## Load workspace, packages and scripts
```{r load workspace, message=FALSE, warning=FALSE}
# load packages
library(knitr)
library(ggplot2)
library(reshape)
library(reshape2)
library(pander)
library(plyr)
library(RODBC)
# library(RMySQL)
library(yaml)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# options(width=600)

# load workspace
```


## Set up database connection to GMD
[How to access MS-SQL databases in linux](http://wiki.mpimp-golm.mpg.de/index.php/Access_MSSQL_databases)
[How to identify operating system from R](http://conjugateprior.org/2015/06/identifying-the-os-from-r/)

```{r set up database connection to GMD}

login_gmd <- yaml.load_file("../functions/login_gmd.yaml")

if (.Platform$OS.type == 'windows'){
  dbhandle <- odbcDriverConnect('driver={SQL Server}; 
                                server=r710sqljahu; 
                                database=GMD; 
                                trusted_connection=true')
} else { # if .Platform$OS.type == 'unix'
  dbhandle <- odbcDriverConnect(
    connection=paste0("server=",login_gmd$host,
                      ";database=",login_gmd$db,
                      ";uid=",login_gmd$user,
                      ";pwd=",login_gmd$passwd,
                      ";Port=1433;driver=FreeTDS;TDS_Version=8.0;"))
}

# test of connection
test <- sqlQuery(dbhandle, 'select * from GMD.tf.TagList')
str(test)
```


## Read table with GMD ids and "real" experiment names

**gmd_trost_experiments.txt** was prepared before any analysis in R!

```{r read table with GMD ids and "real" experiment names}
gmd_trost_experiments <- read.table("data/gmd_trost_experiments.txt", sep="\t", header=TRUE)
# order table by ID
gmd_trost_experiments <- gmd_trost_experiments[order(gmd_trost_experiments$id),]

dim(gmd_trost_experiments)
pander(head(gmd_trost_experiments))
```


## Get ``trost_TagList`` from GMD
**``trost_TagList`` shows all experiments that belong to the TROST/VALDIS project**
```{r get trost_TagList from GMD}
trost_TagList <- sqlQuery(dbhandle, 'select * from tf.TagList where comment=\'trost\' or name like \'%trost%\'')
head(trost_TagList)
# order table
trost_TagList <- trost_TagList[order(trost_TagList$id),]

dim(trost_TagList)
```


## Join ``trost_TagList`` and ``gmd_trost_experiments`` tables
```{r join trost_TagList and gmd_trost_experiments tables}
trost_TagList <- join(trost_TagList, gmd_trost_experiments, by="id")
colnames(trost_TagList)
# remove duplicate column
trost_TagList <- trost_TagList[,-6]

# save table
write.table(trost_TagList, "data/trost_TagList.txt", sep="\t")
```


## Query for table of all TROST chromatogram IDs and respective names
* FK_chromatogram = chromatogram IDs, long IDs (e.g. 06E072D8-B9DE-4BDE-B927-BFF3C4EDDB4B)
* names (e.g. 11290if_9) from GMD are used as IDs in Phenotyper

```{r query for table of ALL TROST chromatogram IDs and respective names}
trost_gmd_ids <- sqlQuery(dbhandle, 'SELECT GC_Chromatogram.name, 
                                            tf.IntensityValue.FK_chromatogram 
                                     FROM tf.IntensityValue 
                                     INNER JOIN GC_Chromatogram 
                                          ON tf.IntensityValue.FK_chromatogram = GC_Chromatogram.id 
                                     INNER JOIN tf.TagList 
                                          ON tf.IntensityValue.FK_TagList = tf.TagList.id
                                     WHERE (tf.TagList.comment = \'trost\') or 
                                           (tf.TagList.name like \'%trost%\')
                                     GROUP BY FK_chromatogram, GC_Chromatogram.name')

colnames(trost_gmd_ids) <- c("GMD_id", "chromatogram")
dim(trost_gmd_ids)
# 6806 2

pander(head(trost_gmd_ids))

# order table by chromatogram id
trost_gmd_ids <- trost_gmd_ids[order(trost_gmd_ids$chromatogram),]

# save table
write.table(trost_gmd_ids, "data/trost_gmd_ids.txt", sep="\t")
```


## Import TROST phenotyper query results 

**gained by metaquery and masterquery --> ``phenotyper_metadata.Rmd``**
**and join phenotyper results with trost_gmd_ids by common GMD_id**
```{r import TROST phenotyper query results}
phenotyper_result_joined <- read.table("data/phenotyper_result_joined.txt", sep="\t", header=TRUE)
dim(phenotyper_result_joined)
# 4790 24

# phenotyper_result_joined_with_gmd_ids <- join(phenotyper_result_joined_valdis, trost_gmd_ids, by="GMD_id")
phenotyper_result_joined_with_gmd_ids <- join(phenotyper_result_joined, trost_gmd_ids, by="GMD_id")
dim(phenotyper_result_joined_with_gmd_ids)
# 4790 25

# order table by chromatogram id
phenotyper_result_joined_with_gmd_ids <- phenotyper_result_joined_with_gmd_ids[order(phenotyper_result_joined_with_gmd_ids$chromatogram),]

# save table
write.table(phenotyper_result_joined_with_gmd_ids, "data/phenotyper_result_joined_with_gmd_ids.txt", sep="\t")
```


## Import analytes overlap table 

**not really necessary because already done in ``gmd_analytes.Rmd``**

* overlap regarding 6 selected TROST experiments, not the QC experiments) includes 110 analytes (without contaminations/internal standards)
* overlap regarding 7 selected TROST experiments, not the QC experiments) includes 107 analytes (without contaminations/internal standards)

```{r import analytes overlap table}
#analytes_6sel_exp <- read.table("data/gmd_metabolites_overlap_6selected_exp.txt", sep="\t", header=TRUE)
#analytes_7sel_exp <- read.table("data/gmd_metabolites_overlap_7selected_exp.txt", sep="\t", header=TRUE, allowEscapes=T)

# number of ALL overlapping analytes
#length(analytes_6sel_exp$analyte)
# 110
#length(analytes_7sel_exp$analyte)
# 107

# order table
#analytes_6sel_exp_sort <- analytes_6sel_exp[order(analytes_6sel_exp$analyte),]
#analytes_7sel_exp_sort <- analytes_7sel_exp[order(analytes_7sel_exp$analyte),]

# save sorted table
#write.table(analytes_6sel_exp_sort, "data/analytes_6sel_exp_sort.txt", sep="\t")
#write.table(analytes_7sel_exp_sort, "data/analytes_7sel_exp_sort.txt", sep="\t", eol = "\r")
```


## Save workspace
```{r save workspace}
save.image("phenotyper.RData")
#sessionInfo()
```
