GMD raw values for all Trost experiments
========================================================

Date: 14.07.2014  
Author: Heike Sprenger  

### set working directory
```{r set working directory}
# setwd("D:/work/repos/trost_metabolomics")
# setwd("~/work/repos/trost_metabolomics")
```

[solution for issue with working directory and knitr](https://github.com/yihui/knitr/issues/277)

### Load workspace, packages and scripts
```{r load workspace, message=FALSE}
# load packages
library(knitr)
library(ggplot2)
library(reshape)
library(reshape2)
library(pander)
library(reshape)
library(plyr)
library(pcaMethods)
library(psych)
library(corrgram)
library(corrplot)
library(RODBC)
# library(RMySQL)
# library(yaml)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# options(width=600)

# load workspace
# load("gmd_data.RData")
```


### source R functions
```{r source R functions, include=FALSE}
source("../libpurzel/colors.R")
source("../libpurzel/names.R")
```


### Set up database connection
```{r set up database connection}
dbhandle <- odbcDriverConnect('driver={SQL Server}; 
                              server=r710sqljahu; 
                              database=GMD; 
                              trusted_connection=true')

# login <- yaml.load_file("../libpurzel/login.yaml")

# phenotyper <- dbConnect(MySQL(), user=login$user, password=login$passwd, dbname=login$db, host=login$host)  

# gmd_connection <- dbConnect(MySQL(), 
#                             user = "MPIMP-GOLM\HSprenger", 
#                             password = Potato, 
#                             dbname = r710sqljahu, 
#                             host = gmd.mpimp-golm.mpg.de)  
```


## ALL values
### SQL Query for Intesity values of ALL Analytes and Chromatograms (= Samples) from one Experiment where the Analyte-MST is a quantitative Cluster.
```{r query for values}
all_values <- sqlQuery(dbhandle, 'SELECT FK_Analyte, GMD.tf.IntensityValue.FK_chromatogram, SUM(value) FROM GMD.tf.MSTAnnotation 
                       INNER JOIN GMD.tf.IntensityValue ON GMD.tf.MSTAnnotation.FK_MST = GMD.tf.IntensityValue.FK_MST and 
                          GMD.tf.MSTAnnotation.FK_TagList = GMD.tf.IntensityValue.FK_TagList
                       INNER JOIN GMD.dbo.GC_Chromatogram ON GMD.tf.IntensityValue.FK_chromatogram = GMD.dbo.GC_Chromatogram.id
  				             INNER JOIN GMD.tf.TagList ON GMD.tf.IntensityValue.FK_TagList = GMD.tf.TagList.id
                       WHERE FK_Analyte is not NULL and 
                       (GMD.tf.TagList.comment = \'trost\' or GMD.tf.TagList.name like \'%trost%\') and
                       GMD.tf.MSTAnnotation.[Is Quantitative Cluster] = 1
                       GROUP BY FK_Analyte, FK_chromatogram')
# Change colname
dim(all_values)
# 1055975      3
colnames(all_values)[3] <- "value"
head(all_values)

write.table(all_values, "../data/gmd_raw_values.txt", sep="\t")
```


### Reshape the values table to get a matrix with the analytes in columns and chromatograms (= samples) in rows.
```{r reshape values table}
all_values_cast <- cast(all_values, FK_chromatogram ~ FK_Analyte, value="value")
all_values_cast <- as.data.frame(all_values_cast)
dim(all_values_cast)
# 5723 samples, 549 analytes

# all_values_cast[1:3,1:3]

# change rownames and remove first column with redundant rownames
rownames(all_values_cast) <- all_values_cast$FK_chromatogram
all_values_cast <- all_values_cast[, -1]
dim(all_values_cast)
# 5723 548
all_values_cast[1:3,1:3]

write.table(all_values_cast, "../data/gmd_raw_values_matrix.txt", sep="\t")
```


### function to get experiment_id
```{r define experiment id}
trost_TagList <- read.table("../data/trost_TagList.txt", sep="\t", header=TRUE)

source("../../functions/func_get_experiment_id.R")

experiment_id <- func_get_experiment_id("MPI_Test1_2")
```

## Values Per Experiment
### Function to obtain values per experiment and cast them
```{r Function to obtain values per experiment}
source("../../functions/func_get_gmd_values.R")
```


### Execute GMD values function
```{r execute GMD values function}
gmd_values_mpitest1_2 <- func_get_gmd_values( func_get_experiment_id("MPI_Test1_2") )
gmd_values_mpitest2 <- func_get_gmd_values( func_get_experiment_id("MPI_Test2") )
gmd_values_jkitest1 <- func_get_gmd_values( func_get_experiment_id("JKI_Test1") )
gmd_values_jkitest2 <- func_get_gmd_values( func_get_experiment_id("JKI_Test2") )

gmd_values_mpi_field_2011 <- func_get_gmd_values( func_get_experiment_id("MPI_Feld_2011") )
gmd_values_mpi_field_2012 <- func_get_gmd_values( func_get_experiment_id("MPI_Feld_2012") )
gmd_values_jki_field_2012 <- func_get_gmd_values( func_get_experiment_id("JKI_Feld_2012") )
gmd_values_jki_field_2013 <- func_get_gmd_values( func_get_experiment_id("JKI_Feld_2013") )
gmd_values_dethlingen_2011 <- func_get_gmd_values( func_get_experiment_id("Dethlingen_2011") )

gmd_values_jki_shelter_2011 <- func_get_gmd_values( func_get_experiment_id("JKI_Shelter_2011") )
gmd_values_jki_shelter_2012 <- func_get_gmd_values( func_get_experiment_id("JKI_Shelter_2012") )
gmd_values_mpi_pruef1 <- func_get_gmd_values( func_get_experiment_id("MPI_Pruef1") )
gmd_values_mpi_pruef2 <- func_get_gmd_values( func_get_experiment_id("MPI_Pruef2") )
gmd_values_mpi_pruef3 <- func_get_gmd_values( func_get_experiment_id("MPI_Pruef3") )

gmd_values_breeder_trials <- func_get_gmd_values( func_get_experiment_id("breeder_trials") )

# VALDIS TROST2
gmd_values_jki_shelter_2014 <- func_get_gmd_values( func_get_experiment_id("JKI_Shelter_2014") )
gmd_values_mpi_fgh_2014 <- func_get_gmd_values( func_get_experiment_id("MPI_FGH_2014") )
```


### save tables with GMD values
```{r save tables with GMD values}
write.table(gmd_values_mpitest1_2, "../data/gmd_values_mpitest1_2.txt", sep="\t", quote=FALSE)
write.table(gmd_values_mpitest2, "../data/gmd_values_mpitest2.txt", sep="\t", quote=FALSE)
write.table(gmd_values_jkitest1, "../data/gmd_values_jkitest1.txt", sep="\t", quote=FALSE)
write.table(gmd_values_jkitest2, "../data/gmd_values_jkitest2.txt", sep="\t", quote=FALSE)

write.table(gmd_values_mpi_field_2011, "../data/gmd_values_mpi_field_2011.txt", sep="\t", quote=FALSE)
write.table(gmd_values_mpi_field_2012, "../data/gmd_values_mpi_field_2012.txt", sep="\t", quote=FALSE)
write.table(gmd_values_jki_field_2012, "../data/gmd_values_jki_field_2012.txt", sep="\t", quote=FALSE)
write.table(gmd_values_jki_field_2013, "../data/gmd_values_jki_field_2013.txt", sep="\t", quote=FALSE)
write.table(gmd_values_dethlingen_2011, "../data/gmd_values_dethlingen_2011.txt", sep="\t", quote=FALSE)

write.table(gmd_values_jki_shelter_2011, "../data/gmd_values_jki_shelter_2011.txt", sep="\t", quote=FALSE)
write.table(gmd_values_jki_shelter_2012, "../data/gmd_values_jki_shelter_2012.txt", sep="\t", quote=FALSE)
write.table(gmd_values_mpi_pruef1, "../data/gmd_values_mpi_pruef1.txt", sep="\t", quote=FALSE)
write.table(gmd_values_mpi_pruef2, "../data/gmd_values_mpi_pruef2.txt", sep="\t", quote=FALSE)
write.table(gmd_values_mpi_pruef3, "../data/gmd_values_mpi_pruef3.txt", sep="\t", quote=FALSE)

write.table(gmd_values_breeder_trials, "../data/gmd_values_breeder_trials.txt", sep="\t", quote=FALSE)

# VALDIS TROST2
write.table(gmd_values_jki_shelter_2014, "../data/gmd_values_jki_shelter_2014.txt", sep="\t", quote=FALSE)
write.table(gmd_values_mpi_fgh_2014, "../data/gmd_values_mpi_fgh_2014.txt", sep="\t", quote=FALSE)
```



### sessionInfo
```{r sessionInfo}
save.image("gmd_data.RData")
sessionInfo()
```