---
title: GMD metadata for all TROST experiments
author: "Heike Sprenger"
date: "July 14, 2014"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---

## Set working directory
```{r set working directory}
# setwd("D:/work/repos/trost_metabolomics")
# setwd("~/work/repos/trost_metabolomics")
```

[solution for issue with working directory and knitr](https://github.com/yihui/knitr/issues/277)

## Load workspace, packages and scripts
```{r load workspace, message=FALSE,warning=FALSE}
# load packages
library(knitr)
library(ggplot2)
library(reshape)
library(reshape2)
library(pander)
library(plyr)
library(RODBC)
library(RMySQL)
library(yaml)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# options(width=600)

# load workspace
# load("gmd_data.RData")
```


## Source R functions
```{r source R functions, include=FALSE}
source("../functions/func_get_experiment_id.R")
source("../functions/func_get_gmd_metadata.R")
```


## Set up database connection to GMD
[How to access MS-SQL databases in linux](http://wiki.mpimp-golm.mpg.de/index.php/Access_MSSQL_databases)
[How to identify operating system from R](http://conjugateprior.org/2015/06/identifying-the-os-from-r/)

```{r set up database connection to GMD}

login_gmd <- yaml.load_file("../functions/login_gmd.yaml")

if (.Platform$OS.type == 'windows'){
  dbhandle <- odbcDriverConnect('driver={SQL Server}; 
                                server=r710sqljahu; 
                                database=GMD; 
                                trusted_connection=true')
} else { # if .Platform$OS.type == 'unix'
  dbhandle <- odbcDriverConnect(
    connection=paste0("server=",login_gmd$host,
                      ";database=",login_gmd$db,
                      ";uid=",login_gmd$user,
                      ";pwd=",login_gmd$passwd,
                      ";Port=1433;driver=FreeTDS;TDS_Version=8.0;"))
}
```


## Read tables: ``trost_TagList`` and ``phenotyper_result_joined_with_gmd_ids``

* ``trost_TagList`` shows all experiments that belong to the TROST/VALDIS project and comes from ``gmd_first_steps.Rmd``
* ``phenotyper_result_joined_with_gmd_ids`` contains meta information about samples and comes from ``phenotyper_metadata.Rmd``

```{r read tables}
trost_TagList <- read.table("data/trost_TagList.txt", sep="\t", header=TRUE)

# contains also valdis (2014 and 2015 data)
phenotyper_result_joined_with_gmd_ids <- read.table("data/phenotyper_result_joined_with_gmd_ids.txt", sep="\t", header=TRUE)
```


## Try out to get correct paste command
```{r try out to get correct paste command}
#var1 <- "trost"
#test <- sqlQuery(dbhandle, paste("select * from tf.TagList where comment=","\'", var1, "\'", sep=""))
```


## Get experiment_id
```{r get experiment id}
# try out function
experiment_id <- func_get_experiment_id("MPI_Feld_2011")
```


## Get GMD metadata
**Function with SQL query for GMD metadata concerning Sorbitol values (Internal Standard = IS), Average Annotated, dry weight and fresh weight.**
```{r get GMD metadata}
sort(trost_TagList$experiment)

# combination of both functions
gmd_meta_mpi_test_1_2 <- func_get_gmd_metadata( func_get_experiment_id("MPI_Test1_2") , 
                                                 phenotyper_result_joined_with_gmd_ids)
gmd_meta_mpi_test_2 <- func_get_gmd_metadata( func_get_experiment_id("MPI_Test2") , 
                                               phenotyper_result_joined_with_gmd_ids)
gmd_meta_jki_test_1 <- func_get_gmd_metadata( func_get_experiment_id("JKI_Test1") , 
                                               phenotyper_result_joined_with_gmd_ids)
gmd_meta_jki_test_2 <- func_get_gmd_metadata( func_get_experiment_id("JKI_Test2") , 
                                               phenotyper_result_joined_with_gmd_ids)

gmd_meta_mpi_field_2011 <- func_get_gmd_metadata_all( func_get_experiment_id("MPI_Feld_2011") , 
                                                       phenotyper_result_joined_with_gmd_ids)
gmd_meta_mpi_field_2012 <- func_get_gmd_metadata_all( func_get_experiment_id("MPI_Feld_2012") , 
                                                       phenotyper_result_joined_with_gmd_ids)
gmd_meta_jki_field_2012 <- func_get_gmd_metadata( func_get_experiment_id("JKI_Feld_2012") , 
                                                   phenotyper_result_joined_with_gmd_ids)
#gmd_meta_jki_field_2013 <- func_get_gmd_metadata( func_get_experiment_id("JKI_Feld_2013") , phenotyper_result_joined_with_gmd_ids)
gmd_meta_dethlingen_2011 <- func_get_gmd_metadata( func_get_experiment_id("Dethlingen_2011") , 
                                                    phenotyper_result_joined_with_gmd_ids)

gmd_meta_jki_shelter_2011 <- func_get_gmd_metadata( func_get_experiment_id("JKI_Shelter_2011") , 
                                                     phenotyper_result_joined_with_gmd_ids)
gmd_meta_jki_shelter_2012 <- func_get_gmd_metadata( func_get_experiment_id("JKI_Shelter_2012") , 
                                                     phenotyper_result_joined_with_gmd_ids)
gmd_meta_mpi_pruef_1 <- func_get_gmd_metadata( func_get_experiment_id("MPI_Pruef1") , 
                                                phenotyper_result_joined_with_gmd_ids)
gmd_meta_mpi_pruef_2 <- func_get_gmd_metadata( func_get_experiment_id("MPI_Pruef2") , 
                                                phenotyper_result_joined_with_gmd_ids)
gmd_meta_mpi_pruef_3 <- func_get_gmd_metadata( func_get_experiment_id("MPI_Pruef3") , 
                                                phenotyper_result_joined_with_gmd_ids)

gmd_meta_mpi_pruef_1 <- gmd_meta_mpi_pruef_1[-178,]
gmd_meta_mpi_pruef_2 <- gmd_meta_mpi_pruef_2[-48,]

#gmd_meta_breeder_trials <- func_get_gmd_metadata( func_get_experiment_id("breeder_trials") , phenotyper_result_joined_with_gmd_ids)

# VALDIS TROST 2014 and 2015
# 2014
gmd_meta_jki_shelter_2014 <- func_get_gmd_metadata_all( func_get_experiment_id("JKI_Shelter_2014"), 
                                                         phenotyper_result_joined_with_gmd_ids )
gmd_meta_mpi_fgh_2014 <- func_get_gmd_metadata_all( func_get_experiment_id("MPI_FGH_2014"), 
                                                     phenotyper_result_joined_with_gmd_ids)
# 2015
gmd_meta_mpi_fgh_2015 <- func_get_gmd_metadata_all( func_get_experiment_id("MPI_FGH_2015"), 
                                                     phenotyper_result_joined_with_gmd_ids)
gmd_meta_mpi_field_2015 <- func_get_gmd_metadata_all( func_get_experiment_id("MPI_Feld_2015"), 
                                                     phenotyper_result_joined_with_gmd_ids)
gmd_meta_jki_shelter_2015 <- func_get_gmd_metadata_all( func_get_experiment_id("JKI_Shelter_2015"), 
                                                         phenotyper_result_joined_with_gmd_ids )
```


## Save tables with GMD metadata
```{r save tables with GMD metadata}
write.table(gmd_meta_mpi_test_1_2, "data/gmd_meta_mpi_test_1_2.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_test_2, "data/gmd_meta_mpi_test_2.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_jki_test_1, "data/gmd_meta_jki_test_1.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_jki_test_2, "data/gmd_meta_jki_test_2.txt", sep="\t", quote=FALSE)

write.table(gmd_meta_mpi_field_2011, "data/gmd_meta_mpi_field_2011.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_field_2012, "data/gmd_meta_mpi_field_2012.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_jki_field_2012, "data/gmd_meta_jki_field_2012.txt", sep="\t", quote=FALSE)
#write.table(gmd_meta_jki_field_2013, "data/gmd_meta_jki_field_2013.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_dethlingen_2011, "data/gmd_meta_dethlingen_2011.txt", sep="\t", quote=FALSE)

write.table(gmd_meta_jki_shelter_2011, "data/gmd_meta_jki_shelter_2011.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_jki_shelter_2012, "data/gmd_meta_jki_shelter_2012.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_pruef_1, "data/gmd_meta_mpi_pruef_1.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_pruef_2, "data/gmd_meta_mpi_pruef_2.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_pruef_3, "data/gmd_meta_mpi_pruef_3.txt", sep="\t", quote=FALSE)

write.table(gmd_meta_jki_shelter_2014, "data/gmd_meta_jki_shelter_2014.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_fgh_2014, "data/gmd_meta_mpi_fgh_2014.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_fgh_2015, "data/gmd_meta_mpi_fgh_2015.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_mpi_field_2015, "data/gmd_meta_mpi_field_2015.txt", sep="\t", quote=FALSE)
write.table(gmd_meta_jki_shelter_2015, "data/gmd_meta_jki_shelter_2015.txt", sep="\t", quote=FALSE)
```


## Save workspace
```{r save workspace}
save.image("gmd_data.RData")
#sessionInfo()
```