---
title: Merge GC-MS data from 2 MPI greenhouse experiments
author: "Heike Sprenger"
date: "March 14, 2016"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---


# Set pander and knitr options
```{r set pander and knitr options, message=FALSE}
library(knitr)
library(pander)
library(colorRamps)
library(reshape)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width = 5, fig.height = 5, cache = F, 
               highlight = T, fig.show = "asis")
opts_knit$set(root.dir = '../')

#setwd("~/work/repos/trost_metabolomics")
#setwd("X:/mpimp/repos/trost_metabolomics")

#load("merge_mpi_test_rewatering.RData")
```


# Load packages and functions for metabolite data analysis
```{r load packages and functions for metabolite data analysis, message=FALSE, fig.show='hide'}
# load packages
source("../functions/func_load_packages_metabolite_data_analysis.R")

# load functions
source("../functions/func_load_functions_metabolite_data_analysis.R")
```


# Load data
## Load analytes table
this table comes from ``gmd_analytes.Rmd``
```{r load analytes table}
analytes_sel_exp_sort_all <- read.table("data/analytes_table_2exp_overlap_select_ordered.txt", 
                                    sep = "\t", header = T, allowEscapes = T)
```


## Remove duplicate analytes
```{r remove duplicate analytes}
duplicate_analytes <- c("Phenylalanine", "Lysine", "Serine", "Proline", "Valine", 
                        "Spermidine", "Glucose-6-phosphate", "Asparagine", "Psicose")

duplicate_analytes_idx <- which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes)

remove_analytes_idx <- c(36, 143, 95, 155, 82, 145, 25, 63, 138)

analytes_sel_exp_sort <- analytes_sel_exp_sort_all[-remove_analytes_idx,]

write.table(analytes_sel_exp_sort, "output/merge_mpi_test_rewatering/analytes_sel_exp_sort_final.txt", sep = "\t")

# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[1]) # Phenylalanine
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[2]) # Lysine
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[3]) # Serine
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[4]) # Proline
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[5]) # Valine
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[6]) # Spermidine
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[7]) # Glucose-6-phosphate
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[8]) # Asparagine
# which(analytes_sel_exp_sort_all$Name %in% duplicate_analytes[9]) # Psicose
# 
# plot(values[,5], values[,36], pch=19) # Phenylalanine
# res_anova_treatment_sample_time[c(5,36),]
# 
# plot(values[,114], values[,143], pch=19) # Lysine
# res_anova_treatment_sample_time[c(114,143),]
# values_norm_median_rev_t_log2FC[c(114,143),]
# 
# plot(values[,95], values[,147], pch=19) # Serine
# res_anova_treatment_sample_time[c(95,147),]
# values_norm_median_rev_t_log2FC[c(95,147),]
# 
# plot(values[,141], values[,155], pch=19) # Proline
# res_anova_treatment_sample_time[c(141,155),]
# values_norm_median_rev_t_log2FC[c(141,155),]
# 
# plot(values[,82], values[,129], pch=19) # Valine
# res_anova_treatment_sample_time[c(82,129),]
# values_norm_median_rev_t_log2FC[c(82,129),]
# 
# plot(values[,96], values[,145], pch=19) # Spermidine
# res_anova_treatment_sample_time[c(96,145),]
# values_norm_median_rev_t_log2FC[c(96,145),]
# 
# plot(values[,25], values[,33], pch=19) # Glucose-6-phosphate
# res_anova_treatment_sample_time[c(25,33),]
# values_norm_median_rev_t_log2FC[c(25,33),]
# 
# plot(values[,63], values[,171], pch=19) # Asparagine
# res_anova_treatment_sample_time[c(63,171),]
# values_norm_median_rev_t_log2FC[c(63,171),]
# 
# plot(values[,138], values[,144], pch=19) # Psicose
# res_anova_treatment_sample_time[c(138,144),]
# values_norm_median_rev_t_log2FC[c(138,144),]
```


## Load sample information
these tables come from ``gmd_data_analysis.Rmd``
```{r load sample information}
mpi_test_1_factors <- read.table("output/mpi_test_1_2_rewatering/factors.txt", 
                                     sep = "\t", header = T, row.names = 1)

mpi_test_2_factors <- read.table("output/mpi_test_2_rewatering/factors.txt", 
                                     sep = "\t", header = T, row.names = 1)
```


## Load data: log-transformed, WITH NAs, selected samples and selected overlapping analytes (164 of 173)
```{r load data with NAs}
mpi_test_1_values_all <- read.table("output/mpi_test_1_2_rewatering/gmd_values_select_log10.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)

mpi_test_2_values_all <- read.table("output/mpi_test_2_rewatering/gmd_values_select_log10.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)

mpi_test_1_values <- mpi_test_1_values_all[,-remove_analytes_idx]
mpi_test_2_values <- mpi_test_2_values_all[,-remove_analytes_idx]

sum(is.na(mpi_test_1_values))
sum(is.na(mpi_test_2_values))
```


## Load data: log-transformed, WITHOUT NAs, selected samples and selected overlapping analytes (164 of 173)
```{r load data witout NAsh}
mpi_test_1_values_woNA_all <- read.table("output/mpi_test_1_2_rewatering/gmd_values_select_log10_woNA.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)

mpi_test_2_values_woNA_all <- read.table("output/mpi_test_2_rewatering/gmd_values_select_log10_woNA.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)

mpi_test_1_values_woNA <- mpi_test_1_values_woNA_all[,-remove_analytes_idx]
mpi_test_2_values_woNA <- mpi_test_2_values_woNA_all[,-remove_analytes_idx]

sum(is.na(mpi_test_1_values_woNA))
sum(is.na(mpi_test_2_values_woNA))
```


# Combine data
## Combine sample information tables by rows
```{r combine sample information tables}
# all samples
factors <- rbind(mpi_test_1_factors, mpi_test_2_factors)
dim(factors)

length(levels(factors$BatchID)) 
# 4
table(factors$BatchID)

levels(factors$treatment)
# "control" "drought stress"

levels(factors$sample_time)

# reorder cultivar factor (use: check_names_ordered)
factors$genotype_name <- factor(factors$genotype_name, levels = check_names_ordered)
levels(factors$genotype_name)

# reorder sample_time factor (use: check_names_ordered)
factors$sample_time <- factor(factors$sample_time, levels = names_sample_time)
levels(factors$sample_time)

# create vector for trial name
factors$trial <- as.factor(c(rep("mpi_test_1_2", 96),
                             rep("mpi_test_2", 95)))

# create vector for tolerance
factors$tolerance <- func_create_tolerance_factor(factors)
table(factors$tolerance, factors$genotype_name)

# interaction of sample_time and cultivar
factors$sample_time_cultivar <- interaction(factors$sample_time, factors$genotype_name, sep = "-")
levels(factors$sample_time_cultivar)

write.table(factors, "output/merge_mpi_test_rewatering/factors.txt", sep = "\t")
```


## Combine value tables by rows 
```{r combine value tables}
# only late timepoint, check cultivars
values <- rbind(mpi_test_1_values, mpi_test_2_values)
dim(values)
# 191 164

values_woNA <- rbind(mpi_test_1_values_woNA, mpi_test_2_values_woNA)
dim(values_woNA)
```


# Calculate percentage of NAs and plot histograms
```{r calculate percentage of NAs}

func_print_na_statistics(values, analytes_sel_exp_sort)

pdf("figures/merge_mpi_test_rewatering/NAs_hist.pdf")
func_plot_na_statistics(values)
dev.off()
```


# Normalization: apply removeFactors function
```{r apply removeFactors function}
values_norm <- func_normalize(values, factors, 
                              facs=c("genotype_name", "treatment", "sample_time","SequenceID", "BatchID", "log10_AvgAnnotated"),
                              keep=c("genotype_name", "treatment", "sample_time"))

values_norm_woNA <- func_normalize(values_woNA, factors, 
                              facs=c("genotype_name", "treatment", "sample_time","SequenceID", "BatchID", "log10_AvgAnnotated"),
                              keep=c("genotype_name", "treatment", "sample_time"))

write.table(values_norm, "output/merge_mpi_test_rewatering/values_norm.txt", sep="\t")
write.table(values_norm, "output/merge_mpi_test_rewatering/values_woNA_norm.txt", sep="\t")
```


# Look for outliers after normalization of 3 merged data sets
```{r outlier detection}
pdf("figures/merge_mpi_test_rewatering/outlier_hist.pdf")
func_hist_outlier(values_norm, threshold=1e-6)
dev.off()

pdf("figures/merge_mpi_test_rewatering/outlier_hist_woNA.pdf")
func_hist_outlier(values_norm_woNA, threshold=1e-6)
dev.off()

# keep possible outliers for now


# remove outlier
values_norm_woOutlier <- func_replace_outlier(values_norm, threshold = 1e-6, 
                                                    original_values = values_norm, 
                                                    output = "normalized")

values_norm_woNA_woOutlier <- func_replace_outlier(values_norm_woNA, threshold = 1e-6, 
                                                    original_values = values_norm_woNA, 
                                                    output = "normalized")

sum(is.na(values_norm)) # 1835
sum(is.na(values_norm_woOutlier)) # 1850
# difference of 15

# USED FOR SUPPLEMENTAL DATA OF PAPER #
write.table(values_norm_woOutlier, "output/merge_mpi_test_rewatering/values_norm_log10_woOutlier.txt", sep = "\t")

sum(is.na(values_norm_woNA)) # 0
sum(is.na(values_norm_woNA_woOutlier)) # 21
```


#### check normal distribution --> shapiro test and histogram
```{r check normal distribution}
### plot histograms of normalized values per analyte
pdf("figures/merge_mpi_test_rewatering/values_norm_hist.pdf")
func_plot_dist(values_norm_woOutlier)
dev.off()

pdf("figures/merge_mpi_test_rewatering/values_norm_woNA_hist.pdf")
func_plot_dist(values_norm_woNA_woOutlier)
dev.off()

# shapiro test for normal distribution per analyte
res_shapiro <- func_shapiro_test(values_norm_woOutlier, 1e-6)
res_shapiro_woNA <- func_shapiro_test(values_norm_woNA_woOutlier, 1e-6)
```


# Perform PCA to check effect of normalization
## Prep for PCA
```{r prep for PCA}
# BEFORE normalization
pca_pareto_rnipals <- func_prep_pca(values_woNA, scale_method = "pareto", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")

pca_none_rnipals <- func_prep_pca(values_woNA, scale_method = "none", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")


# AFTER normalization
pca_norm_pareto_rnipals <- func_prep_pca(values_norm_woNA_woOutlier, scale_method = "pareto", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")

pca_norm_none_rnipals <- func_prep_pca(values_norm_woNA_woOutlier, scale_method = "none", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")
```


## PCA plots all trials
```{r PCA plot all trials}
pdf("figures/merge_mpi_test_rewatering/PCA_scoresplots_normalization_effect.pdf")
par(mar=c(4.5, 4.5, 2, 0.5))
# trial
palette( brewer.pal(3, "Dark2"))
func_5pairs_plot(pca_none_rnipals, factors, "trial", 19, "trial effect, before normalization")
func_5pairs_plot(pca_norm_none_rnipals, factors, "trial", 19, "trial effect, after normalization")
func_5pairs_plot(pca_norm_pareto_rnipals, factors, "trial", 19, "trial effect, after normalization (pareto scaling)")

# cultivar
palette(cols_cultivar_check)
func_5pairs_plot(pca_none_rnipals, factors, "genotype_name", 19, "cultivar effect, before normalization")
func_5pairs_plot(pca_norm_none_rnipals, factors, "genotype_name", 19, "cultivar effect, after normalization")
func_5pairs_plot(pca_norm_pareto_rnipals, factors, "genotype_name", 19, "cultivar effect, after normalization (pareto scaling)")

# treatment
palette(cols_treatment)
func_5pairs_plot(pca_none_rnipals, factors, "treatment", 19, "treatment effect, before normalization")
func_5pairs_plot(pca_norm_none_rnipals, factors, "treatment", 19, "treatment effect, after normalization")
func_5pairs_plot(pca_norm_pareto_rnipals, factors, "treatment", 19, "treatment effect, after normalization (pareto scaling)")

# trial
palette( brewer.pal(3, "Dark2"))
func_pca_plot_sym(pca_res = pca_none_rnipals, dim1 = 1, dim2 = 2, factors = factors, 
                  color_factor = "trial", symbols = c(19,17), symbol_factor = "treatment", 
                  symbol_size = 1.5, pos1 = "topleft", leg1 = 0.8, pos2 = "topright", leg2 = 0.8, 
                  maintext = "before normalization", legend.text1 = c("MPI test 1.2", "MPI test 2"))

func_pca_plot_sym(pca_norm_none_rnipals, 1, 2, factors, "trial", c(19,17), 
              "treatment", 1.5, "bottomleft", 0.8, "topright", 0.8, "after normalization",
              legend.text1 = c("MPI test 1.2", "MPI test 2"))

# cultivar
palette(cols_cultivar_check)
func_pca_plot(pca_norm_none_rnipals, 1, 2, factors, "genotype_name", 19, 1.5, "bottomleft", 1)
func_pca_plot(pca_norm_none_rnipals, 2, 3, factors, "genotype_name", 19, 1.5, "bottomright", 1)
func_pca_plot(pca_norm_pareto_rnipals, 2, 3, factors, "genotype_name", 19, 1.5, "bottomleft", 1)

# treatment
palette(cols_treatment)
func_pca_plot(pca_norm_none_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1)
func_pca_plot(pca_norm_pareto_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1,
              maintext = "treatment effect")

dev.off()
```


## PCA plots: cultivar effect
```{r PCA plots: cultivar effect}
palette(cols_cultivar_check)

pdf("figures/merge_mpi_test_rewatering/PCA_scoresplots_cultivar_effect.pdf",6,6)

func_5pairs_plot(pca_norm_pareto_rnipals, factors, "genotype_name", 19, "cultivar effect")
func_5pairs_plot_sym(pca_norm_pareto_rnipals, factors, "genotype_name", c(19,17), "treatment", "cultivar / treatment effect")

# PC1 vs PC2
func_pca_plot_sym(pca_norm_pareto_rnipals, 1, 2, factors, "genotype_name", c(19,17), 
              "treatment", 1.5, "bottomright", 0.8, "bottomleft", 0.8, "cultivar / treatment effect")
# PC1 vs PC3
func_pca_plot_sym(pca_norm_pareto_rnipals, 1, 3, factors, "genotype_name", c(19,17), 
              "treatment", 1.5, "bottomright", 0.8, "topright", 0.8, "cultivar / treatment effect")
# PC2 vs PC3
func_pca_plot_sym(pca_norm_pareto_rnipals, 2, 3, factors, "genotype_name", c(19,17), 
              "treatment", 1.5, "bottomleft", 0.8, "topleft", 0.8, "cultivar / treatment effect")

func_pca_plot(pca_norm_pareto_rnipals, 2, 3, factors, "genotype_name", 19, 1.5, "bottomleft", 1, 
              maintext = "cultivar effect")

dev.off()

#######################################

# for Phd-thesis: cultivar effect, with treatment as symbol
#pdf("../../Doktorarbeit/figures/gcms_all_field_cultivar_treatment.pdf", width=6, height=6)
palette(cols_cultivar2)
par(mar=c(4.5, 4.5, 0.5, 0.5))
func_pca_plot_sym(pca_norm_none_rnipals, 1, 2, factors, "genotype_name", c(19,17), 
              "treatment", 2, "bottomleft", 1.2, "topleft", 1.2)
#dev.off()

# for Publication
#pdf("../../TORST/Manuskripte/tba/gcms_all_field_cultivar_treatment_new.pdf", width=6, height=6)
palette(cols_treatment)
par(mar=c(4.5, 4.5, 1.5, 0.5))
func_pca_plot_sym(pca_norm_none_rnipals, 1, 2, factors, "treatment", 
                  symbols = c(15, 16, 17, 18), 
                  "genotype_name", 2, "topleft", 1.2, "topright", 1.2, 
                  ymin = -3.5, ymax = 3.5)
#dev.off()


#png("../../TROST/Manuskripte/tba/gcms_all_field_cultivar_treatment_new.png", width=2000, height=2000, res=300)
palette(cols_treatment)
par(mar=c(4.5, 4.5, 1.5, 0.5))
func_pca_plot_sym(pca_norm_none_rnipals, 1, 2, factors, "treatment", 
                  symbols = c(15, 16, 17, 18), 
                  "genotype_name", symbol_size = 2, "bottomleft", 1.5, 
                  "topright", 1.5, ymin = -3.5, ymax = 3.5)
#mtext('C', side=3, at=-2.7, line=0, outer=F, cex=1.5)
#dev.off()
```


## PCA plots: treatment and sample time effect
```{r PCA plots: treatment and sample time effect}
# sample time
pdf("figures/merge_mpi_test_rewatering/PCA_scoresplots_sample_time_effect.pdf")
palette(cols_sample_time)
func_5pairs_plot(pca_norm_pareto_rnipals, factors, 
                 "sample_time", 19, "sampling time effect (no scaling, rnipals)")

func_pca_plot(pca_norm_pareto_rnipals, 1, 2, factors, "sample_time", 19, 1.5, 
              "bottomleft", 0.8, maintext = "sampling time effect (pareto scaling, rnipals)")

func_pca_plot_sym(pca_res = pca_norm_pareto_rnipals, dim1 = 1, dim2 = 2, 
                  factors = factors, color_factor = "sample_time", 
                  symbol_size = 1.5, symbols = c(19,17), symbol_factor = "treatment", 
                  pos1 = "bottomleft", leg1 = 0.8, pos2 = "bottomright", leg2 = 0.8,
                  maintext = "sampling time x treatment effect (pareto scaling, rnipals)")

func_pca_plot_sym(pca_res = pca_norm_pareto_rnipals, dim1 = 1, dim2 = 3, 
                  factors = factors, color_factor = "sample_time", 
                  symbol_size = 1.5, symbols = c(19,17), symbol_factor = "treatment", 
                  pos1 = "bottomleft", leg1 = 0.8, pos2 = "topleft", leg2 = 0.8,
                  maintext = "sampling time x treatment effect (pareto scaling, rnipals)")
dev.off()
```



# Aggregate normalized values for heatmap
```{r aggregate normalized values}
# # for matrix with NAs using 4 factors
# # median
# values_norm_median <- func_agg_3fac(values_norm_woOutlier, factors, "genotype_name", "sample_time", "treatment", median, analytes_sel_exp_sort)
# dim(values_norm_median)
# # 32 samples (rows), 174 (1 sampleID, 173 analytes, columns)
# 
# sum(is.na(values_norm_median)) / (ncol(values_norm_median)*nrow(values_norm_median))*100
# # 1.04%
# 
# # export TRANSPOSED matrix for heatmap
# values_norm_median_t <- t(values_norm_median[,-1])
# colnames(values_norm_median_t) <- values_norm_median$sample
# dim(values_norm_median_t)
# # 173 analytes (rows), 32 samples (columns)
# 
# write.table(values_norm_median_t, 
#             "output/merge_mpi_test_rewatering/values_norm_log10_median_for_heatmap.txt", 
#             sep="\t", quote=FALSE, col.names=NA, row.names=TRUE)
```


# ANOVA, boxplots & t-test
## ANOVA for 2 factors: treatment * cultivar (with interaction)
```{r ANOVA with 2 factors (with interaction)}
res_anova_treatment_cultivar <- func_anova_2fac_ia(values_norm_woNA_woOutlier, factors, "treatment", "genotype_name", 0.01, analytes_sel_exp_sort$Name)
res_anova_treatment_sample_time <- func_anova_2fac_ia(values_norm_woNA_woOutlier, factors, 
                                                      "treatment", "sample_time", 0.01, 
                                                      analytes_sel_exp_sort$Name)

res_anova_treatment_cultivar_sig_treat <- which(res_anova_treatment_cultivar[,1] < 0.01)
res_anova_treatment_cultivar_sig_cul <- which(res_anova_treatment_cultivar[,2] < 0.01)
res_anova_treatment_cultivar_sig_both <- intersect (which(res_anova_treatment_cultivar[,1] < 0.01), 
                                                    which(res_anova_treatment_cultivar[,2] < 0.01))

write.table(res_anova_treatment_cultivar, "output/merge_mpi_test_rewatering/res_anova_treatment_cultivar.txt", sep="\t")
write.table(res_anova_treatment_sample_time, "output/merge_mpi_test_rewatering/res_anova_treatment_sample_time.txt", sep="\t")


res_anova_adj_ia_treat_time_cultivar <- func_anova_2fac_ia(normalized_values = values_norm_woNA_woOutlier, 
                                 trial_factors = factors, 
                                 factor1 = "treatment", 
                                 factor2 = "sample_time_cultivar",
                                 threshold = 0.01,
                                 analyte_names = analytes_sel_exp_sort$Name)
```


# Boxplots of normalized values per analyte using 2 factors
## Boxplots per gene: treatment  effect
```{r Boxplots per gene: treatment effect}
pdf("figures/merge_mpi_test_rewatering/boxplot_treatment.pdf")
func_boxplot_1fac(values_norm_woNA_woOutlier, 
                  factors, 
                  "treatment", 
                  res_anova_treatment_cultivar, 
                  cols_treatment, 
                  analytes_sel_exp_sort$Name)
dev.off()
```

## Boxplots per gene: treatment x cultivar effect
```{r boxplots of normalized values per analyte using 2 factors}
pdf("figures/merge_mpi_test_rewatering/boxplot_treatment_cultivar.pdf", width=7, height=8)
par(mar=c(7, 4.1, 7, 2.1))
func_boxplot_2fac(normalized_values = values_norm_woNA_woOutlier, 
                  trial_factors = factors, 
                  factor1 = "treatment", 
                  factor2 = "genotype_name", 
                  res_anova_adj = res_anova_treatment_cultivar, 
                  cols = cols_cultivar_treatment,
                  names_factors = names_treatment_cultivar_reordered,
                  analytes_sel_exp_sort$Name)

dev.off()
```

## Boxplots per gene: treatment x sample_time effect
```{r Boxplots per gene: treatment x sample_time effect}
pdf("figures/merge_mpi_test_rewatering/boxplot_treatment_sample_time.pdf", width=7, height=8)
par(mar=c(7, 4.1, 7, 2.1))
func_boxplot_2fac(normalized_values = values_norm_woNA_woOutlier, 
                  trial_factors = factors, 
                  factor1 = "treatment", 
                  factor2 = "sample_time", 
                  res_anova_adj = res_anova_treatment_sample_time, 
                  cols = cols_treatment_sample_time,
                  names_factors = names_treatment_sample_time_2,
                  analytes_sel_exp_sort$Name)

dev.off()
```


## Boxplots per gene: treatment x sample time x cultivar effect
```{r boxplots per gene: treatment x sample time x cultivar effect}
pdf("figures/merge_mpi_test_rewatering/boxplot_treatment_sample_time_cultivar.pdf", width=7, height=8)
par(mar=c(7, 4.1, 7, 2.1))
func_boxplot_2fac(normalized_values = values_norm_woNA_woOutlier, 
                  trial_factors = factors, 
                  factor1 = "treatment",
                  factor2 = "sample_time_cultivar",
                  res_anova_adj = res_anova_adj_ia_treat_time_cultivar, 
                  cols = cols_treatment_sample_time, 
                  analyte_names = analytes_sel_exp_sort$Name)
dev.off()
```


## Boxplots per gene: treatment x sample time x cultivar effect 2
```{r Boxplots per gene: treatment x sample time x cultivar effect 2}
data_info <- data.frame(factors, values_norm_woNA_woOutlier, check.names = F)

met <- as.character(analytes_sel_exp_sort$analyteID)
met <- gsub("-101", "", met)

colnames(data_info)[18:181] <- met

met_name <- as.character(analytes_sel_exp_sort$Name)
names(met_name) <- met

pdf("figures/merge_mpi_test_rewatering/boxplot_treatment_sample_time_cultivar2.pdf", width = 8, height = 6)
for(i in met){
  p <- ggplot(data_info, aes_string(x = "sample_time", y = i, fill = "treatment")) + 
    geom_boxplot() + ggtitle(met_name[i]) +
    facet_wrap(~ genotype_name, ncol = 2) + 
    scale_fill_manual(values=cols_treatment)
  plot(p)}
dev.off()

pdf("figures/merge_mpi_test_rewatering/boxplot_treatment_sample_time_cultivar3.pdf", width = 8, height = 6)
for(i in met){
  p <- ggplot(data_info, aes_string(x = "genotype_name", y = i, fill = "treatment")) + 
    geom_boxplot() + ggtitle(met_name[i]) +
    facet_wrap(~ sample_time, ncol = 2) + 
    scale_fill_manual(values=cols_treatment)
  plot(p)}
dev.off()
```


# t-test
## t-test: control vs. stress, separately for sample_time
```{r t-test control vs. stress}
res_ttest_treatment <- func_ttest_treatment_all(values_norm_woNA_woOutlier, factors, 0.01)
# 119

res_ttest_early_before <- func_ttest_treatment(values_norm_woNA_woOutlier, factors, fac1 = "sample_time", val1 = "early/before", 0.01)
res_ttest_early_after <- func_ttest_treatment(values_norm_woNA_woOutlier, factors, fac1 = "sample_time", val1 = "early/after", 0.01)
res_ttest_late_before <- func_ttest_treatment(values_norm_woNA_woOutlier, factors, fac1 = "sample_time", val1 = "late/before", 0.01)
res_ttest_late_after <- func_ttest_treatment(values_norm_woNA_woOutlier, factors, fac1 = "sample_time", val1 = "late/after", 0.01)

# res_ttest_alegria <- func_ttest_treatment(values_norm, factors, fac1 = "genotype_name", val1 = "Alegria", 0.01)
# res_ttest_milva <- func_ttest_treatment(values_norm, factors, fac1 = "genotype_name", val1 = "Milva", 0.01)
# res_ttest_desiree <- func_ttest_treatment(values_norm, factors, fac1 = "genotype_name", val1 = "Desiree", 0.01)
# res_ttest_saturna <- func_ttest_treatment(values_norm, factors, fac1 = "genotype_name", val1 = "Saturna", 0.01)

res_ttest_idx <- order(res_ttest_treatment)
head(analytes_sel_exp_sort[res_ttest_idx, "Name"], n=20)

# union across all sample_times
union_sample_time <- sort(union(union(which(res_ttest_early_before < 0.01), 
                                      which(res_ttest_early_after < 0.01)), 
                                union(which(res_ttest_late_before < 0.01), 
                                      which(res_ttest_late_after < 0.01))))
length(union_sample_time)
# 137

length( intersect(which(res_ttest_treatment < 0.01), union_sample_time)) # 112
length( setdiff(which(res_ttest_treatment < 0.01), union_sample_time)) # 7
length( setdiff(union_sample_time, which(res_ttest_treatment < 0.01))) # 25

res_ttest_treatment_specific <- setdiff(which(res_ttest_treatment < 0.01), union_sample_time)

length( union(which(res_ttest_treatment < 0.01), union_sample_time)) # 144
```


## t-test: control vs. stress, separately for sample_time and cultivar
```{r t-test: control vs. stress, separately for sample_time and cultivar}
res_ttest <- matrix(NA, ncol = 16, nrow = 164)
colnames(res_ttest) <- levels(factors$sample_time_cultivar)
rownames(res_ttest) <- analytes_sel_exp_sort$Name

for(i in levels(factors$sample_time_cultivar)){
  
  res_ttest[,i] <- func_ttest_treatment(values_norm_woNA_woOutlier, factors, 
                                        fac1 = "sample_time_cultivar", val1 = i, 0.01)
}
```


# Reverse log10-transformation
```{r reverse log10-transformation}
values_norm_rev <- 10^(values_norm_woNA_woOutlier)
```


# Calculate log2FC: up or down regulated
** use log10 reverse transformed values for log2FC calculation **
```{r calculate log2FC}
# across all cultivars!
log2FC_treatment <- func_log2FC_1fac(values_norm_rev, factors, average_func = median)

# per cultivar
log2FC_early_before <- func_log2FC_2fac(values_norm_rev, factors, average_func = median,
                                   fac1 = "sample_time", fac1_val = "early/before",
                                   fac2 = "treatment", fac2_val1 = "control", fac2_val2 = "drought stress")

log2FC_early_after <- func_log2FC_2fac(values_norm_rev, factors, average_func = median,
                                   fac1 = "sample_time", fac1_val = "early/after",
                                   fac2 = "treatment", fac2_val1 = "control", fac2_val2 = "drought stress")

log2FC_late_before <- func_log2FC_2fac(values_norm_rev, factors, average_func = median,
                                   fac1 = "sample_time", fac1_val = "late/before",
                                   fac2 = "treatment", fac2_val1 = "control", fac2_val2 = "drought stress")

log2FC_late_after <- func_log2FC_2fac(values_norm_rev, factors, average_func = median,
                                   fac1 = "sample_time", fac1_val = "late/after",
                                   fac2 = "treatment", fac2_val1 = "control", fac2_val2 = "drought stress")
```

## Up or down regulated
```{r up or down regulated}
# get a factor with the value up or down for all analytes
log2FC_up_down_treatment <- func_log2FC_up_down(log2FC_treatment)

log2FC_up_down_early_before <- func_log2FC_up_down(log2FC_early_before)
log2FC_up_down_early_after <- func_log2FC_up_down(log2FC_early_after)
log2FC_up_down_late_before <- func_log2FC_up_down(log2FC_late_before)
log2FC_up_down_late_after <- func_log2FC_up_down(log2FC_late_after)

# get a list with ALL analytes that are up or down
log2FC_dir_early_before <- func_log2FC_dir(log2FC_early_before)
log2FC_dir_early_after <- func_log2FC_dir(log2FC_early_after)
log2FC_dir_late_before <- func_log2FC_dir(log2FC_late_before)
log2FC_dir_late_after <- func_log2FC_dir(log2FC_late_after)

# get a list with significantly changed analytes that are up or down
log2FC_dir_sig_early_before <- func_log2FC_dir_sig(log2FC_dir_early_before, res_ttest_early_before, 0.01)
log2FC_dir_sig_early_after <- func_log2FC_dir_sig(log2FC_dir_early_after, res_ttest_early_after, 0.01)
log2FC_dir_sig_late_before <- func_log2FC_dir_sig(log2FC_dir_late_before, res_ttest_late_before, 0.01)
log2FC_dir_sig_late_after <- func_log2FC_dir_sig(log2FC_dir_late_after, res_ttest_late_after, 0.01)
```


## Volcano plot: log2FC vs. t-test p-value
```{r volcano plot: log2FC vs. t-test p-value}
pdf("figures/merge_mpi_test_rewatering/volcano_plots_sample_time.pdf", 6, 6)
par(mar=c(4.5, 4.5, 0.5, 0.5))

func_volcano_plot(res_ttest_early_before, log2FC_early_before)
func_volcano_plot(res_ttest_early_after, log2FC_early_after)
func_volcano_plot(res_ttest_late_before, log2FC_late_before)
func_volcano_plot(res_ttest_late_after, log2FC_late_after)

dev.off()
```


## Combine t-test results with log2FC
```{r combine t-test results with log2FC}
colnames(analytes_sel_exp_sort)
res_ttest_all <- cbind(analytes_sel_exp_sort[,c(3,4)],
                       res_ttest_early_before, log2FC_early_before, log2FC_up_down_early_before,
                       res_ttest_early_after, log2FC_early_after, log2FC_up_down_early_after, 
                       res_ttest_late_before, log2FC_late_before, log2FC_up_down_late_before,
                       res_ttest_late_after, log2FC_late_after, log2FC_up_down_late_after,
                       res_ttest_treatment)

write.table(res_ttest_all, "output/merge_mpi_test_rewatering/res_ttest_control_vs_stress.txt", sep="\t", col.names=NA, row.names=T)

res_ttest_p <- cbind(p_early_before = res_ttest_early_before, 
                     p_early_after = res_ttest_early_after,
                     p_late_before = res_ttest_late_before, 
                     p_late_after = res_ttest_late_after)

head(res_ttest_p)
rownames(res_ttest_p) <- analytes_sel_exp_sort$Name
write.table(res_ttest_p, "output/merge_mpi_test_rewatering/res_ttest_p.txt", sep="\t")
```


## Compare lists of up/down regulated analytes
```{r compare lists of up/down regulated analytes}
# compare Alegria and Milva
# UP
func_compare_2lists_up(log2FC_dir_sig_early_before, log2FC_dir_sig_early_after)
# DOWN
func_compare_2lists_down(log2FC_dir_sig_early_before, log2FC_dir_sig_early_after)


# compare Desiree and Saturna
# UP
func_compare_2lists_up(log2FC_dir_sig_late_before, log2FC_dir_sig_late_after)
# DOWN
func_compare_2lists_down(log2FC_dir_sig_late_before, log2FC_dir_sig_late_after)
```


## Indices of not-significantly regulated analytes
```{r indices of not-significantly regulated analytes}
not_sig_early <- intersect(which(res_ttest_early_before > 0.01), which(res_ttest_early_after > 0.01))
not_sig_late <- intersect(which(res_ttest_late_before > 0.01), which(res_ttest_late_after > 0.01))
not_sig <- intersect(not_sig_early, not_sig_late)
not_sig <- intersect(not_sig, which(res_ttest_treatment > 0.01))
length(not_sig)
# 20 metabolites are not changed at any timepoint, 144 remain
write.table(not_sig, "output/merge_mpi_test_rewatering/not_sig_regulated_analytes.txt", sep="\t")
```


# Aggregate normalized values for heatmap
## Aggregate normalized values for heatmap: per treatment and sample time
```{r Aggregate normalized values for heatmap: per treatment and sample time}
# for matrix with NAs using 2 factors (treatment, cultivar)
# MEDIAN
values_norm_median <- func_agg_2fac(values_norm_woNA_woOutlier, factors, 
                                    "sample_time", "treatment", median, 
                                    analytes_sel_exp_sort)
dim(values_norm_median)
# 8 samples (rows), 165 (1 sampleID, 164 analytes, columns)

# export TRANSPOSED matrix for heatmap
values_norm_median_t <- t(values_norm_median[,-1])
colnames(values_norm_median_t) <- values_norm_median$sample
dim(values_norm_median_t)
# 164 analytes (rows), 8 samples (columns)

head(values_norm_median_t)

write.table(values_norm_median_t, 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_for_heatmap.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)
```

## Aggregate normalized values for heatmap: per cultivar, treatment and sample time
```{r aggregate normalized values: per cultivar, treatment and sample time}
# for matrix with NAs using 3 factors (treatment, sample time, cultivar)
# MEDIAN
values_norm_median_3fac <- func_agg_3fac(values_norm_woNA_woOutlier, factors, 
                                         "sample_time", "genotype_name","treatment", 
                                         mean, #median
                                         analytes_sel_exp_sort)
dim(values_norm_median_3fac)
# 32 samples (rows), 165 (1 sample, 164 metabolites, columns)

# export TRANSPOSED matrix for heatmap
values_norm_median_3fac_t <- t(values_norm_median_3fac[,-1])
colnames(values_norm_median_3fac_t) <- values_norm_median_3fac$sample
dim(values_norm_median_3fac_t)
# 164 metabolites (rows), 32 samples (columns)

values_norm_median_3fac_t[1:5, 1:2]

write.table(values_norm_median_3fac_t, 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_for_heatmap_3fac.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)
```


# Reverse log-transformation
```{r reverse log-transformation}
values_norm_median_rev <- 10^(values_norm_median[,-1]) # without sample names
dim(values_norm_median_rev)
# 8 samples (rows), 164 analytes (columns)

values_norm_median_3fac_rev <- 10^(values_norm_median_3fac[,-1]) # without sample names
dim(values_norm_median_3fac_rev)

## samples are in rows, analytes in columns
## for transformation: samples should be in columns and analytes in rows --> row-wise median is median over all samples of 1 analyte

# transpose matrix
values_norm_median_rev_t <- t(values_norm_median_rev)
dim(values_norm_median_rev_t)
# 164 analytes (rows), 8 samples (columns)
colnames(values_norm_median_rev_t) <- values_norm_median$sample

values_norm_median_3fac_rev_t <- t(values_norm_median_3fac_rev)
dim(values_norm_median_3fac_rev_t)
colnames(values_norm_median_3fac_rev_t) <- values_norm_median_3fac$sample
```


# Calculation of log-median-ratio
```{r calculation of log-median-ratio}
values_norm_median_rev_transformed <- func_log_transform(values_norm_median_rev_t)
dim(values_norm_median_rev_transformed)
# 164 analytes, 8 samples

values_norm_median_3fac_rev_transformed <- func_log_transform(values_norm_median_3fac_rev_t)
dim(values_norm_median_3fac_rev_transformed)

# export matrix for heatmap
write.table(values_norm_median_rev_transformed, 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_rev_transformed_for_heatmap.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)

write.table(values_norm_median_3fac_rev_transformed, 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_3fac_rev_transformed_for_heatmap.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)
```


# Calculation of log2-fold-change
## Calculation of log2-fold-change: per sample time
```{r calculation of log2-fold-change: per sample time}
# manual calculation
colnames(values_norm_median_rev_t)[c(1,5)]
log2(values_norm_median_rev_t[1,5] / values_norm_median_rev_t[1,1])

colnames(values_norm_median_rev_t)[c(4,8)]
log2(values_norm_median_rev_t[1,8] / values_norm_median_rev_t[1,4])

values_norm_median_rev_t_log2FC <- log2_fold_change(values_norm_median_rev_t)
dim(values_norm_median_rev_t_log2FC)
range(values_norm_median_rev_t_log2FC, na.rm=TRUE)

# change colnames ("trial", "genotype_name", "sample_time")
colnames(values_norm_median_rev_t_log2FC) <- levels(factors$sample_time)

head(values_norm_median_rev_t_log2FC)

# export matrix for heatmap
write.table(values_norm_median_rev_t_log2FC, 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_rev_t_log2FC_for_heatmap.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)

write.table(values_norm_median_rev_t_log2FC, 
            "output/merge_mpi_test_rewatering/log2FC_treatment_per_sampletime.txt", 
            sep = "\t", quote = F, row.names = T)
```


## Calculation of log2-fold-change: per sample time and cultivar
```{r calculation of log2-fold-change: per sample time and cultivar}
# manual calculation
colnames(values_norm_median_3fac_rev_t)[c(1,17)]
log2(values_norm_median_3fac_rev_t[1,17] / values_norm_median_3fac_rev_t[1,1])

colnames(values_norm_median_3fac_rev_t)[c(4,20)]
log2(values_norm_median_3fac_rev_t[1,20] / values_norm_median_3fac_rev_t[1,4])

values_norm_median_3fac_t_log2FC <- log2_fold_change(values_norm_median_3fac_rev_t)
dim(values_norm_median_3fac_t_log2FC)
range(values_norm_median_3fac_t_log2FC, na.rm=TRUE)

# change colnames ("trial", "genotype_name", "sample_time")
colnames(values_norm_median_3fac_t_log2FC) <- gsub(".control", "", colnames(values_norm_median_3fac_t_log2FC))
head(values_norm_median_3fac_t_log2FC)

# export matrix for heatmap
write.table(values_norm_median_3fac_t_log2FC, 
            "output/merge_mpi_test_rewatering/log2FC_treatment_per_sampletime_cultivar.txt", 
            sep = "\t", quote = F, row.names = T)
```


# Change not significant values to NA
```{r change not significant values to NA}
dim(values_norm_median_rev_t_log2FC)
dim(res_ttest_p)

log2FC_sample_time_sig <- values_norm_median_rev_t_log2FC

for (i in 1:164){
  for (j in 1:4){
    if(is.na(res_ttest_p[i,j])){
      log2FC_sample_time_sig[i,j] <- NA
    }
    else if (res_ttest_p[i,j] > 0.01){
      log2FC_sample_time_sig[i,j] <- NA
    }
  }
}

log2FC_sample_time_sig [union_sample_time,]

# only 36 metabolites that changed per cultivar!
# write.table(log2FC_cultivars_sig[union_cultivar,], 
#             "../heatmaps/merge_mpi_test_rewatering/log2FC_cultivars_sig_for_heatmap.txt", 
#             sep = "\t", quote = F, col.names = NA, row.names = T)

###############

# log2FC per sample time and cultivar
log2FC_sample_time_cultivar_sig <- values_norm_median_3fac_t_log2FC

for (i in 1:164){
  for (j in 1:16){
    if(is.na(res_ttest[i,j])){
      log2FC_sample_time_cultivar_sig[i,j] <- NA
    }
    else if (res_ttest[i,j] > 0.01){
      log2FC_sample_time_cultivar_sig[i,j] <- NA
    }
  }
}
```


# Significant changes for heatmap
- p < 0.05 *
- p < 0.01 **
- p < 0.001 ***
```{r significant changes for heatmap}
res_ttest_p_m <- melt(res_ttest_p[union_sample_time,])
res_ttest_p_m$value3 <- cut(res_ttest_p_m$value,
                            breaks=c(-Inf, 0.001, 0.01),
                            label=c("***", "**")) 

signif2 <- matrix(res_ttest_p_m$value3, ncol=4)

##############

# for t-test of stress effect for sample time and cultivar
res_ttest_melt <- melt(res_ttest)

res_ttest_melt$value3 <- cut(res_ttest_melt$value,
                            breaks = c(-Inf, 0.001, 0.01, 0.05),
                            label = c("***", "**", "*")) 

signif_sample_time_cultivar <- matrix(res_ttest_melt$value3, ncol = 16)

res_ttest_melt_ <- res_ttest_melt[,-3]
colnames(res_ttest_melt_)[1:2] <- c("id","variable")
res_ttest_melt_cast <- cast(res_ttest_melt_, id ~ variable)
res_ttest_melt_cast <- res_ttest_melt_cast[analytes_sel_exp_sort$Name,]
```


# Heatmap
## heatmap of log2FC per sample time
```{r heatmap of log2FC per sample time}
mat_data <- as.matrix(values_norm_median_rev_t_log2FC[union_sample_time,])

distance = dist(mat_data, method = "euclidian")
cluster = hclust(distance, method = "average")

#mycluster <- cutree(cluster, h=max(cluster$height)/1.5)
mycluster <- cutree(cluster, k=4)
#mycluster[cluster$order]

mycolor <- brewer.pal(8, "Set2")
mycolor <- mycolor[as.vector(mycluster)]

my_palette <- colorRampPalette(c("blue", "white", "red"))(n = 299)

col_breaks = c(seq(-3,-0.81,length=100), # for blue
               seq(-0.8,0.8,length=100), # for white
               seq(0.81,3,length=100)) # for red

#pdf("../../Doktorarbeit/figures/heatmap_log2FC_signif_cultivars_field_new.pdf", width=6.5, height=10)
#pdf("figures/merge_mpi_test_rewatering/heatmap_log2FC_signif.pdf", width=6, height=12)
dpi <- 300
png("figures/merge_mpi_test_rewatering/heatmap_log2FC_signif_note.png", 
    width=9*dpi, height=18*dpi, res = dpi)

lmat=rbind(4:3, 2:1)
lhei=c(0.8, 4)
lwid=c(1, 4)

layout(mat = lmat, widths = lwid, heights = lhei)

heatmap.2(mat_data,
  cellnote = signif2,  # same data set for cell labels
  notecex=1.0, # numeric scaling factor for cellnote items
  colsep=0:ncol(mat_data), 
  rowsep=0:nrow(mat_data),
  sepwidth=c(0.01,0.05),
  sepcolor="black",
  main = "", # heat map title
  notecol="black",      # change font color of cell labels to black
  density.info="none",  # turns off density plot inside color legend
  trace="none",         # turns off trace lines inside the heat map
  margins =c(6,15),     # widens margins around plot
  keysize=1.5,
  cexRow=1.2,
  cexCol=1.5,
  labCol = c("early/\n before", "early/\n after", "late/\n before", "late/\n after"),
  lhei=c(0.6, 4),
  col=my_palette,       # use on color palette defined earlier 
  breaks=col_breaks,    # enable color transition at specified limits
  RowSideColors=mycolor,
  Rowv=as.dendrogram(cluster),
  dendrogram="row",     # only draw a row dendrogram
  Colv="NA",
  adjCol = c(1,0.5))            # turn off column clustering```

dev.off()
```

### heatmap reordered
```{r heatmap REORDERED}
mat_data <- as.matrix(values_norm_median_rev_t_log2FC[union_sample_time,])

distance <- dist(mat_data, method = "euclidian")
cluster <- hclust(distance, method = "average")

cluster2 <- dendsort(cluster, type = "average") 

#mycluster <- cutree(cluster, h=max(cluster$height)/1.5)
mycluster <- cutree(cluster2, k=4)
mycluster[cluster2$order]

mycolor <- brewer.pal(8, "Set2")
mycolor <- mycolor[as.vector(mycluster)]

my_palette <- colorRampPalette(c("blue", "white", "red"))(n = 299)

# col_breaks = c(seq(-1,-0.3,length=100), # for blue
#                seq(-0.3,0.3,length=100), # for white
#                seq(0.3,2,length=100)) # for red

col_breaks = c(seq(-3,-0.81,length=100), # for blue
               seq(-0.8,0.8,length=100), # for white
               seq(0.81,3,length=100)) # for red

# for Phd-thesis: cultivar effect, with treatment as symbol
#pdf("../../TROST/Manuskripte/tba/heatmap_log2FC_signif_cultivars_field_reordered_average_new_scale2.pdf", width=6, height=10)
pdf("figures/merge_mpi_test_rewatering/heatmap_log2FC_signif_reordered.pdf", width=6, height=12)
# dpi <- 300
# png("figures/merge_mpi_test_rewatering/heatmap_log2FC_signif_reordered.png", width=3*dpi, height=6*dpi, res = dpi)

lmat=rbind(4:3, 2:1)
lhei=c(0.8, 4)
lwid=c(1, 4)

layout(mat = lmat, widths = lwid, heights = lhei)

heatmap.2(mat_data,
  cellnote = signif2,  # same data set for cell labels
  notecex=1.0, # numeric scaling factor for cellnote items
  colsep=0:ncol(mat_data), 
  rowsep=0:nrow(mat_data),
  sepwidth=c(0.01,0.05),
  sepcolor="black",
  main = "", # heat map title
  notecol="black",      # change font color of cell labels to black
  density.info="none",  # turns off density plot inside color legend
  trace="none",         # turns off trace lines inside the heat map
  margins =c(6,20),     # widens margins around plot
  keysize=1.5,
  cexRow=0.8,
  cexCol=1.3,
  labCol = c("early/\n before", "early/\n after", "late/\n before", "late/\n after"),
  lhei=c(0.6, 4),
  col=my_palette,       # use on color palette defined earlier 
  breaks=col_breaks,    # enable color transition at specified limits
  RowSideColors=mycolor,
  Rowv=as.dendrogram(cluster2),
  dendrogram="row",     # only draw a row dendrogram
  Colv="NA",
  adjCol = c(1,0.5))           # turn off column clustering```

dev.off()
```


## heatmap of log2FC per sample time and cultivar
```{r heatmap of log2FC per sample time and cultivar}
mat_data <- as.matrix(values_norm_median_3fac_t_log2FC)
# change order of columns
mat_data <- mat_data[,c(1,5,9,13, 2,6,10,14, 3,7,11,15, 4,8,12,16)]
signif_sample_time_cultivar <- res_ttest_melt_cast[,c(8,6,7,9, 4,2,3,5, 16,14,15,17, 12,10,11,13)]

distance = dist(mat_data, method = "euclidian")
cluster = hclust(distance, method = "average")

#mycluster <- cutree(cluster, h=max(cluster$height)/1.5)
mycluster <- cutree(cluster, k=6)
#mycluster[cluster$order]

mycolor <- brewer.pal(8, "Set2")
mycolor <- mycolor[as.vector(mycluster)]

my_palette <- colorRampPalette(c("blue", "white", "red"))(n = 299)

col_breaks = c(seq(-5,-0.81,length=100), # for blue
               seq(-0.8,0.8,length=100), # for white
               seq(0.81,5,length=100)) # for red

#pdf("figures/merge_mpi_test_rewatering/heatmap_log2FC_signif_sample_time_cultivar_note.pdf", width = 12, height = 14)
dpi <- 300
png("figures/merge_mpi_test_rewatering/heatmap_log2FC_signif_sample_time_cultivar_note.png", 
    width=12*dpi, height=20*dpi, res = dpi)

lmat=rbind(4:3, 2:1)
lhei=c(0.8, 4)
lwid=c(1, 4)

layout(mat = lmat, widths = lwid, heights = lhei)

heatmap.2(mat_data,
  cellnote = signif_sample_time_cultivar,  # same data set for cell labels
  notecex=0.8, # numeric scaling factor for cellnote items
  colsep=0:ncol(mat_data), 
  rowsep=0:nrow(mat_data),
  sepwidth=c(0.01,0.05),
  sepcolor="black",
  main = "", # heat map title
  notecol="black",      # change font color of cell labels to black
  density.info="none",  # turns off density plot inside color legend
  trace="none",         # turns off trace lines inside the heat map
  margins =c(7,25),     # widens margins around plot
  keysize=1.5,
  cexRow=1,
  cexCol=1.3,
  labCol = names_sample_time_cultivar,
  lhei=c(0.6, 4),
  col=my_palette,       # use on color palette defined earlier 
  breaks=col_breaks,    # enable color transition at specified limits
  RowSideColors=mycolor,
  Rowv=as.dendrogram(cluster),
  dendrogram="row",     # only draw a row dendrogram
  Colv="NA",
  adjCol = c(1,0.5))            # turn off column clustering```

dev.off()
```




## export tables that contain only sig changed analytes (treatment effect)
```{r export tables that contain only sig changed analytes}
# only significantly changed analytes

# 144 metabolites that changed either per sample_time or across all samples in t-test (control vs stress)
write.table(values_norm_median_rev_t_log2FC[-not_sig,], 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_rev_t_log2FC_for_heatmap_sig.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)

# only 137 metabolites that changed per sample_time!
write.table(values_norm_median_rev_t_log2FC[union_sample_time,], 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_rev_t_log2FC_for_heatmap_sig_per_cul.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)

#########################

write.table(values_norm_median_rev_transformed[-not_sig,], 
            "output/merge_mpi_test_rewatering/values_norm_log10_median_rev_transformed_for_heatmap_sig.txt", 
            sep = "\t", quote = F, col.names = NA, row.names = T)
```


### export of tables for heatmap
```{r export of tables for heatmap}
# # export only rows with significant results of ANOVA for cultivar
# write.table(values_norm_median_rev_transformed[res_anova_treatment_cultivar_sig_cul,], 
#             "output/merge_mpi_test_rewatering/values_norm_log10_median_rev_transformed_for_heatmap_cul.txt", 
#             sep = "\t", quote = F, col.names = NA, row.names = T)
# 
# # export only rows with significant results of ANOVA for cultivar AND treatment
# write.table(values_norm_median_rev_transformed[res_anova_treatment_cultivar_sig_both,], 
#             "output/merge_mpi_test_rewatering/values_norm_log10_median_rev_transformed_for_heatmap_both.txt", 
#             sep = "\t", quote = F, col.names = NA, row.names = T)
```



### Save workspace and sessionInfo
```{r save workspace}
save.image("merge_mpi_test_rewatering.RData")
sessionInfo()
```

