---
title: "Merge GC-MS data from three VALDIS experiments from 2014"
author: "Heike Sprenger"
date: "November 21, 2015"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_float: true
    toc_depth: 4
---


# Set pander and knitr options
```{r set pander and knitr options, message=FALSE}
library(knitr)
library(pander)
library(colorRamps)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width = 5, fig.height = 5, cache = F, 
               highlight = T, fig.show = "asis")
opts_knit$set(root.dir = '../')

#setwd("D:/work/repos/trost_metabolomics")
setwd("X:/mpimp/repos/trost_metabolomics")
#load("merge_valdis_2014_B.RData")
```


# Load packages and functions for metabolite data analysis
```{r load packages and functions for metabolite data analysis, message=FALSE, fig.show='hide'}
# load packages
source("../functions/func_load_packages_metabolite_data_analysis.R")

# load functions
source("../functions/func_load_functions_metabolite_data_analysis.R")
```


# Load data
## Load analytes table
this table comes from ``gmd_analytes.Rmd``
```{r load analytes table}
# analytes_sel_exp_sort <- read.table("data/analytes_table_21exp_overlap_select_ordered.txt", 
#                                     sep = "\t", header = T, allowEscapes = T)

analytes_sel_exp_sort <- read.table("data/analytes_table_valdis_overlap_select_ordered.txt",
                                    sep = "\t", header = T, allowEscapes = T)
```

**The analytes table contains `r nrow(analytes_sel_exp_sort)` entries.**


## Load sample information
these tables come from ``gmd_data_analysis.Rmd``
```{r load sample information}
mpi_fgh_2014_factors <- read.table("output/mpi_fgh_2014_B/factors.txt", 
                                     sep = "\t", header = T, row.names = 1)

jki_shelter_2014_factors <- read.table("output/jki_shelter_2014_B/factors.txt", 
                                     sep = "\t", header = T, row.names = 1)
```


## Load data: log-transformed, with NAs, selected samples and selected overlapping analytes (`r nrow(analytes_sel_exp_sort)`)
```{r load data}
mpi_fgh_2014_values <- read.table("output/mpi_fgh_2014_B/gmd_values_select_log10.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)

jki_shelter_2014_values <- read.table("output/jki_shelter_2014_B/gmd_values_select_log10.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)

# without NAs
mpi_fgh_2014_values_woNA <- read.table("output/mpi_fgh_2014_B/gmd_values_select_log10_woNA.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)

jki_shelter_2014_values_woNA <- read.table("output/jki_shelter_2014_B/gmd_values_select_log10_woNA.txt", 
                                    sep = "\t", header = T, row.names = 1, check.names = F)
```


# Combine data
## Combine sample information tables by rows
```{r combine sample information tables}
factors <- rbind(mpi_fgh_2014_factors,  jki_shelter_2014_factors)
dim(factors) # 782 15

length(levels(factors$BatchID)) 
# 18
table(factors$BatchID)

levels(factors$treatment)
# "control" "drought stress"

levels(factors$sample_time)
# only late!

# reorder cultivar factor (use: check_names_ordered)
#levels(factors$genotype_name)
levels(factors$genotype_class)

# create vector for trial name
factors$trial <- as.factor(c(rep("mpi_fgh_2014", 398),
                             rep("jki_shelter_2014", 384)))

# create vector for tolerance
#factors$tolerance <- func_create_tolerance_factor(factors)

write.table(factors, "output/merge_valdis_2014_B/factors.txt", sep = "\t", row.names = F)
write.table(factors, "output/merge_valdis_2014_B/metabolite_samples_2014.txt",  sep = "\t", quote = F, col.names = NA, row.names = "chromatogram")
```


## Combine value tables by rows 
```{r combine value tables}
# only late timepoint, check cultivars
values <- rbind(mpi_fgh_2014_values, jki_shelter_2014_values)
dim(values)
# 782 85

# without NAs
values_woNA <- rbind(mpi_fgh_2014_values_woNA, jki_shelter_2014_values_woNA)
```

**The combined dataset contains `r nrow(values)` samples and `r ncol(values)` analytes.**

# Calculate percentage of NAs and plot histograms

```{r calculate percentage of NAs}

func_print_na_statistics(values, analytes_sel_exp_sort$Name)
# [1] "percentage of NAs:  10.376 %"
# [1] "num of analytes with more than 40% NAs:  7"
# [1] "num of NAs for problematic analytes:"
#              analyte_name number_of_NAs
# 1           Gluconic acid           582
# 2 Trehalose, alpha,alpha-           609
# 3           Phenylalanine           352
# 4            Malonic acid           550
# 5                  Uracil           703
# 6          Saccharic acid           504
# 7                 Proline           532

# analytes that have too many NAs and need to be removed for ANOVA normalization (by removeFactors function)
problematic_analytes <- c("Gluconic acid",
                          "Trehalose, alpha,alpha-",
                          #"Phenylalanine",   --> can inside data set
                          "Malonic acid",
                          "Uracil",
                          #"Saccharic acid",  --> can inside data set
                          "Proline")

problematic_analytes_id <- which(analytes_sel_exp_sort$Name %in% problematic_analytes)
analytes_sel_exp_sort_part <- analytes_sel_exp_sort[-problematic_analytes_id,]

pdf("figures/merge_valdis_2014_B/NAs_hist.pdf")
func_plot_na_statistics(values)
dev.off()

# without Gluconic acid, Saccharic acid and Proline
values_part <- values[,-problematic_analytes_id]
```


# Normalization: apply removeFactors function
```{r apply removeFactors function}
# values_norm <- func_normalize(values, factors,
#                               facs = c("genotype_class", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"),
#                               keep = c("genotype_class", "treatment"))
# ERROR!

# without 5 problematic metabolites
values_part_norm <- func_normalize(values_part, factors,
                              facs = c("genotype_class", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"),
                              keep = c("genotype_class", "treatment"))

# ALL data, but without considering treatment as factor
values_norm <- func_normalize(values, factors, 
                              facs = c("genotype_class", "SequenceID", "BatchID", "log10_AvgAnnotated"),
                              keep = c("genotype_class"))

# without NAs
values_woNA_norm <- func_normalize(values_woNA, factors,
                              facs = c("genotype_class", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"),
                              keep = c("genotype_class", "treatment"))

values_woNA_norm_2 <- func_normalize(values_woNA, factors,
                              facs = c("genotype_class", "SequenceID", "BatchID", "log10_AvgAnnotated"),
                              keep = c("genotype_class"))

write.table(values_norm, "output/merge_valdis_2014_B/values_norm.txt", sep="\t")
write.table(values_part_norm, "output/merge_valdis_2014_B/values_part_norm.txt", sep="\t")
```


# Modify table for qtl analysis
```{r modify table for qtl analysis}
values_export <- values_part_norm

colnames(values_export) <- analytes_sel_exp_sort_part$Name
rownames(values_export) <- factors$sample_id

write.table(values_export, "output/merge_valdis_2014_B/metabolite_data_2014_normalized.txt", sep="\t", quote = F, col.names = NA)
```


# Look for outliers after normalization of 3 merged data sets
```{r outlier detection}
pdf("figures/merge_valdis_2014_B/outlier_hist.pdf")
func_hist_outlier(values_norm, threshold=1e-6)
dev.off()

# keep possible outliers for now
```


#### check normal distribution --> shapiro test and histogram
```{r check normal distribution}
### plot histograms of normalized values per analyte
pdf("figures/merge_valdis_2014_B/values_norm_log10_hist.pdf")
func_plot_dist(values_norm)
dev.off()

# shapiro test for normal distribution per analyte
res_shapiro <- func_shapiro_test(values_norm, 1e-8)
```


# Perform PCA to check effect of normalization
## Prep for PCA
```{r prep for PCA}
# BEFORE normalization
pca_pareto_rnipals <- func_prep_pca(values, scale_method = "pareto", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")

pca_none_rnipals <- func_prep_pca(values, scale_method = "none", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")


# AFTER normalization
# pca_norm_pareto_rnipals <- func_prep_pca(values_norm, scale_method = "pareto", center_option = FALSE, 
#                                              pc_number = 5, pca_method = "rnipals")
# 
# pca_norm_none_rnipals <- func_prep_pca(values_norm, scale_method = "none", center_option = FALSE, 
#                                              pc_number = 5, pca_method = "rnipals")

pca_norm_pareto_rnipals <- func_prep_pca(values_part_norm, scale_method = "pareto", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")

pca_norm_none_rnipals <- func_prep_pca(values_part_norm, scale_method = "none", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")

# without NAs
pca_woNA_norm_pareto_rnipals <- func_prep_pca(values_woNA_norm, scale_method = "pareto", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")

pca_woNA_norm_2_pareto_rnipals <- func_prep_pca(values_woNA_norm_2, scale_method = "pareto", center_option = FALSE, 
                                             pc_number = 5, pca_method = "rnipals")
```


## Plot effect of missing value imputation by PCA 
```{r plot effect of missing value imputation by PCA}
# analyte 58: proline
sum(is.na(values_norm[,58]))
boxplot(values_norm[,58] ~ factors$treatment) # with NAs

boxplot(pca_norm_none_rnipals@completeObs[,58] ~ factors$treatment) # with NAs --> missing value imputation by PCA
boxplot(pca_norm_pareto_rnipals@completeObs[,58] ~ factors$treatment) # with NAs --> missing value imputation by PCA
# different scaling due to pareto scaling!
```


## PCA plots all trials
```{r PCA plot all trials}
pdf("figures/merge_valdis_2014_B/PCA_scoresplots_normalization_effect.pdf")
par(mar=c(4.5, 4.5, 2, 0.5))
# trial
palette( brewer.pal(2, "Dark2"))
func_5pairs_plot(pca_none_rnipals, factors, "trial", 19, "trial effect, before normalization")
func_5pairs_plot(pca_norm_none_rnipals, factors, "trial", 19, "trial effect, after normalization")
func_5pairs_plot(pca_norm_pareto_rnipals, factors, "trial", 19, "trial effect, after normalization (pareto scaling)")

# genotype
palette(cols_genotype_report)
func_5pairs_plot(pca_none_rnipals, factors, "genotype_class", 19, "genotype effect, before normalization")
func_5pairs_plot(pca_norm_none_rnipals, factors, "genotype_class", 19, "genotype effect, after normalization")
func_5pairs_plot(pca_norm_pareto_rnipals, factors, "genotype_class", 19, "genotype effect, after normalization (pareto scaling)")

# treatment
palette(cols_treatment_report)
func_5pairs_plot(pca_none_rnipals, factors, "treatment", 19, "treatment effect, before normalization")
func_5pairs_plot(pca_norm_none_rnipals, factors, "treatment", 19, "treatment effect, after normalization")
func_5pairs_plot(pca_norm_pareto_rnipals, factors, "treatment", 19, "treatment effect, after normalization (pareto scaling)")

# trial
palette( brewer.pal(2, "Dark2"))
func_pca_plot_sym(pca_none_rnipals, 1, 2, factors, "trial", c(19,17), 
              "treatment", 1.5, "topleft", 0.8, "topright", 0.8, "before normalization", 
              legend.text1 = c("JKI shelter 2014", "MPI FGH 2014"))

func_pca_plot_sym(pca_norm_none_rnipals, 1, 2, factors, "trial", c(19,17), 
              "treatment", 1.5, "bottomleft", 0.8, "topright", 0.8, "after normalization (no scaling)",
              legend.text1 = c("JKI shelter 2014", "MPI FGH 2014"))

func_pca_plot_sym(pca_norm_pareto_rnipals, 1, 2, factors, "trial", c(19,17), 
              "treatment", 1.5, "bottomleft", 0.8, "topright", 0.8, "after normalization (pareto scaling)",
              legend.text1 = c("JKI shelter 2014", "MPI FGH 2014"))

# without NAs
func_pca_plot_sym(pca_woNA_norm_pareto_rnipals, 1, 2, factors, "trial", c(19,17), 
              "treatment", 1.5, "bottomleft", 0.8, "topright", 0.8, "woNAs, after normalization (pareto scaling)",
              legend.text1 = c("JKI shelter 2014", "MPI FGH 2014"))

func_pca_plot_sym(pca_woNA_norm_2_pareto_rnipals, 1, 2, factors, "trial", c(19,17), 
              "treatment", 1.5, "bottomleft", 0.8, "topright", 0.8, "woNAs, after normalization (pareto scaling)",
              legend.text1 = c("JKI shelter 2014", "MPI FGH 2014"))


# cultivar
palette(cols_genotype_report)
func_pca_plot(pca_none_rnipals, 1, 2, factors, "genotype_class", 19, 1.5, "bottomleft", 1, maintext = "before normalization (no scaling)")
func_pca_plot(pca_norm_none_rnipals, 1, 2, factors, "genotype_class", 19, 1.5, "bottomleft", 1, maintext = "after normalization (no scaling)")
func_pca_plot(pca_norm_none_rnipals, 1, 3, factors, "genotype_class", 19, 1.5, "bottomright", 1, maintext = "after normalization (no scaling)")
func_pca_plot(pca_norm_pareto_rnipals, 1, 2, factors, "genotype_class", 19, 1.5, "bottomleft", 1, maintext = "after normalization (pareto scaling)")
func_pca_plot(pca_woNA_norm_pareto_rnipals, 1, 2, factors, "genotype_class", 19, 1.5, "bottomleft", 1, maintext = "woNAs, after normalization (pareto scaling)")
func_pca_plot(pca_woNA_norm_2_pareto_rnipals, 1, 2, factors, "genotype_class", 19, 1.5, "bottomleft", 1, maintext = "woNAs, after normalization (pareto scaling)")


# treatment
palette(cols_treatment_report)
func_pca_plot(pca_none_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1, maintext = "before normalization (no scaling)")
func_pca_plot(pca_norm_none_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1, maintext = "after normalization (no scaling)")
func_pca_plot(pca_norm_pareto_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1, maintext = "after normalization (pareto scaling)")
func_pca_plot(pca_woNA_norm_pareto_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1, maintext = "woNAs, after normalization (pareto scaling)")
func_pca_plot(pca_woNA_norm_2_pareto_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1, maintext = "woNAs, after normalization (pareto scaling)")

dev.off()
```


## PCA plots: genotype effect
```{r PCA plots: genotype effect}
palette(cols_genotype_report)

pdf("figures/merge_valdis_2014_B/PCA_scoresplots_genotype_effect.pdf",6, 6)
par(mar=c(4.5, 4.5, 2, 0.5))

func_5pairs_plot(pca_norm_pareto_rnipals, factors, "genotype_class", 19, "genotype effect")
func_5pairs_plot_sym(pca_norm_pareto_rnipals, factors, "genotype_class", c(19,17), "treatment", "genotype / treatment effect")

# PC1 vs PC2
func_pca_plot_sym(pca_norm_pareto_rnipals, 1, 2, factors, "genotype_class", c(19,17), 
              "treatment", 1.5, "bottomright", 0.8, "topright", 0.8, "genotype / treatment effect")
# PC1 vs PC3
func_pca_plot_sym(pca_norm_pareto_rnipals, 1, 3, factors, "genotype_class", c(19,17), 
              "treatment", 1.5, "bottomright", 0.8, "topright", 0.8, "genotype / treatment effect")
# PC2 vs PC3
func_pca_plot_sym(pca_norm_pareto_rnipals, 2, 3, factors, "genotype_class", c(19,17), 
              "treatment", 1.5, "bottomleft", 0.8, "topleft", 0.8, "genotype / treatment effect")

func_pca_plot(pca_norm_pareto_rnipals, 2, 3, factors, "genotype_class", 19, 1.5, "bottomleft", 1)
func_pca_plot(pca_norm_pareto_rnipals, 2, 4, factors, "genotype_class", 19, 1.5, "topleft", 1, ymax = 3)
func_pca_plot(pca_norm_pareto_rnipals, 3, 4, factors, "genotype_class", 19, 1.5, "topleft", 1, ymax = 3)
func_pca_plot(pca_norm_pareto_rnipals, 3, 5, factors, "genotype_class", 19, 1.5, "topleft", 1)

# NO SCALING --> shows separation of lines by PC2 (e.g. A179012)
func_pca_plot(pca_norm_none_rnipals, 1, 2, factors, "genotype_class", 19, 1.5, "bottomleft", 1,
              maintext = "genotype effect (no scaling)")
func_pca_plot_sym(pca_norm_none_rnipals, 1, 2, factors, "genotype_class", c(19,17), 
              "treatment", 1.5, "bottomright", 0.8, "topright", 0.8, 
              maintext = "genotype / treatment effect (no scaling)", xmax = 4.5)
dev.off()
```


## PCA plots: treatment effect
```{r PCA plots: treatment effect}
palette(cols_treatment_report)

pdf("figures/merge_valdis_2014_B/PCA_scoresplots_treatment_effect.pdf",6, 6)
par(mar=c(4.5, 4.5, 2, 0.5))

func_5pairs_plot(pca_norm_pareto_rnipals, factors, "treatment", 19, "cultivar effect")
func_5pairs_plot_sym(pca_norm_pareto_rnipals, factors, "treatment", c(19,17,15), "trial", "treatment / trial effect")

# PC1 vs PC2
func_pca_plot_sym(pca_norm_none_rnipals, 1, 2, factors, "treatment", c(19,17,15), 
              "trial", 1.5, "bottomright", 0.8, "topright", 0.8, "treatment / trial effect")

func_pca_plot_sym(pca_norm_pareto_rnipals, 1, 2, factors, "treatment", c(19,17,15), 
              "trial", 1.5, "bottomright", 0.8, "bottomleft", 0.8, "treatment / trial effect (pareto)")

func_pca_plot(pca_norm_none_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1)

func_pca_plot(pca_norm_pareto_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1)

dev.off()
```


## PCA plots: report
```{r PCA plots: report}
#pdf("figures/merge_valdis_2014_B/PCA_scoresplots_report.pdf", 12, 6)
png("figures/merge_valdis_2014_B/PCA_scoresplots_report.png", 4000, 2000, res = 300)

par(mar=c(4.5, 4.5, 0.5, 0.5))
par(mfrow=c(1,2))

palette(cols_treatment_report)
func_pca_plot(pca_norm_pareto_rnipals, 1, 2, factors, "treatment", 19, 1.5, "bottomleft", 1, legend.text = c("Kontrolle", "Trockenstress"))

palette(cols_genotype_report)
func_pca_plot(pca_norm_pareto_rnipals, 2, 4, factors, "genotype_class", 19, 1.5, "topleft", 1, ymax = 3)

dev.off()
```


# Boxplot of A179012
```{r boxplot of A179012}

which(analytes_sel_exp_sort$analyteID == "A179012-101")

pdf("figures/merge_valdis_2014_B/boxplot_A179012-101.pdf", 9, 6)

boxplot(values_norm[,33] ~ factors$genotype_class, 
        las = 2, col = cols_genotype_report, 
        ylab = "log10-normalisierter Metabolit-Gehalt")

boxplot(values_norm[,33] ~ factors$genotype_name, las = 2)

# sorted boxplot
matrix_df <- data.frame(value = values_norm[, 33], 
                        genotype = factors$genotype_name)
    
# sort genotypes according to median for boxplot
matrix_df_sorted <- with(matrix_df, reorder(genotype, value, median, na.rm = T))

# sorted boxplot
boxplot(value ~ matrix_df_sorted, data = matrix_df, 
        ylab = "log10-normalisierter Metabolit-Gehalt", 
        las = 2, cex.lab = 1, cex.axis = 0.7, 
        main = analytes_sel_exp_sort$Name[33],
        col = c(rep(cols_treatment[1], 32), 
                rep(cols_treatment[2],28)) )

dev.off()

#############

#pdf("figures/merge_valdis_2014_B/boxplot_A179012-101_report.pdf", 12, 4)
png("figures/merge_valdis_2014_B/boxplot_A179012-101_report.png", 5000, 1200, res=300)

par(mar=c(4.5, 4.5, 0.5, 0.5))

layout(matrix(c(1,2), 1, 2, byrow = TRUE),   
       widths = c(1,2), heights = 1)

boxplot(values_norm[,33] ~ factors$genotype_class, 
        las = 2, col = cols_genotype_report, 
        ylab = "log10-normalisierter Metabolit-Gehalt")

# sorted boxplot
boxplot(value ~ matrix_df_sorted, data = matrix_df, 
        ylab = "log10-normalisierter Metabolit-Gehalt", 
        las = 2, cex.lab = 1, cex.axis = 0.7, 
        col = c(rep(cols_treatment[1], 32), 
                rep(cols_treatment[2],28)) )

dev.off()
```


# ANOVA, boxplots & t-test
## ANOVA for 2 factors: treatment * cultivar (with interaction)
```{r ANOVA with 2 factors (with interaction)}
res_anova_treatment_genotype <- func_anova_2fac_ia(values_norm, factors, "treatment", "genotype_class", 0.01, analytes_sel_exp_sort$Name)

res_anova_treatment_cultivar_sig_treatment <- which(res_anova_treatment_genotype[,1] < 0.01)
res_anova_treatment_cultivar_sig_genotype <- which(res_anova_treatment_genotype[,2] < 0.01)
res_anova_treatment_cultivar_sig_both <- intersect (which(res_anova_treatment_genotype[,1] < 0.01), 
                                                    which(res_anova_treatment_genotype[,2] < 0.01))

write.table(res_anova_treatment_genotype, "output/merge_valdis_2014_B/res_anova_treatment_genotype.txt", sep="\t")

# try out genotype_name (for QTL analysis)
res_anova_genotype_name <- func_anova_1fac(values_norm, factors, "genotype_name", 0.01, analytes_sel_exp_sort$Name)
names(res_anova_genotype_name) <- analytes_sel_exp_sort$Name
head(sort(res_anova_genotype_name))

write.table(data.frame(res_anova_genotype_name, row.names = make.names(names(res_anova_genotype_name), unique = T)), 
            "output/merge_valdis_2014_B/res_anova_genotype.txt", col.names = F, row.names = T)
```


## Boxplots of normalized values per analyte using 2 factors
```{r boxplots of normalized values per analyte using 2 factors}
pdf("figures/merge_valdis_2014_B/boxplot_treatment_genotype.pdf", width=7, height=8)
par(mar=c(7, 4.1, 7, 2.1))
func_boxplot_2fac(normalized_values = values_norm, 
                  trial_factors = factors, 
                  factor1 = "treatment", 
                  factor2 = "genotype_class", 
                  res_anova_adj = res_anova_treatment_genotype, 
                  cols = cols_treatment_report,
                  names_factors = levels(interaction(factors$treatment, factors$genotype_class)),
                  analytes_sel_exp_sort$Name)

dev.off()

################################

pdf("figures/merge_valdis_2014_B/boxplot_treatment.pdf")
func_boxplot_1fac(values_norm, 
                  factors, 
                  "treatment", 
                  res_anova_treatment_genotype, 
                  cols_treatment, 
                  analytes_sel_exp_sort$Name)
dev.off()
```



## Sorted boxplots per genotype
```{r sorted boxplots per genotype}
pdf("figures/merge_valdis_2014_B/boxplot_sorted_per_genotype.pdf", 10, 6)
func_boxplot_sorted2(values_norm, factors$genotype_name, analytes_sel_exp_sort)
dev.off()


pdf("figures/merge_valdis_2014_B/boxplot_genotype_class.pdf", 5, 5)
func_boxplot_1fac(values_norm, 
                  factors, 
                  "genotype_class", 
                  res_anova_treatment_genotype, 
                  cols_genotype_report, 
                  analytes_sel_exp_sort$Name)
dev.off()


pdf("figures/merge_valdis_2014_B/boxplot_sorted_per_genotype_class.pdf", 10, 6)
func_boxplot_sorted2(values_norm, factors$genotype_class, analytes_sel_exp_sort)
dev.off()
```



### Save workspace and sessionInfo
```{r save workspace}
save.image("merge_valdis_2014_B.RData")
sessionInfo()
```

