---
title: "Combine metabolite data from TROST FIELD trials and Breeder trials"
author: "Heike Sprenger"
date: "May 4, 2016"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_float: true
    toc_depth: 4
---

# Set working directory  
```{r set working directory}
getwd()
#setwd("X:/mpimp/repos/trost_metabolomics/")
```


# Load workspace, packages and scripts
```{r load workspace, message=FALSE}
# load packages
library(knitr)
library(reshape)
library(pander)
library(pcaMethods)
library(ggplot2)
library(corrplot)
library(gplots)
library(RColorBrewer)
library(FactoMineR)
library(psych)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# load workspace
#load("metabolite_data_field_and_breeder.RData")
```


# Source R functions
```{r source R functions}
# source("../functions/colors.R")
# source("../functions/names.R")
# source("../functions/func_histograms.R")
# source("../functions/func_prep_pca.R")
# source("../functions/func_pca_plots.R")
# source("../functions/RemoveFactors_function.R")
```


# Load packages and functions for metabolite data analysis
```{r load packages and functions for metabolite data analysis, message=FALSE, fig.show='hide'}
# load packages
source("../functions/func_load_packages_metabolite_data_analysis.R")

# load functions
source("../functions/func_load_functions_metabolite_data_analysis.R")
```


# Load phenotyper results and tolerance
this table comes from ``phenotyper_metadata.Rmd``
```{r load phenotyper results and tolerance}
phenotyper_result_joined <- read.table("data/phenotyper_result_joined_with_gmd_ids.txt", 
                                       sep="\t", header = T)

# SHOULD BE ZERO!
sum(is.na(phenotyper_result_joined$genotype_name))
# IF NOT EXECUTE ~\work\repos\database_scripts\update_plantlines_alias.sql
# AND REPEAT gmd_complete_workflow.Rmd

# tolerance
tolerance_table <- read.table("../trost_phenotypes/output/tolerance_phenotyper_metadata_2sub.txt", 
                        sep = "\t", header = TRUE)

# change names of cultivars from upper to lower case
levels(tolerance_table$cultivar) <- names_cultivars_31

# relevel tolerance factors
tolerance_table$tol_cat2_fve <- factor(tolerance_table$tol_cat2_fve, levels = c("low", "high"))
tolerance_table$tol_cat3_fve <- factor(tolerance_table$tol_cat3_fve, levels = c("low","mid", "high"))
```


# Load analytes table

* overlapping analytes for 5 TROST field experiments and breeder trials
* this table comes from ``gmd_analytes.Rmd``

```{r load analytes table}

# 115 overlapping analytes
analytes_sel_exp_sort <- read.table("data/analytes_table_model_overlap_select_ordered.txt", 
                                    sep ="\t", header = T, allowEscapes = T)
```


# Load `trost_TagList`
`trost_TagList` shows all experiments that belong to the TROST/VALDIS project and comes from `gmd_first_steps.Rmd`

```{r load trost taglist}
trost_TagList <- read.table("data/trost_TagList.txt", sep="\t", header=TRUE)

trost_gmd_ids <- read.table("data/trost_gmd_ids.txt", sep="\t", header=TRUE)
head(trost_gmd_ids)
```


# Load tables with GMD values
```{r Load tables with GMD values}
gmd_values <- read.table("data/gmd_raw_values_matrix.txt", sep = "\t", header = T, check.names = F)
dim(gmd_values)
```


# Load tables with GMD metadata
```{r load tables with GMD metadata}
# TROST
gmd_meta_mpi_field_2011 <- read.table("data/gmd_meta_mpi_field_2011.txt", sep = "\t", header = T)
gmd_meta_mpi_field_2012 <- read.table("data/gmd_meta_mpi_field_2012.txt", sep = "\t", header = T)
gmd_meta_jki_field_2012 <- read.table("data/gmd_meta_jki_field_2012.txt", sep = "\t", header = T)
gmd_meta_jki_field_2013 <- read.table("data/gmd_meta_jki_field_2013.txt", sep = "\t", header = T)
gmd_meta_dethlingen_2011 <- read.table("data/gmd_meta_dethlingen_2011.txt", sep = "\t", header = T)
gmd_meta_breeder_trials <- read.table("data/gmd_meta_breeder_trials.txt", sep = "\t", header = T)
```


# Bind GMD metadata

* After update of Phenotyper DB it might be necessary to repeat this script: ``~\work\repos\database_scripts\update_plantlines_alias.sql`` because genotype_name of some cultivars are NA (78)

```{r bind GMD metadata}
gmd_meta <- rbind(gmd_meta_mpi_field_2011, gmd_meta_mpi_field_2012, 
                  gmd_meta_jki_field_2012, gmd_meta_jki_field_2013,
                  gmd_meta_dethlingen_2011, gmd_meta_breeder_trials)
dim(gmd_meta)

# change treatment: 50% nFK to Kontrolle
levels(gmd_meta$treatment)[4] <- levels(gmd_meta$treatment)[1] 

# change genotype names for 31 cultivars
levels(gmd_meta$genotype_name) <- names_cultivars_34

# use only control and stress treatments
gmd_meta_2 <- droplevels(subset(gmd_meta, gmd_meta$treatment %in% c("Kontrolle", "Trockenstress")))
levels(gmd_meta_2$treatment) <- c("control", "drought stress")
dim(gmd_meta_2)

# use only late samples
gmd_meta_3 <- droplevels(subset(gmd_meta_2, gmd_meta_2$sample_time == "late"))
dim(gmd_meta_3)

# use only relevant genotypes (31 cultivars and 193 lines from 2014)
genotypes_model <- union(names_cultivars_31, names_lines_2014)
gmd_meta_model <- droplevels(subset(gmd_meta_3, gmd_meta_3$genotype_name %in% genotypes_model))
dim(gmd_meta_model)
# 1403 29

table(gmd_meta_model$experiment_name)
table(gmd_meta_model$experiment_id)
```


# Trial names
```{r trial names}

trial_1 <- as.character(gmd_meta_model$experiment_name)

trial_1[which(gmd_meta_model$experiment_id==44443)] = "mpi_feld_2011"
trial_1[which(gmd_meta_model$experiment_id==56726)] = "mpi_feld_2012"
trial_1[which(gmd_meta_model$experiment_id==56875)] = "jki_feld_2012"
trial_1[which(gmd_meta_model$experiment_id==62327)] = "jki_feld_2013"
trial_1[which(gmd_meta_model$experiment_id==46150)] = "dethlingen_2011"

trial_1 <- as.factor(trial_1)
table(trial_1)

############

trial_2 <- rep("breeder_2011", nrow(gmd_meta_model))

trial_2[which(gmd_meta_model$experiment_id %in% 
                c(56876, 56878, 56879, 56880, 56881, 56882,  56883, 56884))] = "breeder_2012"

trial_2[which(gmd_meta_model$experiment_id==44443)] = "mpi_feld_2011"
trial_2[which(gmd_meta_model$experiment_id==56726)] = "mpi_feld_2012"
trial_2[which(gmd_meta_model$experiment_id==56875)] = "jki_feld_2012"
trial_2[which(gmd_meta_model$experiment_id==62327)] = "jki_feld_2013"
trial_2[which(gmd_meta_model$experiment_id==46150)] = "dethlingen_2011"

trial_2 <- as.factor(trial_2)
table(trial_2)

###########

gmd_meta_model <- data.frame("trial" = trial_1, "trial_2" = trial_2, gmd_meta_model)

# add combined column for treatment and breeder year
gmd_meta_model$treatment2 <- as.character(gmd_meta_model$treatment)
gmd_meta_model$treatment2[which(gmd_meta_model$trial_2 == "breeder_2011")] <- "2011"
gmd_meta_model$treatment2[which(gmd_meta_model$trial_2 == "breeder_2012")] <- "2012"
gmd_meta_model$treatment2 <- factor(gmd_meta_model$treatment2, 
                                    levels = c("control", "drought stress", "2011","2012"))
table(gmd_meta_model$treatment2)
```


# Subset of metadata for field, TROST or VALDIS
```{r subset of metadata for field, trost or valdis}
# sort table by chromatogram ID
gmd_meta_model$chromatogram <- as.character(gmd_meta_model$chromatogram)
gmd_meta_model_sort <- gmd_meta_model[order(gmd_meta_model$chromatogram),]
```


# Export final metadata tables
```{r export final metadata tables}
# write.table(gmd_meta_model, "output/gmd_meta_model.txt", sep="\t")
write.table(gmd_meta_model_sort, "output/field_breeder/metabolite_meta.txt", sep="\t")
```


# Define subset of selected chromatograms 
```{r subset of chromatograms}
# select real biological samples
gmd_values_model <- subset(gmd_values, rownames(gmd_values) %in% gmd_meta_model$chromatogram)
dim(gmd_values_model)
# 1403 566
```


# Define subset of overlapping analytes
* SELECT: `r nrow(analytes_sel_exp_sort)` overlapping analytes regarding all TROST field experiments, WITHOUT internal standards and contaminations
```{r subset of analytes}
gmd_values_select <- func_get_overlapping_analytes(analytes_sel_exp_sort, gmd_values_model)
dim(gmd_values_select)
# 1403 115

write.table(gmd_values_select, "output/field_breeder/metabolite_data.txt", sep="\t")
```


# Calculate percentage of NAs and plot histograms

* print percentage of NAs, number of analytes with more than 40% NAs and ID of those analytes

```{r calculate percentage of NAs and plot histograms}
func_print_na_statistics(gmd_values_select, analytes_sel_exp_sort$Name)

pdf("figures/field_breeder/NAs_hist_all_samples.pdf")
func_plot_na_statistics(gmd_values_select)
dev.off()
```


# Replace NAs with minimal values of dataset and calculate LOG10 of raw intensity values 
```{r replace NAs and calc LOG10 of raw values}
# log10 transformation (with NAs)
gmd_values_select_log10 <- log10(gmd_values_select)

# replace NAs with lowest value of dataset
gmd_values_select_woNA <- func_replace_na_with_min_value(gmd_values_select)

# log10 transformation after NA replacement -> without NAs 
gmd_values_select_log10_woNA <- log10(gmd_values_select_woNA)

sum(is.na(gmd_values_select_log10))
sum(is.na(gmd_values_select_log10_woNA))

# all
write.table(gmd_values_select_log10, "output/field_breeder/metabolite_data_log10.txt", sep="\t")
write.table(gmd_values_select_log10_woNA, "output/field_breeder/metabolite_data_log10_woNA.txt", sep="\t")
```


# Extract Batch and SequenceID
```{r Extract Batch and SequenceID}
gmd_meta_model_sort$BatchID <- func_get_batch_ids(gmd_meta_model_sort)
gmd_meta_model_sort$SequenceID <- func_get_sequence_ids(gmd_meta_model_sort)
gmd_meta_model_sort$log10_AvgAnnotated <- log10(gmd_meta_model_sort$AvgAnnotated)
```


# Relevel factors and create factors table
```{r relevel factors and create factors table}
# define tolerance factor
# tolerance <- func_create_tolerance_factor(gmd_meta_model_sort)

# create factors table for treatment, cultivar, sample_time, Dw, Fw, Is, AvgAnnotated, AvgAnnotatedLog10, BatchID, SequenceID
factors <- func_create_factors_table_field(gmd_meta_model_sort, 
                                     gmd_meta_model_sort$SequenceID, 
                                     gmd_meta_model_sort$BatchID)

colnames(factors)

dim(factors)
dim(gmd_meta_model_sort)

write.table(factors, "output/field_breeder/factors.txt", sep="\t")
```


# ANOVA normalization
```{r ANOVA normalization}
dim(gmd_values_select_log10_woNA)
dim(gmd_values_select_log10)

# without NAs (replaced by min)
norm_values_log10_woNA <- apply(gmd_values_select_log10_woNA, 2, RemoveFactors, sam = factors, 
                              facs = c("genotype_name", "SequenceID", "BatchID", "log10_AvgAnnotated"), 
                              keep = c("genotype_name"))

norm_values_log10 <- apply(gmd_values_select_log10, 2, RemoveFactors, sam = factors, 
                              facs = c("genotype_name", "SequenceID", "BatchID", "log10_AvgAnnotated"), 
                              keep = "genotype_name")

sum(is.na(norm_values_log10)) / (ncol(norm_values_log10) * nrow(norm_values_log10))*100
# 5.4%

write.table(norm_values_log10_woNA, "output/field_breeder/metabolite_data_log10_woNA_norm.txt", sep="\t")
write.table(norm_values_log10, "output/field_breeder/metabolite_data_log10_norm.txt", sep="\t")
```


# PCA
## PCA TROST field and breeder trials
```{r PCA TROST field and breeder trials}

gmd_values_log10_prep <- prep(gmd_values_select_log10, scale = "none", center = F) # before normalization
norm_values_log10_prep <- prep(norm_values_log10, scale = "none", center = FALSE) # after normalization (with NA)
# with scaling
norm_values_log10_prep_pareto <- prep(norm_values_log10, scale = "pareto", center = FALSE) 
norm_values_log10_prep_vector <- prep(norm_values_log10, scale = "vector", center = FALSE) 

gmd_values_log10_prep_rnipals <- pca(gmd_values_log10_prep, nPcs = 5, method="rnipals")
norm_values_log10_prep_rnipals <- pca(norm_values_log10_prep, nPcs = 5, method="rnipals")
# with scaling
norm_values_log10_prep_pareto_rnipals <- pcaMethods::pca(norm_values_log10_prep_pareto, nPcs = 5, method="rnipals")
norm_values_log10_prep_vector_rnipals <- pcaMethods::pca(norm_values_log10_prep_vector, nPcs = 5, method="rnipals")

norm_values_log10_prep_rnipals@R2
norm_values_log10_prep_pareto_rnipals@R2
norm_values_log10_prep_vector_rnipals@R2

# complete observations of TROST data
compObs <- norm_values_log10_prep_rnipals@completeObs
dim(compObs)

# export normalized values with complete observations (no NAs) for model
write.table(compObs, "output/field_breeder/metabolite_data_log10_norm_completeObs.txt", sep="\t")
```


## PCA plots
```{r PCA plots}
pdf("figures/field_breeder/PCA_plot.pdf")
# trial effect
# before normalization
palette(heike_palette_7)
pairs(gmd_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$trial_2)
plot(gmd_values_log10_prep_rnipals@scores[,1],
     gmd_values_log10_prep_rnipals@scores[,2], 
     col = gmd_meta_model_sort$trial_2, pch = 19,
     main = "PCA effect of trial before normalization")
legend("topright", legend = levels(gmd_meta_model_sort$trial_2), 
       fill = 1:7, cex = 0.8)

# after normalization
pairs(norm_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$trial_2)
plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,2], 
     col = gmd_meta_model_sort$trial_2, pch = 19,
     main = "PCA effect of trial after normalization")
legend("bottomright", legend = levels(gmd_meta_model_sort$trial_2), 
       fill = 1:7, cex = 0.8)

# treatment effect
palette(heike_palette_4)
pairs(gmd_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$treatment2,
      main = "before normalization")

pairs(norm_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$treatment2,
      main = "after normalization")

plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,2], 
     col = gmd_meta_model_sort$treatment2, pch = 19,
     xlab = "PC1", ylab = "PC2",
     main = "PCA effect of treatment after normalization")
legend("bottomright", legend = levels(gmd_meta_model_sort$treatment2), 
       fill = 1:4, cex = 0.8)

plot(norm_values_log10_prep_rnipals@scores[,2],
     norm_values_log10_prep_rnipals@scores[,4], 
     col = gmd_meta_model_sort$treatment2, pch = 19,
     xlab = "PC2", ylab = "PC4",
     main = "PCA effect of treatment after normalization")
legend("bottomright", legend = levels(gmd_meta_model_sort$treatment2), 
       fill = 1:4, cex = 0.8)

# genotype effect
palette(rainbow(34))
pairs(norm_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$genotype_name)

# PC1 vs PC2
plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,2], 
     col = gmd_meta_model_sort$genotype_name, pch = 19,
     main = "PCA effect of genotype after normalization")
legend("bottomright", legend = levels(gmd_meta_model_sort$genotype_name), 
       fill = 1:34, cex = 0.6)

# PC1 vs PC3
plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,3], 
     col = gmd_meta_model_sort$genotype_name, pch = 19,
     main = "PCA effect of genotype after normalization")
legend("bottomright", legend = levels(gmd_meta_model_sort$genotype_name), 
       fill = 1:34, cex = 0.6)

plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,3], 
     col = gmd_meta_model_sort$genotype_name, pch = 19,
     main = "PCA effect of genotype after normalization")
text(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,3], 
     labels = gmd_meta_model_sort$genotype_name, cex=0.5)
legend("bottomright", legend = levels(gmd_meta_model_sort$genotype_name), 
       fill = 1:34, cex = 0.6)
dev.off()
```


## PCA plot for publication
```{r PCA plot for publication}
norm_values_log10_prep_rnipals@R2

pdf("figures/field_breeder/metabolites_PCA_plot_for_publication_normalization.pdf", width = 14, height = 7)
par(mar=c(4.3, 4.5, 2, 1), mfrow=c(1,2))
palette(c("#023fa5", "#8e063b", "#11c638", "#ef9708", "#00BFFF", "#BEBEBE", "#000000")) # darkblue red green orange deepskyblue gray black

func_pca_plot(gmd_values_log10_prep_rnipals, 1, 2, gmd_meta_model_sort, "trial_2", 19, 1, 
              "topright", 1, maintext = "before correction",
              legend.text = c("A1-A8", "A9-A16", "F2", "F4", "F7", "F1", "F3"))

func_pca_plot(norm_values_log10_prep_rnipals, 1, 2, gmd_meta_model_sort, "trial_2", 19, 1, 
              "topleft", 1, xmin = -3, xmax = 3, ymax = 3.5,
              maintext = "after correction",
              legend.text = c("A1-A8", "A9-A16", "F2", "F4", "F7", "F1", "F3"))
dev.off()

pdf("figures/field_breeder/metabolites_PCA_plot_for_publication_treatment.pdf", width = 7, height = 7)
par(mar=c(4.3, 4.5, 2, 1))

# source: http://www.ucl.ac.uk/~zctpep9/Archived%20webpages/Cookbook%20for%20R%20%C2%BB%20Colors%20(ggplot2).htm
# c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# orange,       lightblue,  green,      yellow,     blue,     red,        pink

#palette(c("darkblue", "deeppink4", "forestgreen", "darkorange3"))
palette(c("#023fa5", "#8e063b", "#11c638", "#ef9708")) # darkblue red green orange

func_pca_plot(norm_values_log10_prep_rnipals, 1, 2, gmd_meta_model_sort, "treatment2", 19, 1, 
              "topleft", 1.4, xmin = -3, xmax = 3, ymax = 4,
              legend.text = c("experimental field trials: control", "experimental field trials: drought stress", 
                              "agronomic field trials 2011", "agronomic field trials 2012"))
dev.off()


pdf("figures/field_breeder/metabolites_PCA_plot_for_publication_treatment_PC2-4.pdf", width = 7, height = 7)
par(mar=c(4.3, 4.5, 2, 1))
palette(brewer.pal(4, "Set1"))

func_pca_plot(norm_values_log10_prep_rnipals, 2, 4, gmd_meta_model_sort, "treatment2", 19, 1, 
              "topleft", 1, 
              legend.text = c("experimental field trials: control", "experimental field trials: drought stress", 
                              "agronomic field trials 2011", "agronomic field trials 2012"))
dev.off()


pdf("figures/field_breeder/metabolites_PCA_plot_for_publication_genotype.pdf", width = 7, height = 7)
par(mar=c(4.3, 4.5, 2, 1))
palette(cols_distinct_31)

# PC1 vs PC3
func_pca_plot(norm_values_log10_prep_rnipals, 1, 3, gmd_meta_model_sort, "genotype_name", 19, 1, 
              "topleft", 0.8, xmin = -3.5, xmax = 3,
              legend.text = levels(gmd_meta_model_sort$genotype_name))

# PC1 vs PC2
func_pca_plot(norm_values_log10_prep_rnipals, 1, 2, gmd_meta_model_sort, "genotype_name", 19, 1, 
              "topleft", 0.8, xmin = -3.5, xmax = 3,
              legend.text = levels(gmd_meta_model_sort$genotype_name))

# PC2 vs PC3
func_pca_plot(norm_values_log10_prep_rnipals, 2, 3, gmd_meta_model_sort, "genotype_name", 19, 1, 
              "topleft", 0.8, xmin = -3.5, xmax = 3,
              legend.text = levels(gmd_meta_model_sort$genotype_name))
dev.off()
```


### Plots for scaled PCA
```{r Plots for scaled PCA}
################## SCALED
pdf("figures/field_breeder/metabolites_PCA_plot_for_publication_treatment_scaled.pdf", width = 7, height = 7)
par(mar=c(4.3, 4.5, 2, 1))
palette(c("#023fa5", "#8e063b", "#11c638", "#ef9708")) # darkblue red green orange

func_pca_plot(norm_values_log10_prep_pareto_rnipals, 1, 2, gmd_meta_model_sort, "treatment2", 19, 1, 
              "topleft", 1, xmin = -6.6, xmax = 5.5, ymax = 3.8, ymin = -5, maintext = "pareto scaling",
              legend.text = c("experimental field trials: control", "experimental field trials: drought stress", 
                              "agronomic field trials 2011", "agronomic field trials 2012"))

func_pca_plot(norm_values_log10_prep_vector_rnipals, 1, 2, gmd_meta_model_sort, "treatment2", 19, 1, 
              "topleft", 1, xmin = -0.03, xmax = 0.03, ymax = 0.03, ymin = -0.03, maintext = "vector scaling",
              legend.text = c("experimental field trials: control", "experimental field trials: drought stress", 
                              "agronomic field trials 2011", "agronomic field trials 2012"))
dev.off()
############################

pdf("figures/field_breeder/metabolites_PCA_plot_for_publication_genotype_scaled.pdf", width = 7, height = 7)
par(mar=c(4.3, 4.5, 2, 1))
palette(cols_distinct_31)

# PC1 vs PC3
func_pca_plot(norm_values_log10_prep_pareto_rnipals, 1, 3, gmd_meta_model_sort, "genotype_name", 19, 1, 
              "topleft", 0.8, xmin = -6.6, xmax = 5.5, ymax = 5.2, ymin = -4,
              legend.text = levels(gmd_meta_model_sort$genotype_name))

# PC1 vs PC2
func_pca_plot(norm_values_log10_prep_pareto_rnipals, 1, 2, gmd_meta_model_sort, "genotype_name", 19, 1, 
              "topleft", 0.8, xmin = -6.6, xmax = 5.5, ymax = 3.8, ymin = -5,
              legend.text = levels(gmd_meta_model_sort$genotype_name))
dev.off()
```


# Correlation matrices
```{r correlation matrices}
dim(norm_values_log10_woNA)
norm_values_log10_df <- data.frame(norm_values_log10_woNA)
colnames(norm_values_log10_df) <- analytes_sel_exp_sort$Name

dim(gmd_meta_model_sort)
trials <- levels(gmd_meta_model_sort$trial)
trials_id <- c("F2", "A10", "A11", "A12", "A13", "A14", "A15", "A16", "A9", "A1", "A2", "A3", "A4", "A5", "A6", "A7", "A8", "F4", "F7", "F1", "F3")
years_for_heatmap <- c("2011", rep("2012", 8), rep("2011", 8), "2012", "2013", "2011", "2012")

cols_set3 <- brewer.pal(11, "Set3")
cols_set1 <- brewer.pal(9, "Set1")

col_years_for_heatmap <- c(cols_set1[2], rep(cols_set1[3], 8), rep(cols_set1[2], 8), cols_set1[3], cols_set1[1], cols_set1[2], cols_set1[3])

col_loc_for_heatmap <- c(cols_set3[1], cols_set3[2], cols_set3[3], cols_set3[4], cols_set3[5], 
                        cols_set3[6], cols_set3[7], cols_set3[8], cols_set3[9], cols_set3[2],
                        cols_set3[3], cols_set3[4], cols_set3[5], cols_set3[6], cols_set3[7], 
                        cols_set3[8], cols_set3[9], cols_set3[10], cols_set3[10], cols_set3[11], cols_set3[11])

# --- 1. Step: Create subsets for each experiment --- #

norm_values_log10_per_exp <- split(norm_values_log10_df, gmd_meta_model_sort$trial)
class(norm_values_log10_per_exp$`Feldanbau 2012 Buetow`)
names(norm_values_log10_per_exp)

# --- 2. Step: Create correlation matrizes - spearman (for non-normal distributed data)
# For 

# corr_list <- lapply(norm_values_log10_per_exp, function(x) corr.test(x, use = "pairwise", 
#                                                                      method="spearman", adjust="BH", 
#                                                                      alpha=.05, ci = TRUE))

corr_list_2 <- lapply(norm_values_log10_per_exp, function(x) cor(x, use = "complete", method = "spearman"))

dim(corr_list_2$dethlingen_2011)

#pdf("figures/test.pdf", 10, 10)
corrplot(corr_list_2$dethlingen_2011, order = "hclust", method = "color", tl.col = "black", tl.cex = 0.7, outline = TRUE, mar = c(1, 0.5, 2, 0.5))
#dev.off()
```


# coeffRV
```{r coeffRV}
coeffRV(corr_list_2$mpi_feld_2011, corr_list_2$mpi_feld_2012)$rv
coeffRV(corr_list_2$jki_feld_2012, corr_list_2$jki_feld_2013)$rv
coeffRV(corr_list_2$mpi_feld_2012, corr_list_2$jki_feld_2013)$rv

coeffRV_res <- list()

for (i in trials){
  for (j in trials){
    coeffRV_res[[i]][j] <- coeffRV(corr_list_2[[j]], corr_list_2[[i]])$rv
  }
}

#coeffRV_res

head(unlist(coeffRV_res))
length(unlist(coeffRV_res))

coeffRV_res_mat <- matrix(unlist(coeffRV_res), ncol = 21)
coeffRV_res_mat[1:3, 1:3]
colnames(coeffRV_res_mat) <- rownames(coeffRV_res_mat) <- trials
colnames(coeffRV_res_mat) <- rownames(coeffRV_res_mat) <- trials_id


col_spec <- colorRampPalette(brewer.pal(11,"Spectral"))(100)
col_grey <- colorRampPalette(brewer.pal(9,"Greys"))(100)

corrplot(coeffRV_res_mat, order = "hclust", method = "color", tl.col = "black", tl.cex = 0.7, 
         outline = TRUE, col=col_spec, is.corr = FALSE)

pdf("figures/field_breeder/heatmap_coeffRV.pdf")
heatmap.2(coeffRV_res_mat, col = col_grey, trace = "none", density.info = "none",
          RowSideColors = col_years_for_heatmap,
          ColSideColors = col_loc_for_heatmap)
dev.off()
```



# Save workspace and sessioninfo
```{r save workspace}
save.image("metabolite_data_field_and_breeder.RData")
sessionInfo()
```

