---
title: Prediction model with TROST and VALDIS data
author: "Heike Sprenger"
date: "March 6, 2016"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---

# Set pander and knitr options
```{r set pander and knitr options, message=FALSE}
library(knitr)
library(pander)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')
```


# Load packages and functions for metabolite data analysis
```{r load packages and functions for metabolite data analysis, message=FALSE, fig.show='hide'}
# load packages
source("../functions/func_load_packages_metabolite_data_analysis.R")

# load functions
source("../functions/func_load_functions_metabolite_data_analysis.R")
```


# Paste file names with experiment_string
```{r paste file names with experiment_string, tidy=TRUE}
file_analytes_table <- paste("data/", analytes_table, ".txt", sep = "")
 
# file_gmd_values_select <- paste("output/", experiment_string, "/gmd_values_select.txt", sep = "")
# file_gmd_values_select_log10 <- paste("output/", experiment_string, "/gmd_values_select_log10.txt", sep = "")
# file_gmd_values_select_log10_woNA <- paste("output/", experiment_string, "/gmd_values_select_log10_woNA.txt", sep = "")
# 
# file_values_norm_log10_woOutlier <- paste("output/", experiment_string, "/values_norm_log10_woOutlier.txt", sep = "")
# file_gmd_values_select_log10_woOutlier <- paste("output/", experiment_string, "/gmd_values_select_log10_woOutlier.txt", sep = "")
```


# Set up database connection to GMD
```{r set up database connection to GMD}
login_gmd <- yaml.load_file("../functions/login_gmd.yaml")

if (.Platform$OS.type == 'windows'){
  dbhandle <- odbcDriverConnect('driver={SQL Server}; 
                                server=r710sqljahu; 
                                database=GMD; 
                                trusted_connection=true')
} else { # if .Platform$OS.type == 'unix'
  dbhandle <- odbcDriverConnect(
    connection=paste0("server=",login_gmd$host,
                      ";database=",login_gmd$db,
                      ";uid=",login_gmd$user,
                      ";pwd=",login_gmd$passwd,
                      ";Port=1433;driver=FreeTDS;TDS_Version=8.0;"))
}
```


# Load data
## Load phenotyper results
this table comes from ``phenotyper_metadata.Rmd``
```{r load phenotyper results}
phenotyper_result_joined <- read.table("data/phenotyper_result_joined_with_gmd_ids.txt", 
                                       sep="\t", header = T)
```


## Load analytes table
this table comes from ``gmd_analytes.Rmd``
```{r load analytes table}
analytes_sel_exp_sort <- read.table(file_analytes_table, sep ="\t", header = T, allowEscapes = T)
```


## Load `trost_TagList`
`trost_TagList` shows all experiments that belong to the TROST/VALDIS project and comes from `gmd_first_steps.Rmd`

```{r}
trost_TagList <- read.table("data/trost_TagList.txt", sep="\t", header=TRUE)

trost_gmd_ids <- read.table("data/trost_gmd_ids.txt", sep="\t", header=TRUE)
head(trost_gmd_ids)
```


## Load tables with GMD metadata
```{r load tables with GMD metadata}
# TROST
gmd_meta_mpi_field_2011 <- read.table("data/gmd_meta_mpi_field_2011.txt", sep = "\t", header = T)
gmd_meta_mpi_field_2012 <- read.table("data/gmd_meta_mpi_field_2012.txt", sep = "\t", header = T)
gmd_meta_jki_field_2012 <- read.table("data/gmd_meta_jki_field_2012.txt", sep = "\t", header = T)
gmd_meta_dethlingen_2011 <- read.table("data/gmd_meta_dethlingen_2011.txt", sep = "\t", header = T)
gmd_meta_jki_field_2013 <- read.table("../../TROST/GC-MS/GMD_Phenotyper/data/gmd_meta_jki_field_2013.txt", 
                                      sep = "\t", header = T)
gmd_meta_breeder_trials <- read.table("../../TROST/GC-MS/GMD_Phenotyper/data/gmd_meta_breeder_trials.txt", 
                                      sep = "\t", header = T)

colnames(gmd_meta_jki_field_2013) <- colnames(gmd_meta_breeder_trials) <- colnames(gmd_meta_jki_field_2012)

# VALDIS
gmd_meta_jki_shelter_2014 <- read.table("data/gmd_meta_jki_shelter_2014.txt", sep = "\t", header = T)
gmd_meta_jki_shelter_2015 <- read.table("data/gmd_meta_jki_shelter_2015.txt", sep = "\t", header = T)
gmd_meta_mpi_fgh_2014 <- read.table("data/gmd_meta_mpi_fgh_2014.txt", sep = "\t", header = T)
gmd_meta_mpi_fgh_2015 <- read.table("data/gmd_meta_mpi_fgh_2015.txt", sep = "\t", header = T)
gmd_meta_mpi_field_2015 <- read.table("data/gmd_meta_mpi_field_2015.txt", sep="\t", header = T)

#####################################

# exp_idx <- which(trost_TagList$experiment %in% experiment_names)
# trost_TagList_model <- droplevels(trost_TagList[exp_idx,])
# gmd_meta_per_exp_list <- vector("list", nrow(trost_TagList_model))
# for (i in trost_TagList_model$id){
#   gmd_meta_per_exp_list[[i]] <- func_get_gmd_metadata_3(i)
#   }
# gmd_meta_per_exp <- do.call(rbind, gmd_meta_per_exp_list)
# dim(gmd_meta_per_exp)
```


## Bind GMD metadata
```{r bind GMD metadata}
gmd_meta <- rbind(gmd_meta_mpi_field_2011, gmd_meta_mpi_field_2012, 
                  gmd_meta_jki_field_2012, gmd_meta_jki_field_2013,
                  gmd_meta_dethlingen_2011, gmd_meta_breeder_trials,
                  gmd_meta_jki_shelter_2014, gmd_meta_jki_shelter_2015,
                  gmd_meta_mpi_fgh_2014, gmd_meta_mpi_fgh_2015, gmd_meta_mpi_field_2015)

# change treatment
levels(gmd_meta$treatment)[4] <- levels(gmd_meta$treatment)[1] 

# change genotype names for cultivars
levels(gmd_meta$genotype_name)[1:34] <- names_cultivars_34

# use only control and stress treatments
gmd_meta_model <- droplevels(subset(gmd_meta, gmd_meta$treatment %in% c("Kontrolle", "Trockenstress")))
levels(gmd_meta_model$treatment) <- c("control", "drought stress")

# use only relevant genotypes (34 cultivars and 193 lines from 2014)
genotypes_model <- union(names_cultivars_34, names_lines_2014)
gmd_meta_model <- droplevels(subset(gmd_meta_model, gmd_meta_model$genotype_name %in% genotypes_model))

# use only late samples
gmd_meta_model <- droplevels(subset(gmd_meta_model, gmd_meta_model$sample_time == "late"))
dim(gmd_meta_model)
```


## Trial names
```{r trial names}
year <- rep("breeder_2011", nrow(gmd_meta_breeder_trials))
year[which(gmd_meta_breeder_trials$experiment_id %in% c("56876", "56878", "56879", 
                                                        "56880", "56881", "56882", 
                                                        "56883", "56884"))] = "breeder_2012"
year <- as.factor(year)
table(year)

mpi_feld_factors_trial <- as.factor(c(rep("mpi_feld_2011", 304),rep("mpi_feld_2012", 304)))
jki_feld_factors_trial <- as.factor(c(rep("jki_feld_2012", 136),rep("jki_feld_2013", 131)))
all_feld_factors_trial <- as.factor(c(rep("mpi_feld_2011", 304),rep("mpi_feld_2012", 304),
                                    rep("jki_feld_2012", 136),rep("jki_feld_2013", 131),
                                    rep("dethlingen_2011", 120)))

breeder_factors_trial <- gmd_meta_breeder_trials$experiment_name

valdis_factors_trial <- as.factor(c(rep("jki_shelter_2014", 384), rep("jki_shelter_2015", 252),
                                    rep("mpi_fgh_2014", 404), rep("mpi_fgh_2015", 254),
                                    rep("mpi_feld_2015", 377)))

trial <- as.factor(c(as.character(all_feld_factors_trial), 
                     as.character(breeder_factors_trial), 
                     as.character(valdis_factors_trial)))

trial_2 <- as.factor(c(as.character(all_feld_factors_trial), 
                       as.character(year), 
                       as.character(valdis_factors_trial)))
table(trial_2)

gmd_meta_model <- data.frame("trial" = trial, "trial_2" = trial_2, gmd_meta_model)

# the first 1530 lines contain only TROST samples
gmd_meta_model_trost <- droplevels(gmd_meta_model[1:1530,])
table(gmd_meta_model_trost$trial_2)

write.table(gmd_meta_model, "output/gmd_meta_model.txt", sep="\t")
write.table(gmd_meta_model_trost, "output/gmd_meta_model_trost.txt", sep="\t")
```


## Load tables with GMD values
```{r Load tables with GMD values}
gmd_values <- read.table("data/gmd_raw_values_matrix.txt", sep = "\t", header = T, check.names = F)
dim(gmd_values)
```


# Define subset of selected chromatograms 
```{r subset of chromatograms}
# select real biological samples
gmd_values_model <- subset(gmd_values, rownames(gmd_values) %in% gmd_meta_model$chromatogram)
```


## Remove outlier, see chunk XXX: find outlier
```{r remove outlier, see chunk XXX: find outlier}
# which(rownames(gmd_values_model)=="098BC4DF-28D8-4966-A089-E6F2B0405EDF")
# # 23
# gmd_values_part <- gmd_values_part[-23,]
# 
# which(gmd_meta$chromatogram =="098BC4DF-28D8-4966-A089-E6F2B0405EDF")
# # 23
# gmd_meta <- gmd_meta[-23,]
```


# Define subset of overlapping analytes
## SELECT: `r nrow(analytes_sel_exp_sort)` overlapping analytes regarding ALL TROST and VALDIS experiments, WITHOUT internal standards and contaminations
```{r subset of analytes}
gmd_values_select <- func_get_overlapping_analytes(analytes_sel_exp_sort, gmd_values_model)

write.table(gmd_values_select, "output/gmd_values_prediction_model.txt", sep="\t")
```


# Replace NAs with minimal values of dataset and calculate LOG10 of raw intensity values 
```{r replace NAs and calc LOG10 of raw values}
# log10 transformation (with NAs)
gmd_values_select_log10 <- log10(gmd_values_select)

# replace NAs with lowest value of dataset
gmd_values_select_woNA <- func_replace_na_with_min_value(gmd_values_select)

# log10 transformation after NA replacement -> without NAs 
gmd_values_select_log10_woNA <- log10(gmd_values_select_woNA)

sum(is.na(gmd_values_select_log10))
sum(is.na(gmd_values_select_log10_woNA))

write.table(gmd_values_select_log10, "output/gmd_values_prediction_model_log10.txt", sep="\t")
write.table(gmd_values_select_log10_woNA, "output/gmd_values_prediction_model_log10_woNA.txt", sep="\t")
```




# Save workspace
```{r save workspace}
save.image(paste("prediction_model_trost_valdis_", Sys.Date(), ".RData", sep="") )
```


# SessionInfo
```{r sessionInfo}
 sessionInfo()
```