---
title: "Prediction model with metabolite data - Load Data"
author: "Heike Sprenger"
date: "April 14, 2016"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_float: true
    toc_depth: 4
---

## Set working directory  
```{r setwd}
getwd()
#setwd("D:/work/repos/trost_metabolomics")
```


## Set pander and knitr options
```{r set pander and knitr options, message=FALSE}
library(knitr)
library(pander)
library(randomForest)
library(caret)
library(varSelRF)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# load workspace
# load("prediction_model.RData")
```


## Load packages and functions for metabolite data analysis
```{r load packages and functions for metabolite data analysis, message=FALSE, fig.show='hide'}
# load packages
source("../functions/func_load_packages_metabolite_data_analysis.R")

# load functions
source("../functions/func_load_functions_metabolite_data_analysis.R")
```


## Set up database connection to GMD
```{r set up database connection to GMD}
# login_gmd <- yaml.load_file("../functions/login_gmd.yaml")
# 
# if (.Platform$OS.type == 'windows'){
#   dbhandle <- odbcDriverConnect('driver={SQL Server}; 
#                                 server=r710sqljahu; 
#                                 database=GMD; 
#                                 trusted_connection=true')
# } else { # if .Platform$OS.type == 'unix'
#   dbhandle <- odbcDriverConnect(
#     connection=paste0("server=",login_gmd$host,
#                       ";database=",login_gmd$db,
#                       ";uid=",login_gmd$user,
#                       ";pwd=",login_gmd$passwd,
#                       ";Port=1433;driver=FreeTDS;TDS_Version=8.0;"))
# }
```


## Load phenotyper results and tolerance
this table comes from ``phenotyper_metadata.Rmd``
```{r load phenotyper results and tolerance}
phenotyper_result_joined <- read.table("data/phenotyper_result_joined_with_gmd_ids.txt", 
                                       sep="\t", header = T)

# SHOULD BE ZERO!
sum(is.na(phenotyper_result_joined$genotype_name))
# IF NOT EXECUTE ~\work\repos\database_scripts\update_plantlines_alias.sql
# AND REPEAT gmd_complete_workflow.Rmd

# tolerance
tolerance_table <- read.table("../trost_phenotypes/output/tolerance_phenotyper_metadata_2sub.txt", 
                        sep = "\t", header = TRUE)

# change names of cultivars from upper to lower case
levels(tolerance_table$cultivar) <- names_cultivars_31

# relevel tolerance factors
tolerance_table$tol_cat2_fve <- factor(tolerance_table$tol_cat2_fve, levels = c("low", "high"))
tolerance_table$tol_cat3_fve <- factor(tolerance_table$tol_cat3_fve, levels = c("low","mid", "high"))
```


## Load analytes table
this table comes from ``gmd_analytes.Rmd``
```{r load analytes table}

# 115 overlapping analytes
analytes_sel_exp_sort <- read.table("data/analytes_table_model_overlap_select_ordered.txt", 
                                    sep ="\t", header = T, allowEscapes = T)

# 74 overlapping analytes (TROST + VALDIS experiments)
analytes_sel_exp_sort_all <- read.table("data/analytes_table_20exp_overlap_select_ordered.txt", 
                                           sep ="\t", header = T, allowEscapes = T)
```


## Load `trost_TagList`
`trost_TagList` shows all experiments that belong to the TROST/VALDIS project and comes from `gmd_first_steps.Rmd`

```{r load trost taglist}
trost_TagList <- read.table("data/trost_TagList.txt", sep="\t", header=TRUE)

trost_gmd_ids <- read.table("data/trost_gmd_ids.txt", sep="\t", header=TRUE)
head(trost_gmd_ids)
```


## Load tables with GMD values
```{r Load tables with GMD values}
gmd_values <- read.table("data/gmd_raw_values_matrix.txt", sep = "\t", header = T, check.names = F)
dim(gmd_values)
```


## Load tables with GMD metadata
```{r load tables with GMD metadata}
# TROST
gmd_meta_mpi_field_2011 <- read.table("data/gmd_meta_mpi_field_2011.txt", sep = "\t", header = T)
gmd_meta_mpi_field_2012 <- read.table("data/gmd_meta_mpi_field_2012.txt", sep = "\t", header = T)
gmd_meta_jki_field_2012 <- read.table("data/gmd_meta_jki_field_2012.txt", sep = "\t", header = T)
gmd_meta_jki_field_2013 <- read.table("data/gmd_meta_jki_field_2013.txt", sep = "\t", header = T)
gmd_meta_dethlingen_2011 <- read.table("data/gmd_meta_dethlingen_2011.txt", sep = "\t", header = T)

gmd_meta_breeder_trials <- read.table("data/gmd_meta_breeder_trials.txt", sep = "\t", header = T)

# VALDIS
gmd_meta_jki_shelter_2014 <- read.table("data/gmd_meta_jki_shelter_2014.txt", sep = "\t", header = T)
gmd_meta_jki_shelter_2015 <- read.table("data/gmd_meta_jki_shelter_2015.txt", sep = "\t", header = T)
gmd_meta_mpi_fgh_2014 <- read.table("data/gmd_meta_mpi_fgh_2014.txt", sep = "\t", header = T)
gmd_meta_mpi_fgh_2015 <- read.table("data/gmd_meta_mpi_fgh_2015.txt", sep = "\t", header = T)
gmd_meta_mpi_field_2015 <- read.table("data/gmd_meta_mpi_field_2015.txt", sep="\t", header = T)

#####################################

# exp_idx <- which(trost_TagList$experiment %in% experiment_names)
# trost_TagList_model <- droplevels(trost_TagList[exp_idx,])
# gmd_meta_per_exp_list <- vector("list", nrow(trost_TagList_model))
# for (i in trost_TagList_model$id){
#   gmd_meta_per_exp_list[[i]] <- func_get_gmd_metadata_3(i)
#   }
# gmd_meta_per_exp <- do.call(rbind, gmd_meta_per_exp_list)
# dim(gmd_meta_per_exp)
```


## Bind GMD metadata

* After update of Phenotyper DB it might be necessary to repeat this script: ``~\work\repos\database_scripts\update_plantlines_alias.sql`` because genotype_name of some cultivars are NA (78)

```{r bind GMD metadata}
gmd_meta <- rbind(gmd_meta_mpi_field_2011, gmd_meta_mpi_field_2012, 
                  gmd_meta_jki_field_2012, gmd_meta_jki_field_2013,
                  gmd_meta_dethlingen_2011, gmd_meta_breeder_trials,
                  gmd_meta_jki_shelter_2014, gmd_meta_jki_shelter_2015,
                  gmd_meta_mpi_fgh_2014, gmd_meta_mpi_fgh_2015, gmd_meta_mpi_field_2015)
dim(gmd_meta)

# change treatment: 50% nFK to Kontrolle
levels(gmd_meta$treatment)[4] <- levels(gmd_meta$treatment)[1] 

# change genotype names for cultivars
levels(gmd_meta$genotype_name)[1:34] <- names_cultivars_34

# use only control and stress treatments
gmd_meta_2 <- droplevels(subset(gmd_meta, gmd_meta$treatment %in% c("Kontrolle", "Trockenstress")))
levels(gmd_meta_2$treatment) <- c("control", "drought stress")
dim(gmd_meta_2)

# use only late samples
gmd_meta_3 <- droplevels(subset(gmd_meta_2, gmd_meta_2$sample_time == "late"))
dim(gmd_meta_3)

# use only relevant genotypes (34 cultivars and 193 lines from 2014)
genotypes_model <- union(names_cultivars_34, names_lines_2014)
gmd_meta_model <- droplevels(subset(gmd_meta_3, gmd_meta_3$genotype_name %in% genotypes_model))
dim(gmd_meta_model)

table(gmd_meta_model$experiment_name)
```


## Trial names
```{r trial names}
year <- rep("breeder_2011", nrow(gmd_meta_breeder_trials))
year[which(gmd_meta_breeder_trials$experiment_id %in% c("56876", "56878", "56879", 
                                                        "56880", "56881", "56882", 
                                                        "56883", "56884"))] = "breeder_2012"
year <- as.factor(year)
table(year)

mpi_feld_factors_trial <- as.factor(c(rep("mpi_feld_2011", 304),rep("mpi_feld_2012", 304)))
jki_feld_factors_trial <- as.factor(c(rep("jki_feld_2012", 136),rep("jki_feld_2013", 133)))
all_feld_factors_trial <- as.factor(c(rep("mpi_feld_2011", 304),rep("mpi_feld_2012", 304),
                                    rep("jki_feld_2012", 136),rep("jki_feld_2013", 133),
                                    rep("dethlingen_2011", 120)))

breeder_factors_trial <- gmd_meta_breeder_trials$experiment_name

valdis_factors_trial <- as.factor(c(rep("jki_shelter_2014", 384), rep("jki_shelter_2015", 252),
                                    rep("mpi_fgh_2014", 404), rep("mpi_fgh_2015", 254),
                                    rep("mpi_feld_2015", 377)))

trial <- as.factor(c(as.character(all_feld_factors_trial), 
                     as.character(breeder_factors_trial), 
                     as.character(valdis_factors_trial)))

trial_2 <- as.factor(c(as.character(all_feld_factors_trial), 
                       as.character(year), 
                       as.character(valdis_factors_trial)))
table(trial_2)

gmd_meta_model <- data.frame("trial" = trial, "trial_2" = trial_2, gmd_meta_model)
```


## Column for genotype class (for VALDIS lines)
```{r column for genotype class}
# get all names that start with AR or EA
AxR <- which(grepl("^AR", gmd_meta_model$genotype_name))
ExA <- which(grepl("^EA", gmd_meta_model$genotype_name))

genotype_class <- as.character(gmd_meta_model$genotype_name)
genotype_class[AxR] <- "AxR"
genotype_class[ExA] <- "ExA"
gmd_meta_model$genotype_class <- factor(genotype_class)
```


## Subset of metadata for field, TROST or VALDIS
```{r subset of metadata for field, trost or valdis}
# sort table by chromatogram ID
gmd_meta_model$chromatogram <- as.character(gmd_meta_model$chromatogram)
gmd_meta_model_sort <- gmd_meta_model[order(gmd_meta_model$chromatogram),]

# only the 997 samples from trost field trials
feld_idx <- which(gmd_meta_model_sort$trial %in% levels(all_feld_factors_trial))
feld_chrom <- gmd_meta_model_sort$chromatogram[feld_idx]
length(feld_chrom)

gmd_meta_model_feld <- droplevels(gmd_meta_model_sort[feld_idx,])
table(gmd_meta_model_feld$trial_2)

# only the 1532 samples from trost field and breeder trials
trost_idx <- which(gmd_meta_model_sort$trial_2 %in% union(levels(all_feld_factors_trial), levels(year)))
trost_chrom <- gmd_meta_model_sort$chromatogram[trost_idx]
length(trost_chrom)

gmd_meta_model_trost <- droplevels(gmd_meta_model_sort[trost_idx,])
table(gmd_meta_model_trost$trial_2)

# only 1671 valdis samples
valdis_idx <- which(gmd_meta_model_sort$trial %in% levels(valdis_factors_trial))
valdis_chrom <- gmd_meta_model_sort$chromatogram[valdis_idx]
length(valdis_chrom)

gmd_meta_model_valdis <- droplevels(gmd_meta_model_sort[valdis_idx,])
dim(gmd_meta_model_valdis)
```


## Export final metadata tables
```{r export final metadata tables}
write.table(gmd_meta_model, "output/prediction_model/gmd_meta_model.txt", sep="\t")
write.table(gmd_meta_model_sort, "output/prediction_model/gmd_meta_model_sort.txt", sep="\t")
write.table(gmd_meta_model_feld, "output/prediction_model/gmd_meta_model_feld.txt", sep="\t")
write.table(gmd_meta_model_trost, "output/prediction_model/gmd_meta_model_trost.txt", sep="\t")
```


## Define subset of selected chromatograms 
```{r subset of chromatograms}
# select real biological samples
gmd_values_model <- subset(gmd_values, rownames(gmd_values) %in% gmd_meta_model$chromatogram)
dim(gmd_values_model)
# 3203 566
```


### Remove outlier, see chunk XXX: find outlier
```{r remove outlier, see chunk XXX: find outlier}
# which(rownames(gmd_values_model)=="098BC4DF-28D8-4966-A089-E6F2B0405EDF")
# # 23
# gmd_values_part <- gmd_values_part[-23,]
# 
# which(gmd_meta$chromatogram =="098BC4DF-28D8-4966-A089-E6F2B0405EDF")
# # 23
# gmd_meta <- gmd_meta[-23,]
```


## Define subset of overlapping analytes
* SELECT: `r nrow(analytes_sel_exp_sort)` overlapping analytes regarding ALL TROST and VALDIS experiments, WITHOUT internal standards and contaminations
```{r subset of analytes}
gmd_values_select <- func_get_overlapping_analytes(analytes_sel_exp_sort, gmd_values_model)
dim(gmd_values_select)
# 3203 115

# only the values for 1532 TROST samples 
gmd_values_select_trost <- gmd_values_select[trost_idx, ]
dim(gmd_values_select_trost)
head(rownames(gmd_values_select_trost))

# 74 analytes overlapping for all TROST and VALDIS data (20 experiments)
gmd_values_select_all <- func_get_overlapping_analytes(analytes_sel_exp_sort_all, gmd_values_model)
dim(gmd_values_select_all)
# 3203 74 

write.table(gmd_values_select, "output/prediction_model/gmd_values_prediction_model.txt", sep="\t")
write.table(gmd_values_select_trost, "output/prediction_model/gmd_values_trost_prediction_model.txt", sep="\t")
write.table(gmd_values_select_all, "output/prediction_model/gmd_values_all_prediction_model.txt", sep="\t")
```


## Save workspace
```{r save workspace}
save.image("prediction_model.RData")
```


