---
title: "Prediction model with metabolite data - Load Data"
author: "Heike Sprenger"
date: "April 14, 2016"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_float: true
    toc_depth: 4
---

## Set working directory  
```{r setwd}
getwd()
#setwd("D:/work/repos/trost_metabolomics")
```


## Set pander and knitr options
```{r set pander and knitr options, message=FALSE}
library(knitr)
library(pander)
library(randomForest)
library(caret)
library(varSelRF)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# load workspace
# load("prediction_model.RData")
```


## Load workspace
```{r load workspace}
load("prediction_model.RData")
```


## Calculate percentage of NAs and plot histograms

* print percentage of NAs, number of analytes with more than 40% NAs and ID of those analytes

```{r calculate percentage of NAs and plot histograms}
func_print_na_statistics(gmd_values_select, analytes_sel_exp_sort$Name)

pdf("figures/prediction_model/NAs_hist_all_samples.pdf")
func_plot_na_statistics(gmd_values_select)
dev.off()


# only the values for 1532 TROST samples 
func_print_na_statistics(gmd_values_select_trost, analytes_sel_exp_sort$Name)

pdf("figures/prediction_model/NAs_hist_trost_samples.pdf")
func_plot_na_statistics(gmd_values_select_trost)
dev.off()
```


## Replace NAs with minimal values of dataset and calculate LOG10 of raw intensity values 
```{r replace NAs and calc LOG10 of raw values}
# log10 transformation (with NAs)
gmd_values_select_log10 <- log10(gmd_values_select)
gmd_values_select_trost_log10 <- log10(gmd_values_select_trost)

# replace NAs with lowest value of dataset
gmd_values_select_woNA <- func_replace_na_with_min_value(gmd_values_select)
gmd_values_select_trost_woNA <- func_replace_na_with_min_value(gmd_values_select_trost)

# log10 transformation after NA replacement -> without NAs 
gmd_values_select_log10_woNA <- log10(gmd_values_select_woNA)
gmd_values_select_trost_log10_woNA <- log10(gmd_values_select_trost_woNA)

sum(is.na(gmd_values_select_log10))
sum(is.na(gmd_values_select_log10_woNA))

# all
write.table(gmd_values_select_log10, "output/prediction_model/gmd_values_prediction_model_log10.txt", sep="\t")
write.table(gmd_values_select_log10_woNA, "output/prediction_model/gmd_values_prediction_model_log10_woNA.txt", sep="\t")
# trost
write.table(gmd_values_select_trost_log10, "output/prediction_model/gmd_values_prediction_model_trost_log10.txt", sep="\t")
write.table(gmd_values_select_trost_log10_woNA, "output/prediction_model/gmd_values_prediction_model_trost_log10_woNA.txt", sep="\t")
```


## Extract Batch and SequenceID
```{r Extract Batch and SequenceID}
gmd_meta_model_trost$BatchID <- func_get_batch_ids(gmd_meta_model_trost)
gmd_meta_model_trost$SequenceID <- func_get_sequence_ids(gmd_meta_model_trost)
gmd_meta_model_trost$log10_AvgAnnotated <- log10(gmd_meta_model_trost$AvgAnnotated)

gmd_meta_model_sort$BatchID <- func_get_batch_ids(gmd_meta_model_sort)
gmd_meta_model_sort$SequenceID <- func_get_sequence_ids(gmd_meta_model_sort)
gmd_meta_model_sort$log10_AvgAnnotated <- log10(gmd_meta_model_sort$AvgAnnotated)
```


## Relevel factors and create factors table
```{r relevel factors and create factors table}
# define tolerance factor
tolerance <- func_create_tolerance_factor(gmd_meta_model_sort)
tolerance_trost <- func_create_tolerance_factor(gmd_meta_model_trost)

# create factors table for treatment, cultivar, sample_time, Dw, Fw, Is, AvgAnnotated, AvgAnnotatedLog10, BatchID, SequenceID
factors <- func_create_factors_table(gmd_meta_model_sort, 
                                     gmd_meta_model_sort$SequenceID, 
                                     gmd_meta_model_sort$BatchID, 
                                     tolerance)
levels(factors$genotype_class)

# only TROST data, without VALDIS
factors_trost <- func_create_factors_table(gmd_meta_model_trost, 
                                     gmd_meta_model_trost$SequenceID, 
                                     gmd_meta_model_trost$BatchID, 
                                     tolerance_trost)

levels(factors_trost$genotype_class)

write.table(factors, "output/prediction_model/factors_prediction_model.txt", sep="\t")
write.table(factors_trost, "output/prediction_model/factors_prediction_model_trost.txt", sep="\t")
```


## ANOVA normalization: only TROST
```{r ANOVA normalization only TROST}
dim(gmd_values_select_trost_log10)

# without NAs (replaced by min)
norm_values_trost_log10_woNA <- apply(gmd_values_select_trost_log10_woNA, 2, RemoveFactors, sam = factors_trost, 
                              facs = c("genotype_class", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"), 
                              keep = c("genotype_class", "treatment"))

# with NAs
norm_values_trost_log10 <- apply(gmd_values_select_trost_log10, 2, RemoveFactors, sam = factors_trost, 
                              facs = c("genotype_class", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"), 
                              keep = c("genotype_class", "treatment"))

sum(is.na(norm_values_trost_log10)) / (ncol(norm_values_trost_log10) * nrow(norm_values_trost_log10))*100
# 5.38%

write.table(norm_values_trost_log10_woNA, "output/prediction_model/norm_values_trost_log10_woNA_predicion_model.txt", sep="\t")
write.table(norm_values_trost_log10, "output/prediction_model/norm_values_trost_log10_prediction_model.txt", sep="\t")
```


## ANOVA normalization: TROST and VALDIS
```{r ANOVA normalization TROST and VALDIS}
# all TROST + VALDIS trials, with/without NA
# gmd_values_select_log10
# gmd_values_select_log10_woNA

dim(gmd_values_select_log10)

# without NAs (replaced by min)
norm_values_log10_woNA <- apply(gmd_values_select_log10_woNA, 2, RemoveFactors, sam = factors, 
                              facs = c("genotype_class", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"), 
                              keep = c("genotype_class", "treatment"))

# equivalent to all_valdis_merge_IIIc
# with NAs
norm_values_log10 <- apply(gmd_values_select_log10, 2, RemoveFactors, sam = factors, 
                              facs = c("genotype_class", "treatment", "SequenceID", "BatchID", "log10_AvgAnnotated"), 
                              keep = c("genotype_class", "treatment"))

sum(is.na(norm_values_log10)) / (ncol(norm_values_log10) * nrow(norm_values_log10))*100
# 13.93%

write.table(norm_values_log10_woNA, "output/prediction_model/norm_values_log10_woNA_predicion_model.txt", sep="\t")
write.table(norm_values_log10, "output/prediction_model/norm_values_log10_prediction_model.txt", sep="\t")
```


## PCA
### PCA only TROST
```{r PCA only TROST}
# only TROST
gmd_values_trost_log10_prep <- prep(gmd_values_select_trost_log10, scale = "none", center = F) # before normalization
norm_values_trost_log10_prep <- prep(norm_values_trost_log10, scale = "none", center = FALSE) # after normalization (with NA)

gmd_values_trost_log10_prep_rnipals <- pca(gmd_values_trost_log10_prep, nPcs = 5, method="rnipals")
norm_values_trost_log10_prep_rnipals <- pca(norm_values_trost_log10_prep, nPcs = 5, method="rnipals")

# complete observations of TROST data
compObs_trost <- norm_values_trost_log10_prep_rnipals@completeObs
dim(compObs_trost)

# export normalized values with complete observations (no NAs) for model
write.table(compObs_trost, "output/prediction_model/norm_values_trost_log10_rnipals_completeObs.txt", sep="\t")
```


### PCA plots only TROST
```{r PCA plots only TROST}
# trial effect
# before normalization
pairs(gmd_values_trost_log10_prep_rnipals@scores, col = gmd_meta_model_trost$trial_2)
plot(gmd_values_trost_log10_prep_rnipals@scores[,1],
     gmd_values_trost_log10_prep_rnipals@scores[,2], 
     col = gmd_meta_model_trost$trial_2, pch = 19,
     main = "PCA effect of trial before normalization")
legend("topright", legend = levels(gmd_meta_model_trost$trial_2), 
       fill = 1:7, cex = 0.8)

# after normalization
pairs(norm_values_trost_log10_prep_rnipals@scores, col = gmd_meta_model_trost$trial_2)
plot(norm_values_trost_log10_prep_rnipals@scores[,1],
     norm_values_trost_log10_prep_rnipals@scores[,2], 
     col = gmd_meta_model_trost$trial_2, pch = 19,
     main = "PCA effect of trial after normalization")
legend("bottomright", legend = levels(gmd_meta_model_trost$trial_2), 
       fill = 1:7, cex = 0.8)

# genotype effect
palette(rainbow(34))
pairs(norm_values_trost_log10_prep_rnipals@scores, col = gmd_meta_model_trost$genotype_class)

# PC1 vs PC2
plot(norm_values_trost_log10_prep_rnipals@scores[,1],
     norm_values_trost_log10_prep_rnipals@scores[,2], 
     col = gmd_meta_model_trost$genotype_class, pch = 19,
     main = "PCA effect of genotype after normalization")
legend("bottomright", legend = levels(gmd_meta_model_trost$genotype_class), 
       fill = 1:34, cex = 0.6)

# PC1 vs PC3
plot(norm_values_trost_log10_prep_rnipals@scores[,1],
     norm_values_trost_log10_prep_rnipals@scores[,3], 
     col = gmd_meta_model_trost$genotype_class, pch = 19,
     main = "PCA effect of genotype after normalization")
legend("bottomright", legend = levels(gmd_meta_model_trost$genotype_class), 
       fill = 1:34, cex = 0.6)
```


### PCA TROST and VALDIS
```{r PCA TROST and VALDIS}
# all TROST + VALDIS trials
gmd_values_log10_prep <- prep(gmd_values_select_log10, scale = "none", center = FALSE) # before normalization
norm_values_log10_prep <- prep(norm_values_log10, scale = "none", center = FALSE) # after normalization (with NA)
#all_valdis_merge_IIIc_woNA_none <- prep(all_valdis_merge_IIIc_woNA, scale = "none", center = FALSE) # after normalization

gmd_values_log10_prep_rnipals <- pca(gmd_values_log10_prep, nPcs = 5, method="rnipals")
norm_values_log10_prep_rnipals <- pca(norm_values_log10_prep, nPcs = 5, method="rnipals")

# complete observations of TROST and VALDIS data
compObs <- norm_values_log10_prep_rnipals@completeObs
dim(compObs)

# export normalized values with complete observations (no NAs) for model
write.table(compObs, "output/prediction_model/norm_values_log10_rnipals_completeObs.txt", sep="\t")
```


### PCA plots TROST and VALDIS
```{r PCA plots TROST and VALDIS}
# effect of treatment
palette(cols_treatment)
pairs(norm_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$treatment, pch = 19)

# effect of genotype
palette(rainbow(36))
pairs(gmd_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$genotype_class)
pairs(norm_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$genotype_class)

plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,2],
     col = gmd_meta_model_sort$genotype_class, 
     pch=c(17,19)[gmd_meta_model_sort$treatment],
     main = "PCA effect of genotype after normalization")
text(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,2],
     labels = gmd_meta_model_sort$genotype_class, cex=0.4)
legend("topright", legend = levels(gmd_meta_model_sort$genotype_class), fill=1:36, cex=0.6)


# effect of trial
palette(rainbow(12))
# before normalization
pairs(gmd_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$trial_2)

plot(gmd_values_log10_prep_rnipals@scores[,1],
     gmd_values_log10_prep_rnipals@scores[,2],
     col = gmd_meta_model_sort$trial_2, pch = 19,
     main = "PCA effect of trial before normalization")
legend("topright", legend = levels(gmd_meta_model_sort$trial_2), fill=1:12, cex=0.6)


# after normalization
pairs(norm_values_log10_prep_rnipals@scores, col = gmd_meta_model_sort$trial_2)

plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,2],
     col = gmd_meta_model_sort$trial_2, pch=19,
     main = "PCA effect of trial after normalization")
legend("topright", legend = levels(gmd_meta_model_sort$trial_2), fill=1:12, cex=0.6)
```


### PCA Loadings
```{r loadings}
# top6 loadings for PC2
head(sort(norm_values_log10_prep_rnipals@loadings[,2]))
tail(sort(norm_values_log10_prep_rnipals@loadings[,2]))

# A179012 shows high loading value for PC2
summary(norm_values_log10[,"790E11C9-30D1-40BE-A2BB-89B08A545722"])
hist(norm_values_log10[,"790E11C9-30D1-40BE-A2BB-89B08A545722"], breaks=30)

# there is a strong effect of genotype on metabolite level
boxplot(norm_values_log10[,"790E11C9-30D1-40BE-A2BB-89B08A545722"] ~ gmd_meta_model_sort$genotype_class, las = 2)

A179012_low <- which(norm_values_log10[,"790E11C9-30D1-40BE-A2BB-89B08A545722"] < 3.5)
A179012_high <- which(norm_values_log10[,"790E11C9-30D1-40BE-A2BB-89B08A545722"] > 3.5)
A179012_level <- rep("high", 3203)
A179012_level[A179012_low] <- "low"
A179012_level <- factor(A179012_level)
table(A179012_level)

palette(heike_palette_2)
plot(norm_values_log10_prep_rnipals@scores[,1],
     norm_values_log10_prep_rnipals@scores[,2],
     col = A179012_level, pch = 19,
     main = "PCA effect of A179012 level after normalization")
```


## Define training and test data
```{r define training/test data}
# rownames of completeObs
compObs_chrom <- rownames(compObs_trost)
length(compObs_chrom)

# join gmd_meta with tolerance table
gmd_meta_model_feld_joined <- droplevels(merge(gmd_meta_model_feld, tolerance_table, 
                                    by.x = "genotype_name", by.y = "cultivar"))
# sort table by chromatogram ID
gmd_meta_model_feld_joined <- gmd_meta_model_feld_joined[order(gmd_meta_model_feld_joined$chromatogram),]
dim(gmd_meta_model_feld)
dim(gmd_meta_model_feld_joined)

# only the 913 samples (of 997) from trost field trials, WITHOUT 3 cultivars
feld_model_idx <- which(compObs_chrom %in% gmd_meta_model_feld_joined$chromatogram)
feld_model_chrom <- compObs_chrom[feld_model_idx]
length(feld_model_chrom)

feld_model_data <- compObs_trost[feld_model_idx,]
dim(feld_model_data)
head(rownames(feld_model_data))

# test/train data and info
train_idx <- which(gmd_meta_model_feld_joined$model_set=="train")
test_idx <- which(gmd_meta_model_feld_joined$model_set=="test")

train_data <- feld_model_data[train_idx, ]
test_data <- feld_model_data[test_idx, ]
colnames(train_data) = analytes_sel_exp_sort$analyteID
colnames(test_data) = analytes_sel_exp_sort$analyteID

# with short metabolite names
#train_data2 <- all_merge_IIIc_field_exp_compObs[train_idx, ]
#test_data2 <- all_merge_IIIc_field_exp_compObs[test_idx, ]
#colnames(train_data2) = analytes_valdis$name_short
#colnames(test_data2) = analytes_valdis$name_short

train_info <- gmd_meta_model_feld_joined[train_idx,]
test_info <- gmd_meta_model_feld_joined[test_idx,]

## only control
train_data_control <- subset (train_data, train_info$treatment=="control")
test_data_control <- subset (test_data, test_info$treatment=="control")
colnames(train_data_control) = analytes_sel_exp_sort$analyteID
colnames(test_data_control) = analytes_sel_exp_sort$analyteID

train_info_control <- subset (train_info, train_info$treatment=="control")
test_info_control <- subset (test_info, test_info$treatment=="control")
```


## Define ALL data (without 3 cultivars --> model_set=="NA") and subset of all control samples
```{r all data (control)}
all_data <- subset(feld_model_data, !gmd_meta_model_feld_joined$model_set=="NA")
colnames(all_data) = analytes_sel_exp_sort$analyteID

all_info <- subset (gmd_meta_model_feld_joined, !gmd_meta_model_feld_joined$model_set=="NA")

# all control samples (ohne NA -> 3 Kutlivare)
all_data_control <- subset(feld_model_data, 
                           gmd_meta_model_feld_joined$treatment=="control" 
                           & !gmd_meta_model_feld_joined$model_set=="NA")

colnames(all_data_control) = analytes_sel_exp_sort$analyteID

all_info_control <- subset (gmd_meta_model_feld_joined, 
                            gmd_meta_model_feld_joined$treatment=="control" 
                            & !gmd_meta_model_feld_joined$model_set=="NA")
```


## Define VALDIS data
```{r define valdis data}
# length(valdis_idx)
# # 1671
# 
# valdis_data <- compObs[valdis_idx,]
# dim(valdis_data)
# colnames(valdis_data) = analytes_sel_exp_sort$analyteID

# check
#match(rownames(valdis_data), gmd_meta_model_valdis$chromatogram)
```


## Save workspace
```{r save workspace}
save.image("prediction_model.RData")
```


