---
title: "Prediction model with metabolite data - Load Data"
author: "Heike Sprenger"
date: "April 14, 2016"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---

## Set working directory  
```{r setwd}
getwd()
#setwd("D:/work/repos/trost_metabolomics")
```


## Set pander and knitr options
```{r set pander and knitr options, message=FALSE}
library(knitr)
library(pander)
library(randomForest)
library(caret)
library(varSelRF)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')
```


## Load workspace
```{r load workspace}
load("prediction_model.RData")
```


## Random forest classification
### Random forest classification: train data with 3 classes
```{r rf cat3 train}
input_cat3_train <- data.frame(train_data, "tol" = train_info$tol_cat3_fve, check.names = FALSE)
#input_cat3_train <- data.frame(train_data, "tol" = train_info$tol_cat3_fve)
levels(input_cat3_train$tol)

i <- match("tol", names(input_cat3_train)) # i is index of tol column

set.seed(1234)
#rf_cat3_train <- randomForest(tol~. , data=input_cat3_train, ntree=1000)
rf_cat3_train <- randomForest(input_cat3_train[,-i], input_cat3_train[,i], ntree=1000)
print(rf_cat3_train)
confusionMatrix(rf_cat3_train$predicted, train_info$tol_cat3_fve)

head(importance(rf_cat3_train))
tail(sort(importance(rf_cat3_train)))
varImpPlot(rf_cat3_train, main="")


table(predict(rf_cat3_train, test_data), test_info$tol_cat3_fve)
confusionMatrix(predict(rf_cat3_train, test_data), test_info$tol_cat3_fve)
#table(test_info$tol_cat3_fve)


#######################################
# USED FOR FINAL TROST PROJECT REPORT #
#######################################

# 
# ###############################
# # with real metabolite names
# input_cat3_train2 <- as.data.frame( cbind(train_data2, "tol"=train_info$tol_cat3_fve))
# input_cat3_train2$tol <- train_info$tol_cat3_fve
# levels(input_cat3_train2$tol)
# 
# set.seed(1234)
# rf_cat3_train2 <- randomForest(tol~. , data=input_cat3_train2, ntree=1000)
# print(rf_cat3_train)
# 
# tail(sort(importance(rf_cat3_train)))
# varImpPlot(rf_cat3_train, main="")
```


### Cross-Validation
```{r rf cat3 train CV}
set.seed(1)
rf_cat3_train_cv <- rfcv(train_data, train_info$tol_cat3_fve, step=0.8)
rf_cat3_train_cv$n.var
length(rf_cat3_train_cv$n.var)
with(rf_cat3_train_cv, plot(n.var, error.cv, log="x", type="o", lwd=2))
```


## Random forest classification: all data with 3 classes
```{r rf cat3 all}
input_cat3_all <- data.frame(all_data, "tol" = all_info$tol_cat3_fve, check.names = FALSE)
#input_cat3_train <- data.frame(train_data, "tol" = train_info$tol_cat3_fve)
levels(input_cat3_all$tol)

i <- match("tol", names(input_cat3_all)) # i is index of tol column

set.seed(1234)
#rf_cat3_train <- randomForest(tol~. , data=input_cat3_train, ntree=1000)
rf_cat3_all <- randomForest(input_cat3_all[,-i], input_cat3_all[,i], ntree=1000)
print(rf_cat3_all)
confusionMatrix(rf_cat3_all$predicted, all_info$tol_cat3_fve)
table(rf_cat3_all$predicted, all_info$tol_cat3_fve)

head(importance(rf_cat3_all))
tail(sort(importance(rf_cat3_all)))
varImpPlot(rf_cat3_all, main="")
```


### Cross-Validation
```{r rf cat3 all CV}
set.seed(1)
rf_cat3_all_cv <- rfcv(all_data, all_info$tol_cat3_fve, step=0.8)
rf_cat3_all_cv$n.var
length(rf_cat3_all_cv$n.var)
with(rf_cat3_all_cv, plot(n.var, error.cv, log="x", type="o", lwd=2))
```


### Variable Selection (vs)
```{r rf cat3 all VS}
set.seed(2)
rf_cat3_all_vs <- varSelRF(all_data, all_info$tol_cat3_fve, ntree = 500, 
                           ntreeIterat = 300, vars.drop.frac = 0.2, c.sd=1)
rf_cat3_all_vs

plot(rf_cat3_all_vs, which=1)
plot(rf_cat3_all_vs, which=2)

rf_cat3_all_vs_idx <- which(colnames(all_data) %in% rf_cat3_all_vs$selected.vars)
```


## Try LOO-CV with single cultivars for test set

* Idea: use 30 of 31 cultivars in training and the remaining for test

```{r try LOO-CV}
table(gmd_meta_model_feld_joined$genotype_name)
length(levels(gmd_meta_model_feld_joined$genotype_name))

genotype_levels <- levels(gmd_meta_model_feld_joined$genotype_name)

loocv_accuracy <- rep(NA, 31)
names(loocv_accuracy) <- genotype_levels

for (i in genotype_levels){
  loocv_training_data <- subset(feld_model_data, !gmd_meta_model_feld_joined$genotype_name == i)
  loocv_test_data <- subset(feld_model_data, gmd_meta_model_feld_joined$genotype_name == i)
  
  loocv_training_info <- subset(gmd_meta_model_feld_joined, !gmd_meta_model_feld_joined$genotype_name == i)
  loocv_test_info <- subset(gmd_meta_model_feld_joined, gmd_meta_model_feld_joined$genotype_name == i)
  
  loocv_input <- data.frame(loocv_training_data, "tol" = loocv_training_info$tol_cat3_fve, check.names = FALSE)
  tol <- match("tol", names(loocv_input)) # i is index of tol column

  set.seed(1234)
  loocv_rf <- randomForest(loocv_input[,-tol], loocv_input[,tol], ntree=1000)
  #table(predict(loocv_rf, loocv_test_data), loocv_test_info$tol_cat3_fve)
  
  # save accuracy
  loocv_accuracy[i] <- confusionMatrix(predict(loocv_rf, loocv_test_data), 
                                       loocv_test_info$tol_cat3_fve)$overall[1]
}

loocv_accuracy

```


## Save workspace
```{r save workspace}
save.image("prediction_model.RData")
```

