---
title: "QC concentration series data analysis"
author: "Heike Sprenger"
date: "Wednesday, February 25, 2016"
output:
  html_document:
    highlight: tango
    number_section: yes
    theme: cerulean
    toc: yes
    toc_depth: 4
---


# Set options
```{r set-options, echo=FALSE, cache=FALSE}
options(width=600)
```


# Set working directory and load workspace
```{r set up, message=FALSE}
#setwd("D:/work/repos/trost_metabolomics")
#setwd("~/work/repos/trost_metabolomics/")
#load("qc_concentration_series.RData")

library(knitr)
library(pander)
library(reshape)
library(plyr)
library(gplots)
library(plyr)
library(dendextend)
library(MetNorm)
library(RODBC)
library(RMySQL)
library(yaml)
library(pcaMethods)

# set options for pander
panderOptions('table.split.table', 200)

# set options for knitr
opts_chunk$set(fig.width=5, fig.height=5, cache=FALSE, highlight = TRUE, fig.show="asis")
opts_knit$set(root.dir = '../')

# load workspace
# load("yield_data_valdis.RData")
```


# Source R functions
```{r source R functions, include=FALSE}
source("../functions/func_get_experiment_id.R")
source("../functions/func_get_gmd_metadata.R")
source("../functions/func_cv.R")
source("../functions/colors.R")
source("../functions/RemoveFactors_function.R")
source("../functions/func_normalize.R")
```


# Set up database connection to GMD
[How to access MS-SQL databases in linux](http://wiki.mpimp-golm.mpg.de/index.php/Access_MSSQL_databases)
[How to identify operating system from R](http://conjugateprior.org/2015/06/identifying-the-os-from-r/)

```{r set up database connection to GMD}

login_gmd <- yaml.load_file("../functions/login_gmd.yaml")

if (.Platform$OS.type == 'windows'){
  dbhandle <- odbcDriverConnect('driver={SQL Server}; 
                                server=r710sqljahu; 
                                database=GMD; 
                                trusted_connection=true')
} else { # if .Platform$OS.type == 'unix'
  dbhandle <- odbcDriverConnect(
    connection=paste0("server=",login_gmd$host,
                      ";database=",login_gmd$db,
                      ";uid=",login_gmd$user,
                      ";pwd=",login_gmd$passwd,
                      ";Port=1433;driver=FreeTDS;TDS_Version=8.0;"))
}
```

# Read tables: ``trost_TagList``

* ``trost_TagList`` shows all experiments that belong to the TROST/VALDIS project and comes from ``gmd_first_steps.Rmd``

```{r read tables}
trost_TagList <- read.table("data/trost_TagList.txt", sep="\t", header=TRUE)

trost_gmd_ids <- read.table("data/trost_gmd_ids.txt", sep="\t", header=TRUE)
head(trost_gmd_ids)
```


# Load information about QC samples
```{r load information about QC samples}
qc_info_blank <- read.table("data/qc_concentration_series_with_blanks.txt", sep = "\t", header = T, check.names = F)

dim(qc_info_blank)
head(qc_info_blank)
colnames(qc_info_blank)[1] <- "GMD_id"

qc_info_blank$concentration <- factor(qc_info_blank$concentration, levels=c("t0","t6", "t12", "t25", "t50", "t100", "t200"))
table(qc_info_blank$concentration)

# integer
class(qc_info_blank$sequence_id)

qc_info_blank_with_ids <- join(qc_info_blank, trost_gmd_ids, by="GMD_id", type = "inner")
dim(qc_info_blank_with_ids)
```


# Get GMD meta data
**use function ``func_get_gmd_metadata_3`` that omits FW and DW but returns Sorbitol (Is) and AvgAnnotated per chromatogram**

[Solution for rbind and loop](http://stackoverflow.com/questions/26843861/replace-rbind-in-for-loop-with-lapply-2nd-circle-of-hell)

```{r get GMD meta data}
remove_exp <- c("QC_1", "QC_2", "QC_3", "QC_4", "MPI_Test2_old")
remove_exp_idx <- which(trost_TagList$experiment %in% remove_exp)

trost_TagList_qc <- trost_TagList[-remove_exp_idx,]

gmd_meta_per_exp_list <- vector("list", nrow(trost_TagList_qc))

for (i in trost_TagList_qc$id){
  gmd_meta_per_exp_list[[i]] <- func_get_gmd_metadata_3(i)
  }

gmd_meta_per_exp <- do.call(rbind, gmd_meta_per_exp_list)
dim(gmd_meta_per_exp)
```


## Remove duplicate entries in gmd_meta
* these 6 gmd_ids were assigned to 2 experiments: MPI FGH 2014 and JKI Shelter 2014
* remove entries for MPI FGH 2014 (9519AAB5-12D3-47C6-995C-02F25F58CD67)

```{r remove duplicate entries in gmd_meta}
duplicate_gmd_ids <- c('14301if_6','14301if_7','14301if_8','14301if_9','14301if_10','14301if_11')
duplicate_chromatograms <- qc_info_blank_with_ids$chromatogram[which(qc_info_blank_with_ids$GMD_id %in% duplicate_gmd_ids)]

duplicate_idx <- intersect(grep("9519AAB5-12D3-47C6-995C-02F25F58CD67", rownames(gmd_meta_per_exp)),
                           which(gmd_meta_per_exp$chromatogram %in% duplicate_chromatograms))

gmd_meta_per_exp <- gmd_meta_per_exp[-duplicate_idx,]

dim(gmd_meta_per_exp)

# remove rownames
rownames(gmd_meta_per_exp) <- NULL
head(gmd_meta_per_exp)
```


## Merge information about QC samples with gmd meta data
```{r merge information about QC samples with gmd meta data}

qc_info_final <- join(qc_info_blank_with_ids, gmd_meta_per_exp, by="chromatogram")
dim(qc_info_final)

qc_info_final <- qc_info_final[order(qc_info_final$chromatogram),]
```


# Load all GMD raw values
```{r load all GMD raw values}
all_values_cast <- read.table("data/gmd_raw_values_matrix.txt", sep = "\t", header = T, check.names = F)

dim(all_values_cast)

qc_values_idx <- which(rownames(all_values_cast) %in% qc_info_blank_with_ids$chromatogram)

qc_values <- all_values_cast[qc_values_idx,]
dim(qc_values)
```


# Exclude analytes with many NAs
**exclude analytes with more than 600 NAs (of 816 samples)**
```{r exclude analytes with many NAs}

qc_values_NA <- apply(qc_values, 2, FUN = function(x) sum(is.na(x)))

hist(qc_values_NA, breaks=50, col = "gray")

qc_values_NA_idx <- which(qc_values_NA > 500)
qc_values_part <- qc_values[,-qc_values_NA_idx]
dim(qc_values_part)
# 816 156

# calculate CV per analyte
qc_values_part_cv_per_analyte <- apply(qc_values_part, 2, FUN = func_CV)
head(qc_values_part_cv_per_analyte)
```


# Analysis of alkan analytes
```{r analysis of alkan analytes}
alkan_ids <- c("ABD59C73-3CA6-4508-B950-18336DB59BE3", "1310FACD-F2BF-4FD7-BC20-B21DF06EDE79", "B8926297-132A-4B5D-B9C2-294F3F006246",
               "2994BC5D-52BE-44F8-9824-0253A3BD2243", "E66BE919-93E8-4101-AB46-9612FE796394", "379E5588-B955-4C35-88E0-21E7DF38DE0E",
               "896D4B7E-BF33-4D54-82CE-7360D88E8DC8")
alkan_names <- c("Docosane", "Dodecane", "Dotriacontane", "Hexatriacontane", 
                 "Octacosane", "Octadecane", "Pentadecane")

alkan_idx <- which(names(qc_values_part_cv_per_analyte) %in% alkan_ids)
qc_values_part_cv_per_analyte[alkan_idx]

alkan_boolean <- logical(dim(qc_values_part)[2]) # 193x FALSE
alkan_boolean[alkan_idx] <- TRUE # set alkan to TRUE

qc_values_alkan <- data.frame(qc_values_part[, alkan_idx],
                              trial = qc_info_final$trial, 
                              check.names = F)

qc_values_alkan_melt <- melt(qc_values_alkan)
head(qc_values_alkan_melt)

# calculate CV per analyte and experiment/trial
qc_values_alkan_melt_cv <- ddply(qc_values_alkan_melt, c("variable","trial"), summarise, 
                                 cv = sd(value, na.rm = T)/mean(value, na.rm = T))
head(qc_values_alkan_melt_cv)

qc_values_alkan_melt_cv_cast <- cast(qc_values_alkan_melt_cv, variable ~ trial)
heatmap.2(as.matrix(qc_values_alkan_melt_cv_cast), trace = "none")
```


# PCA: all QC samples
```{r PCA all QC samples}
# log10 transformation
dim(qc_values_part) # without analytes with many NAs
# 816 samples, 156 analytes
qc_values_part_log10 <- log10(qc_values_part)

# Scaling
qc_pareto <- prep(as.matrix(qc_values_part_log10), scale = "pareto", center = FALSE)

# PCA
qc_pareto_rnipals <- pca(qc_pareto, nPcs=5, method="rnipals")

# R2cum
qc_pareto_rnipals@R2cum
```


## PCA: score plots
```{r PCA score plots}
pdf("figures/qc_concentration_series/qc_pca_scores_plot_before_norm.pdf")

# concentration effect
palette(heike_palette_7)
pairs(qc_pareto_rnipals@scores[,1:5], col=qc_info_final$concentration, pch=19, cex=1.5,main="concentration effect")

plot(qc_pareto_rnipals@scores[,1], qc_pareto_rnipals@scores[,2], 
     col=qc_info_final$concentration, pch=19, cex=1.5, main="concentration effect")
legend("bottomright", levels(qc_info_final$concentration), fill=1:7, cex=0.8)

plot(qc_pareto_rnipals@scores[,1], qc_pareto_rnipals@scores[,3], 
     col=qc_info_final$concentration, pch=19, cex=1.5, main="concentration effect")
legend("topleft", levels(qc_info_final$concentration), fill=1:7, cex=0.8)

# trial effect
palette(rainbow(20))
pairs(qc_pareto_rnipals@scores[,1:5], col=qc_info_final$trial, pch=19, cex=1.5, main="trial effect")

par(mar=c(5.1, 4.1, 4.1, 10), xpd=T)
plot(qc_pareto_rnipals@scores[,2], qc_pareto_rnipals@scores[,3], 
     col=qc_info_final$trial, pch=19, cex=1.5, main="trial effect")
legend("topright", levels(qc_info_final$trial), fill=1:20, cex=0.8, inset = c(-0.4,0))

plot(qc_pareto_rnipals@scores[,1], qc_pareto_rnipals@scores[,3], 
     col=qc_info_final$trial, pch=19, cex=1.5, main="trial effect")
legend("topright", levels(qc_info_final$trial), fill=1:20, cex=0.8, inset = c(-0.4,0))

dev.off()
```


# Normalized Values
```{r normalized values, warning=FALSE}
qc_values_part_log10_norm <- func_normalize(qc_values_part_log10, 
                                            trial_factors = qc_info_final,
                                            facs = c("concentration", "sequence_id", "batch_id", "AvgAnnotated"),
                                            keep = "concentration")

dim(qc_values_part_log10_norm)
# 816 156

# amount of NAs
sum(is.na(qc_values)) / (nrow(qc_values) * ncol(qc_values)) * 100
sum(is.na(qc_values_part_log10_norm)) / (nrow(qc_values_part_log10_norm) * ncol(qc_values_part_log10_norm)) * 100
# 32.7%

#################################

# using sorbitol = C391707B-E687-477D-9BF9-A3EA182D8DA4
# sorbitol_values_part_log10 <- qc_values_cast_part_log10[,"C391707B-E687-477D-9BF9-A3EA182D8DA4"]
# 
# dim(qc_info_final)
# qc_info_final_sorbitol <- cbind(qc_info_final, sorbitol_values_part_log10)
# colnames(qc_info_final_sorbitol)[9] <- "sorbitol_log10"
# class(qc_info_final_sorbitol$sorbitol_log10)
# 
# qc_values_cast_part_log10_norm_sorbitol <- apply(qc_values_cast_part_log10, 2, RemoveFactors, sam=qc_info_final_sorbitol, 
#                               facs=c("concentration", "sequence_id", "batch_id", "sorbitol_log10"), 
#                               keep="concentration")
# dim(qc_values_cast_part_log10_norm_sorbitol)
# # 155 218
# qc_values_cast_part_log10_norm_sorbitol[1:3,1:3]
```


# PCA: all QC samples normalized
```{r PCA all QC samples normalized}
# Scaling
qc_log10_norm_pareto <- prep(as.matrix(qc_values_part_log10_norm), scale = "pareto", center = FALSE)

# PCA
qc_log10_norm_pareto_rnipals <- pca(qc_log10_norm_pareto, nPcs=5, method="rnipals")

# R2cum
qc_log10_norm_pareto_rnipals@R2cum
qc_log10_norm_pareto_rnipals@R2

# before normalization
qc_pareto_rnipals@R2cum
```


## PCA: score plots normalized
```{r PCA score plots normalized}
pdf("figures/qc_concentration_series/qc_pca_scores_plot_after_norm.pdf")

# concentration effect
palette(heike_palette_7)
pairs(qc_log10_norm_pareto_rnipals@scores[,1:5], col=qc_info_final$concentration, pch=19, cex=1.5,main="concentration effect")

plot(qc_log10_norm_pareto_rnipals@scores[,1], qc_log10_norm_pareto_rnipals@scores[,2], 
     col=qc_info_final$concentration, pch=19, cex=1.5, main="concentration effect")
legend("bottomright", levels(qc_info_final$concentration), fill=1:7, cex=0.8)

plot(qc_log10_norm_pareto_rnipals@scores[,1], qc_log10_norm_pareto_rnipals@scores[,3], 
     col=qc_info_final$concentration, pch=19, cex=1.5, main="concentration effect")
legend("topleft", levels(qc_info_final$concentration), fill=1:7, cex=0.8)

plot(qc_log10_norm_pareto_rnipals@scores[,2], qc_log10_norm_pareto_rnipals@scores[,3], 
     col=qc_info_final$concentration, pch=19, cex=1.5, main="concentration effect")
legend("topleft", levels(qc_info_final$concentration), fill=1:7, cex=0.8)

# trial effect
palette(rainbow(20))
pairs(qc_log10_norm_pareto_rnipals@scores[,1:5], col=qc_info_final$trial, pch=19, cex=1.5, main="trial effect")

par(mar=c(5.1, 4.1, 4.1, 10), xpd=T)
plot(qc_log10_norm_pareto_rnipals@scores[,2], qc_log10_norm_pareto_rnipals@scores[,3], 
     col=qc_info_final$trial, pch=19, cex=1.5, main="trial effect")
legend("topright", levels(qc_info_final$trial), fill=1:20, cex=0.8, inset = c(-0.4,0))

plot(qc_log10_norm_pareto_rnipals@scores[,1], qc_log10_norm_pareto_rnipals@scores[,3], 
     col=qc_info_final$trial, pch=19, cex=1.5, main="trial effect")
legend("topright", levels(qc_info_final$trial), fill=1:20, cex=0.8, inset = c(-0.4,0))

dev.off()
```


# Dataset without blanks
```{r dataset without blanks}
blank_idx <- which(qc_info_final$concentration == "t0")

qc_info_wo_blanks <- qc_info_final[-blank_idx,]
qc_values_part_log10_norm_wo_blanks <- qc_values_part_log10_norm[-blank_idx,]

dim(qc_info_wo_blanks)
dim(qc_values_part_log10_norm_wo_blanks)
```


# Correlation analysis after normalization
```{r correlation analysis}
dim(qc_values_part_log10_norm)
# 816 156
tail(sort(apply(qc_values_part_log10_norm, 2, function(x) 816-length(na.omit(x)))))
hist(apply(qc_values_part_log10_norm, 2, function(x) 816-length(na.omit(x))), breaks=20, xlab="NAs per analyte", col="lightblue")


# before normalization
res_cor_conc_r <- rep(NA, 156)
for (i in 1:156){
    res_cor_conc_r[i] <- cor.test(qc_info_final$concentration_num, qc_values_part_log10[,i])$estimate
}

res_cor_conc_p <- rep(NA, 156)
for (i in 1:156){
    res_cor_conc_p[i] <- cor.test(qc_info_final$concentration_num, qc_values_part_log10[,i])$p.value
}

# after normalization

res_cor_conc_r_norm <- rep(NA, 156)
for (i in 1:156){
    res_cor_conc_r_norm[i] <- cor.test(qc_info_final$concentration_num, qc_values_part_log10_norm[,i])$estimate
}

res_cor_conc_p_norm <- rep(NA, 156)
for (i in 1:156){
    res_cor_conc_p_norm[i] <- cor.test(qc_info_final$concentration_num, qc_values_part_log10_norm[,i])$p.value
}

tail(order(res_cor_conc_r))
hist(res_cor_conc_r, col="grey")
hist(res_cor_conc_p, col="grey")

res_cor_conc_padj <- p.adjust(res_cor_conc_p, method="bonferroni")
res_cor_conc_padj_norm <- p.adjust(res_cor_conc_p_norm, method="bonferroni")

hist(res_cor_conc_padj, breaks=30, col="grey")
length(which(res_cor_conc_padj == 0))
length(which(res_cor_conc_padj_norm == 0))

# plot cor values with/without sorbitol as additional normalization factor
#sig_idx <- which(res_cor_conc_padj==0)
plot(res_cor_conc_r_norm, res_cor_conc_r)
abline(0,1)
#identify(res_cor_conc_r_sorb, res_cor_conc_r)
```

### scatterplot after normalization
```{r scatterplot after normalization}
pdf("figures/qc_concentration_series/qc_boxplot_normalized.pdf")
for (i in 1:156){
  boxplot(qc_values_part_log10_norm[,i] ~ qc_info_final$concentration, 
          main = colnames(qc_values_part_log10_norm)[i])
}
dev.off()


pdf("figures/qc_concentration_series/qc_boxplot_normalized_wo_blanks_log10.pdf")
for (i in 1:156){
  boxplot(qc_values_part_log10_norm_wo_blanks[,i] ~ round(log10(qc_info_wo_blanks$concentration_num),2), 
          main = colnames(qc_values_part_log10_norm_wo_blanks)[i],
          xlab = "log10(µl)", ylab = "log10 intensity")
}
dev.off()


pdf("figures/qc_concentration_series/qc_scatterplot_log10conc_normalized_wo_blanks.pdf")
for (i in 1:156){
  plot(log10(qc_info_wo_blanks$concentration_num), 
       qc_values_part_log10_norm_wo_blanks[,i], 
       main = colnames(qc_values_part_log10_norm_wo_blanks)[i], 
       pch=19, col = rgb(100,100,100,100, maxColorValue = 255), cex=1.5)
}
dev.off()

```


# ruv-2 normalization
```{r ruv-2 normalization}
dim(qc_values_part_log10)
# 816 samples, 156 analytes
sum(is.na(qc_values_part_log10))

dim(qc_pareto_rnipals@completeObs)
sum(is.na(qc_pareto_rnipals@completeObs))

# ruv needs data without NA!

ruv_k3 <- NormalizeRUVRand(Y = qc_pareto_rnipals@completeObs,
                           ctl = alkan_boolean, k = 3)

ruv_k3_norm <- ruv_k3$newY
dim(ruv_k3_norm)
sum(is.na(ruv_k3_norm))

# PCA of ruv-normalized data
# Scaling
ruv_k3_norm_pareto <- prep(as.matrix(ruv_k3_norm), scale = "pareto", center = FALSE)

# PCA
ruv_k3_norm_pareto_rnipals <- pca(ruv_k3_norm_pareto, nPcs = 5, method="rnipals")
round(ruv_k3_norm_pareto_rnipals@R2[1:5]*100,2)

palette(heike_palette_7)
pairs(ruv_k3_norm_pareto_rnipals@scores[,1:5], col=qc_info_final$concentration, pch=19, cex=1.5, main="concentration effect",
      labels=c("PC1 (46.7%)", "PC2 9.9%)","PC3 (4.5%)","PC4 (3%)","PC5 (2.3%)"))

# R2cum
qc_log10_norm_pareto_rnipals@R2cum
qc_log10_norm_pareto_rnipals@R2
```



# Save workspace and sessionInfo
```{r save workspace}
save.image("qc_concentration_series.RData")
sessionInfo()
```


